---
title: "Comparison with OTTERS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

library(knitr)
library(data.table)

```

***

This document is an attempt of simplifying the message: See what methods most accurately predicts gene expression. We will using the eQTLGen data (excluding GTEx) as the discovery sample, and GTEx v8 as the target sample.

Methods to include are:

- SUMMIT
- SUMMIT-FA
- EXPRESSO
- OTTERS
- BGW-TWAS

***

# Literature review

## SUMMIT and SUMMIT-FA

SUMMIT and SUMMIT-FA are not flexible - they requires a specific set of files with required file names and formats. It is going to be hard to implement. They release the models, so perhaps we can compare the R2 in some independent data.

These methods restrict to HapMap3 variants. They also output models in format resembling FUSION format.

They trained the models using the full eQTLGen sumstats, evaluating the 309 independent individuals in GTEx v8.

We can download the model from SUMMIT and SUMMIT-FA can be download from OSF. Do this and predict expression into independent GTEx v8.

***

## EXPRESSO

This software and the manuscript is more impressive. Really comprehensive comparison of methods using sumstats and individual-level data. EXPRESSO appears to be the best method. They also developed a TWAS-based drug repurposing pipeline. They apply their models to multi-ancestry GWAS using TESLA. They don't include SBayesRC or MegaPRS, which are some of the best performing PGS methods.

The authors position EXPRESSO as just one method for generating the models, and that it can be added in combination with the methods in OTTERS for improved power in TWAS, since OTTERS combines across models using Cauchy distribution. 

I could compare SBayesRC to EXPRESSO, and my pipeline to OTTERS. I could also include multi-ancestry eQTL data if there is any.

***

## BGW-TWAS

Develops approach for modelling cis and trans eQTLs. They use ROSMAP to develop the models, and evaluate in MCADGS.

It is novel in its use of trans-eQTL data, but performance is not great.

It would be interesting to see whether we can go beyond this, by applying some of the PGS methods to genome-wide eQTL data - eQTLGen have released this now.

***

# Analysis Phase 1: GenoPred with cis and trans

Try combining the cis- and trans-eQTL sumstats for eQTLGen, splitting by gene, and then using GenoPred to generate the prediction models, and then evaluate in independent GTEx participants

***

## Prepare eQTL sumstats

```{r}
library(data.table)

cis<-fread('~/oliverpainfel/Data/eQTLGen/full/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt.gz')
cis$GenePos <- NULL
cis$NrCohorts <- NULL
cis$FDR <- NULL
cis$BonferroniP <- NULL
cis$GeneChr <- NULL

trans<-fread('~/oliverpainfel/Data/eQTLGen/full/2018-09-04-trans-eQTLsFDR-CohortInfoRemoved-BonferroniAdded.txt.gz')
trans$GenePos <- NULL
trans$NrCohorts <- NULL
trans$FDR <- NULL
trans$BonferroniP <- NULL
trans$GeneChr <- NULL

# Read in EUR MAF
frq<-NULL
for(i in 1:22){
  frq<-rbind(frq, fread(paste0('/users/k1806347/oliverpainfel/Data/1KG/FUSION//1000G.EUR.',i,'.frq')))
}
names(frq)[names(frq) == 'MAF']<-'FREQ'

setkey(cis, Gene)
setkey(trans, Gene)
setkey(frq, SNP)

dir.create('~/oliverpainfel/Data/eQTLGen/full/cis_trans_combined')
genes <- unique(cis$Gene)
for(i in genes[1:50]){
  print(i)
  cis_i<-cis[.(i)]
  trans_i<-trans[.(i)]

  cis_i <- cis[cis$Gene == i,]
  trans_i <- trans[trans$Gene == i,]
  
  both_i<-rbind(cis_i, trans_i)
  
  names(both_i)[names(both_i) == 'AssessedAllele']<-'A1'
  names(both_i)[names(both_i) == 'OtherAllele']<-'A2'
  
  # Insert FREQ from EUR reference
  # There don't seem to be any strand flips
  frq_tmp<-frq[.(both_i$SNP)]
  both_i_match<-merge(both_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by=c('SNP','A1','A2'))
  both_i_flip<-merge(both_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by.x=c('SNP','A1','A2'), by.y=c('SNP','A2','A1'))
  both_i_flip$FREQ<-1-both_i_flip$FREQ
  both_i<-rbind(both_i_match, both_i_flip)
  
  # Approximate BETA, SE, and P
  both_i$BETA<-both_i$Z/sqrt((2*both_i$FREQ)*(1-both_i$FREQ)*(both_i$NrSamples+(both_i$Z^2)))
  both_i$SE<-abs(both_i$BETA)/abs(both_i$Z)
  
  both_i <- both_i[, c('SNPChr','SNPPos','SNP','A1','A2','BETA','SE','Pvalue','NrSamples')]
  names(both_i) <- c('CHR','BP','SNP','A1','A2','BETA','SE','P','N')
  
  fwrite(both_i, paste0('~/oliverpainfel/Data/eQTLGen/full/cis_trans_combined/',i,'.gz'), quote = F, sep = ' ', na= 'NA')
}

############################

library(data.table)

cis<-fread('~/oliverpainfel/Data/eQTLGen/full/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt.gz')
cis$GenePos <- NULL
cis$NrCohorts <- NULL
cis$FDR <- NULL
cis$BonferroniP <- NULL
cis$GeneChr <- NULL

# Read in EUR MAF
frq<-NULL
for(i in 1:22){
  frq<-rbind(frq, fread(paste0('/users/k1806347/oliverpainfel/Data/1KG/FUSION//1000G.EUR.',i,'.frq')))
}
names(frq)[names(frq) == 'MAF']<-'FREQ'

setkey(cis, Gene)
setkey(frq, SNP)

dir.create('~/oliverpainfel/Data/eQTLGen/full/cis_only')
genes <- unique(cis$Gene)
for(i in genes[1:50]){
  print(i)
  cis_i<-cis[.(i)]
  cis_i <- cis[cis$Gene == i,]

  names(cis_i)[names(cis_i) == 'AssessedAllele']<-'A1'
  names(cis_i)[names(cis_i) == 'OtherAllele']<-'A2'
  
  # Insert FREQ from EUR reference
  # There don't seem to be any strand flips
  frq_tmp<-frq[.(cis_i$SNP)]
  cis_i_match<-merge(cis_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by=c('SNP','A1','A2'))
  cis_i_flip<-merge(cis_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by.x=c('SNP','A1','A2'), by.y=c('SNP','A2','A1'))
  cis_i_flip$FREQ<-1-cis_i_flip$FREQ
  cis_i<-rbind(cis_i_match, cis_i_flip)
  
  # Approximate BETA, SE, and P
  cis_i$BETA<-cis_i$Z/sqrt((2*cis_i$FREQ)*(1-cis_i$FREQ)*(cis_i$NrSamples+(cis_i$Z^2)))
  cis_i$SE<-abs(cis_i$BETA)/abs(cis_i$Z)
  
  cis_i <- cis_i[, c('SNPChr','SNPPos','SNP','A1','A2','BETA','SE','Pvalue','NrSamples')]
  names(cis_i) <- c('CHR','BP','SNP','A1','A2','BETA','SE','P','N')
  
  fwrite(cis_i, paste0('~/oliverpainfel/Data/eQTLGen/full/cis_only/',i,'.gz'), quote = F, sep = ' ', na= 'NA')
}

```

***

## Run GenoPred

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

# Create config file
conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml',
  'gwas_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_methods: ['ptclump','sbayesrc','sbayesr','quickprs','lassosum','ldpred2','prscs','dbslmm','megaprs']",
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']",
  "lassosum_pseudo_only: true",
  "megaprs_pseudo_only: true",
  "prscs_phi: ['auto']",
  "dbslmm_h2f: ['1']",
  "cores_prep_pgs: 1",
  "quickprs_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/quickprs/hm3",
  "sbayesrc_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/sbayesrc/hm3"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml', col.names = F, row.names = F, quote = F)

# Create target_list
target_list <- data.table(name = 'gtex_v8',
                          path = '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/processed/gtex_v8.snp.maf001.hm3',
                          type = 'plink2',
                          indiv_report = F)

write.table(
  target_list,
  '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  col.names = T,
  row.names = F,
  quote = F,
  sep = ' '
)

# Create gwas_list
genes <- gsub('.gz', '', list.files('~/oliverpainfel/Data/eQTLGen/full/cis_trans_combined'))

gwas_list <- data.table(
  name = genes,
  path = paste0('/users/k1806347/oliverpainfel/Data/eQTLGen/full/cis_trans_combined/', genes,'.gz'),
  population = 'EUR',
  n = NA,
  sampling = NA,
  prevalence = NA,
  mean = NA,
  sd = NA,
  label = genes)

write.table(gwas_list, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list.txt', col.names = T, row.names = F, quote = F, sep = ' ')

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml \
  target_pgs -n 
```

This is probably an inefficient way of running these PGS methods. If the sumstats were truly genome-wide then it would be perfect, but eQTLGen trans QTLs are restricted to 10K variants. Several methods do run in a fairly efficient manor (ldpred2, lassosum, quickprs, ptclump). quickprs and sbayesrc will have issues due to poor overlap with the reference. dbslmm and prs-cs are run per chromosome so won't work.

Two options, find truly genome-wide trans-eQTL dataset, or continue with cis-eQTLs only.

For cis-eQTL only analysis, we can make GenoPred more efficient by ensuring only the relevant LD reference data is used. We just need to subset the reference to include eQTL sumstats for given gene, and adjsut CHROMS to relevant chromosome. SBayesRC we can subset the reference to the relevant blocks. The only extra overhead of running in GenoPred would be that there is seperate job for every method, but shouldn't be much of an overhead. It will also track jobs completing better. WuickPRS can't be so easily remedied as the reference data assumes a certain overlap with the sumstats. MegaPRS would be fine though. I can make an R function for packaging models in FUSION format for TWAS.

Note. I am running into an issue regarding reference data for sbayesr and sbayesrc 

DBSLMM and LDpred2 require SNP-h2 estimate. Currently this comes from LDSC, which is inaccurate when using subsets of variants. It would be better if this was done using sbayesr or sbayesrc. Problem is SBayesRC is taking a while to run when on subset of variants. We could test when using cis-eQTLs only, or using SBayesR which I know if fairly accurate and quick. We would have to do this prior to running DBSLMM and LDpred2, and provide an option to eithe estimate SNP-h2 using LDSC or SBayesR.

***

# Analysis Phase 2: GenoPred with cis only

## Run GenoPred

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

# Create config file
conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml',
  'gwas_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list_cis_only.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_methods: ['ptclump','quickprs','lassosum','ldpred2','prscs','dbslmm','megaprs','sbayesr']",
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']",
  "lassosum_pseudo_only: true",
  "ldpred2_model: ['auto']",
  "megaprs_pseudo_only: true",
  "prscs_phi: ['auto']",
  "dbslmm_h2f: ['1']",
  "h2_method: 'sbayesr'",
  "cores_prep_pgs: 1",
  "quickprs_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/quickprs/hm3",
  "sbayesrc_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/sbayesrc/hm3"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml', col.names = F, row.names = F, quote = F)

# Create gwas_list
genes <- gsub('.gz', '', list.files('~/oliverpainfel/Data/eQTLGen/full/cis_only'))

gwas_list <- data.table(
  name = genes,
  path = paste0('/users/k1806347/oliverpainfel/Data/eQTLGen/full/cis_only/', genes,'.gz'),
  population = 'EUR',
  n = NA,
  sampling = NA,
  prevalence = NA,
  mean = NA,
  sd = NA,
  label = genes)

write.table(gwas_list, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list_cis_only.txt', col.names = T, row.names = F, quote = F, sep = ' ')

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml \
  target_pgs -n 
```

Testing using ptclump. Check it runs with other methods. 

sbayesrc very slow due to using all 591 LD blocks despite sumstats only including one locus.

target_scoring: Maybe we can speed this up by only include chr for each PGS that have >1 non-zero effect? Haven't tested yet so maybe not worth it. Or we can just export the weights into FUSION format and parse to FeaturePred...

***

## Evaluate PGS

```{r}
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
pgs <- read_pgs(config = config, pop = 'EUR', pseudo_only = T)$gtex_v8$EUR

# Get a list of genes
score_files <- list_score_files(config = config)
genes <- unique(score_files$name)

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in names(pgs)){
  obs_resid_i <- obs_resid[, c('ID', i), with = F]
  names(obs_resid_i)<-c('ID','obs')
  pgs_i <- pgs[[i]]
  for(j in names(pgs_i)){
    pgs_i_j <- pgs[[i]][[j]]
    pgs_i_j$FID<-NULL
    if(ncol(pgs_i_j) == 1){
      next
    }
    names(pgs_i_j)<-c('ID','pred')
    
    both_i_j <- merge(obs_resid_i, pgs_i_j, by = 'ID')
    both_i_j <- both_i_j[complete.cases(both_i_j),]
    
    if(nrow(both_i_j) > 0){
      mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
      
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = coef(mod)[2,1],
          se = coef(mod)[2,2],
          p = coef(mod)[2,4],
          n = nrow(both_i_j)
        ))
    } else {
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = NA,
          se = NA,
          p = NA,
          n = 0
        ))      
    }
  }
}

library(ggplot2)
library(cowplot)

# Restrict to genes available for all methods
cor_res <- cor_res[complete.cases(cor_res),]
cor_res_complete <- NULL
for(i in names(pgs)){
    pgs_i <- pgs[[i]]
    if(all(pgs_methods %in% names(pgs_i))){
      cor_res_complete <- rbind(cor_res_complete, cor_res[cor_res$gene == i,])
    }
}
cor_res <- cor_res_complete

ggplot(cor_res, aes(x=gene, y=r , fill = pgs_method)) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

# Compute medians using aggregate()
medians <- aggregate(r ~ pgs_method, data = cor_res, FUN = median, na.rm = TRUE)

ggplot(cor_res, aes(x = pgs_method, y = r, fill = pgs_method)) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = pgs_method, y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

```

This looks good. It shows different methods perform best for each gene. SBayesR and DBSLMM perform well.

***

# Analysis Phase 3: eqtl_to_twas with cis only

Lets run the same analysis, but derive the models using the eqtl_to_twas code to see how it compares in terms of predictive utility and computational efficiency.

We have generated models using the script previously, so use those models. Use FeaturePred to predict expression into the GTEx sample.

```{bash}

conda activate feature_pred
sbatch -p neurohack_cpu,cpu --mem 30G -n 40 -t 200 --wrap="
  Rscript /users/k1806347/oliverpainfel/Software/MyGit/Predicting-TWAS-features/FeaturePred.V2.0.R \
  	--PLINK_prefix_chr /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/processed/gtex_v8.snp.maf001.hm3.chr \
  	--ref_ld_chr /users/k1806347/oliverpainfel/Data/1KG/FUSION/1000G.EUR. \
  	--ref_maf /users/k1806347/oliverpainfel/Data/1KG/FUSION/1000G.EUR. \
  	--weights /users/k1806347/oliverpainfel/Data/eQTLGen/eQTL_to_TWAS/eQTLGen.eQTL.pos \
  	--weights_dir /users/k1806347/oliverpainfel/Data/eQTLGen/eQTL_to_TWAS/ \
  	--plink /users/k1806347/oliverpainfel/Software/plink1.9.sh \
  	--n_cores 40 \
  	--memory 10000 \
  	--all_mod T \
  	--pigz /users/k1806347/oliverpainfel/Software/pigz \
  	--output /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/eqtl_to_twas/output/gtex_v8.eqtlgen
"

```

## Evaluate PGS

```{r}
library(data.table)

# Get the list of genes tested above
config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
genes <- unique(score_files$name)

# Read in predicted expression by eqtl_to_twas
pred_exp <- fread('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/eqtl_to_twas/output/gtex_v8.eqtlgen/FeaturePredictions_eQTLGen.eQTL.txt.gz')
pred_exp<-pred_exp[, c(1, which(grepl(paste(genes, collapse = '|'), names(pred_exp)))), with=F]
names(pred_exp)[1] <- 'ID'

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in genes){
  pred_exp_i <- pred_exp[, c(1, which(grepl(paste0('\\.', i, '\\.'), names(pred_exp)))), with = F]
  obs_resid_i <- obs_resid[, c('ID', i), with = F]
  names(obs_resid_i)<-c('ID','obs')
  methods <- gsub('.*\\.', '', names(pred_exp_i)[-1])
  for(j in methods){
    pred_exp_i_j <- pred_exp_i[, c(1, which(grepl(paste0('\\.', j, '$'), names(pred_exp_i)))), with = F]
    if(ncol(pgs_i_j) == 1){
      next
    }
    names(pred_exp_i_j)<-c('ID','pred')
    
    both_i_j <- merge(obs_resid_i, pred_exp_i_j, by = 'ID')
    both_i_j <- both_i_j[complete.cases(both_i_j),]
    
    if(nrow(both_i_j) > 0){
      mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
      
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = coef(mod)[2,1],
          se = coef(mod)[2,2],
          p = coef(mod)[2,4],
          n = nrow(both_i_j)
        ))
    } else {
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = NA,
          se = NA,
          p = NA,
          n = 0
        ))      
    }
  }
}

library(ggplot2)
library(cowplot)

# Restrict to genes available for all methods
cor_res <- cor_res[complete.cases(cor_res),]
cor_res_complete <- NULL
for(i in names(pgs)){
    pgs_i <- pgs[[i]]
    if(all(pgs_methods %in% names(pgs_i))){
      cor_res_complete <- rbind(cor_res_complete, cor_res[cor_res$gene == i,])
    }
}
cor_res <- cor_res_complete

ggplot(cor_res, aes(x=gene, y=r , fill = pgs_method)) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

# Compute medians using aggregate()
medians <- aggregate(r ~ pgs_method, data = cor_res, FUN = median, na.rm = TRUE)

ggplot(cor_res, aes(x = pgs_method, y = r, fill = pgs_method)) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = pgs_method, y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

```

The predicted-observed correlations are highly concordant between GenoPred and eqtl_to_twas methods. Benefits of GenoPred are incusion of megaprs and quickprs, integration with wider GenoPred pipeline. Benefits of eqtl_to_twas are that it includes the top1 and susie is probably faster and requires fewer job submissions. I kind of want to keep developing GenoPred as seems more fun and better integrated with GenoPred ecosystem. We can include top1 and susie models if we feel necessary, and can think of ways to speed up target scoring.

***

# Predict expression using EXPRESSO

```{bash}
mkdir ~/
wget https://github.com/LidaWangPSU/EXPRESSO/blob/main/trained_model/EXPRESSO_Whole_Blood_eQTLGen_weights_hg19.txt.zip
```

***

To do:
- Fix SBayesRC reference issue - I think it is fixed but reference data download is slow - 
- Make SBayesRC more efficient when using locus specific sumstats
- Compare prediction when using models from SUMMIT, SUMMIT-FA, OTTERS and EXPRESSO

***
