---
title: "Comparison with OTTERS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

library(knitr)
library(data.table)

```

***

This document is an attempt of simplifying the message: See what methods most accurately predicts gene expression. We will using the eQTLGen data (excluding GTEx) as the discovery sample, and GTEx v8 as the target sample.

Methods to include are:

- SUMMIT
- SUMMIT-FA
- EXPRESSO
- OTTERS
- BGW-TWAS

***

# Literature review

## SUMMIT and SUMMIT-FA

SUMMIT and SUMMIT-FA are not flexible - they requires a specific set of files with required file names and formats. It is going to be hard to implement. They release the models, so perhaps we can compare the R2 in some independent data.

These methods restrict to HapMap3 variants. They also output models in format resembling FUSION format.

They trained the models using the full eQTLGen sumstats, evaluating the 309 independent individuals in GTEx v8.

We can download the model from SUMMIT and SUMMIT-FA can be download from OSF. Do this and predict expression into independent GTEx v8.

***

## EXPRESSO

This software and the manuscript is more impressive. Really comprehensive comparison of methods using sumstats and individual-level data. EXPRESSO appears to be the best method. They also developed a TWAS-based drug repurposing pipeline. They apply their models to multi-ancestry GWAS using TESLA. They don't include SBayesRC or MegaPRS, which are some of the best performing PGS methods.

The authors position EXPRESSO as just one method for generating the models, and that it can be added in combination with the methods in OTTERS for improved power in TWAS, since OTTERS combines across models using Cauchy distribution. 

I could compare SBayesRC to EXPRESSO, and my pipeline to OTTERS. I could also include multi-ancestry eQTL data if there is any.

***

## BGW-TWAS

Develops approach for modelling cis and trans eQTLs. They use ROSMAP to develop the models, and evaluate in MCADGS.

It is novel in its use of trans-eQTL data, but performance is not great.

It would be interesting to see whether we can go beyond this, by applying some of the PGS methods to genome-wide eQTL data - eQTLGen have released this now.

***

# Full eQTLGen

To allow comparison to some previous studies lets start by using the full eQTLGen data, and then evaluate models in the independent subset of GTEx v8.

***

## GenoPred with cis and trans

Try combining the cis- and trans-eQTL sumstats for eQTLGen, splitting by gene, and then using GenoPred to generate the prediction models, and then evaluate in independent GTEx participants

***

### Prepare eQTL sumstats

```{r}
library(data.table)

cis<-fread('~/oliverpainfel/Data/eQTLGen/full/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt.gz')
cis$GenePos <- NULL
cis$NrCohorts <- NULL
cis$FDR <- NULL
cis$BonferroniP <- NULL
cis$GeneChr <- NULL

trans<-fread('~/oliverpainfel/Data/eQTLGen/full/2018-09-04-trans-eQTLsFDR-CohortInfoRemoved-BonferroniAdded.txt.gz')
trans$GenePos <- NULL
trans$NrCohorts <- NULL
trans$FDR <- NULL
trans$BonferroniP <- NULL
trans$GeneChr <- NULL

# Read in EUR MAF
frq<-NULL
for(i in 1:22){
  frq<-rbind(frq, fread(paste0('/users/k1806347/oliverpainfel/Data/1KG/FUSION//1000G.EUR.',i,'.frq')))
}
names(frq)[names(frq) == 'MAF']<-'FREQ'

setkey(cis, Gene)
setkey(trans, Gene)
setkey(frq, SNP)

dir.create('~/oliverpainfel/Data/eQTLGen/full/cis_trans_combined')
genes <- unique(cis$Gene)
for(i in genes[1:50]){
  print(i)
  cis_i<-cis[.(i)]
  trans_i<-trans[.(i)]

  cis_i <- cis[cis$Gene == i,]
  trans_i <- trans[trans$Gene == i,]
  
  both_i<-rbind(cis_i, trans_i)
  
  names(both_i)[names(both_i) == 'AssessedAllele']<-'A1'
  names(both_i)[names(both_i) == 'OtherAllele']<-'A2'
  
  # Insert FREQ from EUR reference
  # There don't seem to be any strand flips
  frq_tmp<-frq[.(both_i$SNP)]
  both_i_match<-merge(both_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by=c('SNP','A1','A2'))
  both_i_flip<-merge(both_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by.x=c('SNP','A1','A2'), by.y=c('SNP','A2','A1'))
  both_i_flip$FREQ<-1-both_i_flip$FREQ
  both_i<-rbind(both_i_match, both_i_flip)
  
  # Approximate BETA, SE, and P
  both_i$BETA<-both_i$Z/sqrt((2*both_i$FREQ)*(1-both_i$FREQ)*(both_i$NrSamples+(both_i$Z^2)))
  both_i$SE<-abs(both_i$BETA)/abs(both_i$Z)
  
  both_i <- both_i[, c('SNPChr','SNPPos','SNP','A1','A2','BETA','SE','Pvalue','NrSamples')]
  names(both_i) <- c('CHR','BP','SNP','A1','A2','BETA','SE','P','N')
  
  fwrite(both_i, paste0('~/oliverpainfel/Data/eQTLGen/full/cis_trans_combined/',i,'.gz'), quote = F, sep = ' ', na= 'NA')
}

############################

library(data.table)

cis<-fread('~/oliverpainfel/Data/eQTLGen/full/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt.gz')
cis$GenePos <- NULL
cis$NrCohorts <- NULL
cis$FDR <- NULL
cis$BonferroniP <- NULL
cis$GeneChr <- NULL

cis<-cis[cis$SNPChr == 22,]

# Read in EUR MAF
frq<-NULL
for(i in 1:22){
  frq<-rbind(frq, fread(paste0('/users/k1806347/oliverpainfel/Data/1KG/FUSION//1000G.EUR.',i,'.frq')))
}
names(frq)[names(frq) == 'MAF']<-'FREQ'

setkey(cis, Gene)
setkey(frq, SNP)

dir.create('~/oliverpainfel/Data/eQTLGen/full/cis_only')
genes <- unique(cis$Gene)
for(i in genes[1:200]){
  print(i)
  cis_i<-cis[.(i)]
  cis_i <- cis[cis$Gene == i,]

  names(cis_i)[names(cis_i) == 'AssessedAllele']<-'A1'
  names(cis_i)[names(cis_i) == 'OtherAllele']<-'A2'
  
  # Insert FREQ from EUR reference
  # There don't seem to be any strand flips
  frq_tmp<-frq[.(cis_i$SNP)]
  cis_i_match<-merge(cis_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by=c('SNP','A1','A2'))
  cis_i_flip<-merge(cis_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by.x=c('SNP','A1','A2'), by.y=c('SNP','A2','A1'))
  cis_i_flip$FREQ<-1-cis_i_flip$FREQ
  cis_i<-rbind(cis_i_match, cis_i_flip)
  
  # Approximate BETA, SE, and P
  cis_i$BETA<-cis_i$Z/sqrt((2*cis_i$FREQ)*(1-cis_i$FREQ)*(cis_i$NrSamples+(cis_i$Z^2)))
  cis_i$SE<-abs(cis_i$BETA)/abs(cis_i$Z)
  
  cis_i <- cis_i[, c('SNPChr','SNPPos','SNP','A1','A2','BETA','SE','Pvalue','NrSamples')]
  names(cis_i) <- c('CHR','BP','SNP','A1','A2','BETA','SE','P','N')
  
  fwrite(cis_i, paste0('~/oliverpainfel/Data/eQTLGen/full/cis_only/',i,'.gz'), quote = F, sep = ' ', na= 'NA')
}

write.table(genes[1:200], '~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', row.names=F, col.names=F, quote=F)
```

***

### Run GenoPred

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

# Create config file
conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml',
  'gwas_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_methods: ['ptclump','sbayesrc','sbayesr','quickprs','lassosum','ldpred2','prscs','dbslmm','megaprs']",
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']",
  "lassosum_pseudo_only: true",
  "megaprs_pseudo_only: true",
  "prscs_phi: ['auto']",
  "dbslmm_h2f: ['1']",
  "cores_prep_pgs: 1",
  "quickprs_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/quickprs/hm3",
  "sbayesrc_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/sbayesrc/hm3"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml', col.names = F, row.names = F, quote = F)

# Create target_list
target_list <- data.table(name = 'gtex_v8',
                          path = '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/processed/gtex_v8.snp.maf001.hm3',
                          type = 'plink2',
                          indiv_report = F)

write.table(
  target_list,
  '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  col.names = T,
  row.names = F,
  quote = F,
  sep = ' '
)

# Create gwas_list
genes <- gsub('.gz', '', list.files('~/oliverpainfel/Data/eQTLGen/full/cis_trans_combined'))

gwas_list <- data.table(
  name = genes,
  path = paste0('/users/k1806347/oliverpainfel/Data/eQTLGen/full/cis_trans_combined/', genes,'.gz'),
  population = 'EUR',
  n = NA,
  sampling = NA,
  prevalence = NA,
  mean = NA,
  sd = NA,
  label = genes)

write.table(gwas_list, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list.txt', col.names = T, row.names = F, quote = F, sep = ' ')

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml \
  target_pgs -n 
```

This is probably an inefficient way of running these PGS methods. If the sumstats were truly genome-wide then it would be perfect, but eQTLGen trans QTLs are restricted to 10K variants. Several methods do run in a fairly efficient manor (ldpred2, lassosum, quickprs, ptclump). quickprs and sbayesrc will have issues due to poor overlap with the reference. dbslmm and prs-cs are run per chromosome so won't work.

Two options, find truly genome-wide trans-eQTL dataset, or continue with cis-eQTLs only.

For cis-eQTL only analysis, we can make GenoPred more efficient by ensuring only the relevant LD reference data is used. We just need to subset the reference to include eQTL sumstats for given gene, and adjsut CHROMS to relevant chromosome. SBayesRC we can subset the reference to the relevant blocks. The only extra overhead of running in GenoPred would be that there is seperate job for every method, but shouldn't be much of an overhead. It will also track jobs completing better. WuickPRS can't be so easily remedied as the reference data assumes a certain overlap with the sumstats. MegaPRS would be fine though. I can make an R function for packaging models in FUSION format for TWAS.

Note. I am running into an issue regarding reference data for sbayesr and sbayesrc 

DBSLMM and LDpred2 require SNP-h2 estimate. Currently this comes from LDSC, which is inaccurate when using subsets of variants. It would be better if this was done using sbayesr or sbayesrc. Problem is SBayesRC is taking a while to run when on subset of variants. We could test when using cis-eQTLs only, or using SBayesR which I know if fairly accurate and quick. We would have to do this prior to running DBSLMM and LDpred2, and provide an option to eithe estimate SNP-h2 using LDSC or SBayesR.

***

## GenoPred with cis only

### Run GenoPred

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

# Create config file
conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml',
  'gwas_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list_cis_only.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_methods: ['ptclump','quickprs','lassosum','ldpred2','prscs','dbslmm','megaprs','sbayesr','sdpr']",
  "pgs_scaling: ['discrete']",
  "ptclump_pts: ['0.001','0.05']",
  "target_populations: ['EUR']",
  "lassosum_pseudo_only: true",
  "ldpred2_model: ['auto']",
  "megaprs_pseudo_only: true",
  "prscs_phi: ['auto']",
  "dbslmm_h2f: ['1']",
  "h2_method: 'sbayesr'",
  "cores_prep_pgs: 1",
  "quickprs_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/quickprs/hm3",
  "sbayesrc_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/sbayesrc/hm3"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml', col.names = F, row.names = F, quote = F)

# Create gwas_list
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

gwas_list <- data.table(
  name = genes,
  path = paste0('/users/k1806347/oliverpainfel/Data/eQTLGen/full/cis_only/', genes,'.gz'),
  population = 'EUR',
  n = NA,
  sampling = NA,
  prevalence = NA,
  mean = NA,
  sd = NA,
  label = genes)

write.table(gwas_list, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list_cis_only.txt', col.names = T, row.names = F, quote = F, sep = ' ')

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml \
  target_pgs -n 
```

Testing using ptclump. Check it runs with other methods. 

sbayesrc very slow due to using all 591 LD blocks despite sumstats only including one locus.

target_scoring: Maybe we can speed this up by only include chr for each PGS that have >1 non-zero effect? Haven't tested yet so maybe not worth it. Or we can just export the weights into FUSION format and parse to FeaturePred...

***

### Check computational resources

```{r}
library(data.table)
library(ggplot2)
library(cowplot)

setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')

# Get some key variables from config
config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in configuration specific benchmark files
bm_files_i <- list.files(paste0(outdir, '/reference/benchmarks/'), full.names = T)

# Subset benchmarks for pgs_methods
bm_files_i <- bm_files_i[grepl('prep_pgs_', bm_files_i)]

# Subset to benchmarks for gwas/gwas_groups in config
scores <- list_score_files(config)
bm_files_i <- bm_files_i[grepl(paste0('-', unique(scores$name),'.txt', collapse = '|'), bm_files_i)]

# Read in benchmark files
bm_dat_all <- do.call(rbind, lapply(bm_files_i, function(file) {
  tmp <- fread(file)
  tmp$file <- basename(file)
  return(tmp)
}))

# Create rule column
bm_dat_all$rule <- gsub('-.*','',bm_dat_all$file)

# Create method column
bm_dat_all$method <-
  gsub('_i', '', gsub('prep_pgs_', '', bm_dat_all$rule))

bm_dat_all <- merge(bm_dat_all, pgs_method_labels, by = 'method', all.x=T)

#############
# Time
#############

# Calculate average time taken for each method
method_avg <- NULL
for(i in unique(bm_dat_all$label)){
  method_avg <- rbind(
    method_avg,
    data.frame(
      method = bm_dat_all$method[bm_dat_all$label == i][1],
      Method = i,
      Time = mean(bm_dat_all$s[bm_dat_all$label == i])
    )
  )
}

# Format the time taken nicely
method_avg$Time_clean[method_avg$Time < 60] <-
  paste0(round(method_avg$Time[method_avg$Time < 60], 1), ' sec')
method_avg$Time_clean[method_avg$Time > 60] <-
  paste0(round(method_avg$Time[method_avg$Time > 60] / 60, 1), ' min')
method_avg$Time_clean[method_avg$Time > 3600] <-
  paste0(round(method_avg$Time[method_avg$Time > 3600] / 60 / 60, 1), ' hr')

# Convert time in seconds to hours
method_avg$Time_hour <- method_avg$Time / 60/60

method_avg$Method <- factor(method_avg$Method, levels = c("DBSLMM", "lassosum", "LDpred2", "MegaPRS", "PRS-CS", "pT+clump", "QuickPRS", "SBayesR", "SDPR"))

ggplot(method_avg, aes(x = Method, y = Time_hour, fill = Method)) +
  geom_bar(stat = "identity", position="dodge") +
  geom_text(aes(label = Time_clean), vjust = 0.5, angle = 90, hjust = -0.2, position = position_dodge(width = 0.9)) +
  labs(x = NULL, y = "Time (hours)") +
  ylim(0, max(method_avg$Time_hour) + (max(method_avg$Time_hour)/5)) +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  panel_border() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#############
# Memory
#############

# Calculate average max_rss for each method
method_avg_mem <- NULL
for(i in unique(bm_dat_all$label)){
  method_avg_mem <- rbind(
    method_avg_mem,
    data.frame(
      method = bm_dat_all$method[bm_dat_all$label == i][1],
      Method = i,
      Memory = mean(bm_dat_all$max_rss[bm_dat_all$label == i])
    )
  )
}

# Format the Memory nicely
method_avg_mem$Memory_clean <-
  paste0(round(method_avg_mem$Memory/1000, 2), ' Gb')

ggplot(method_avg_mem, aes(x = Method, y = Memory, fill = Method)) +
  geom_bar(stat = "identity", position="dodge") +
  geom_text(aes(label = Memory_clean), vjust = -0.5, position = position_dodge(width = 0.9)) +
  labs(x = "PGS Method", y = "Memory (Mb)") +
  theme_half_open() +
  background_grid() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="none")

```

***

## eqtl_to_twas with cis only

Lets run the same analysis, but derive the models using the eqtl_to_twas code to see how it compares in terms of predictive utility and computational efficiency.

We have generated models using the script previously, so use those models. Use FeaturePred to predict expression into the GTEx sample.

```{bash}

conda activate feature_pred
sbatch -p neurohack_cpu,cpu --mem 30G -n 40 -t 200 --wrap="
  Rscript /users/k1806347/oliverpainfel/Software/MyGit/Predicting-TWAS-features/FeaturePred.V2.0.R \
  	--PLINK_prefix_chr /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/processed/gtex_v8.snp.maf001.hm3.chr \
  	--ref_ld_chr /users/k1806347/oliverpainfel/Data/1KG/FUSION/1000G.EUR. \
  	--ref_maf /users/k1806347/oliverpainfel/Data/1KG/FUSION/1000G.EUR. \
  	--weights /users/k1806347/oliverpainfel/Data/eQTLGen/eQTL_to_TWAS/eQTLGen.eQTL.pos \
  	--weights_dir /users/k1806347/oliverpainfel/Data/eQTLGen/eQTL_to_TWAS/ \
  	--plink /users/k1806347/oliverpainfel/Software/plink1.9.sh \
  	--n_cores 40 \
  	--memory 10000 \
  	--all_mod T \
  	--pigz /users/k1806347/oliverpainfel/Software/pigz \
  	--output /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/eqtl_to_twas/output/gtex_v8.eqtlgen
"

```

***

```{r}
library(data.table)

# Get the list of genes tested above
genes <- gsub('.gz', '', list.files('~/oliverpainfel/Data/eQTLGen/full/cis_only'))

# Read in predicted expression by eqtl_to_twas
pred_exp <- fread('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/eqtl_to_twas/output/gtex_v8.eqtlgen/FeaturePredictions_eQTLGen.eQTL.txt.gz')
pred_exp<-pred_exp[, c(1, which(grepl(paste(genes, collapse = '|'), names(pred_exp)))), with=F]
names(pred_exp)[1] <- 'ID'

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in genes){
  pred_exp_i <- pred_exp[, c(1, which(grepl(paste0('\\.', i, '\\.'), names(pred_exp)))), with = F]
  obs_resid_i <- obs_resid[, c('ID', i), with = F]
  names(obs_resid_i)<-c('ID','obs')
  methods <- gsub('.*\\.', '', names(pred_exp_i)[-1])
  for(j in methods){
    pred_exp_i_j <- pred_exp_i[, c(1, which(grepl(paste0('\\.', j, '$'), names(pred_exp_i)))), with = F]
    if(ncol(pgs_i_j) == 1){
      next
    }
    names(pred_exp_i_j)<-c('ID','pred')
    
    both_i_j <- merge(obs_resid_i, pred_exp_i_j, by = 'ID')
    both_i_j <- both_i_j[complete.cases(both_i_j),]
    
    if(nrow(both_i_j) > 0){
      mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
      
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = coef(mod)[2,1],
          se = coef(mod)[2,2],
          p = coef(mod)[2,4],
          n = nrow(both_i_j)
        ))
    } else {
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = NA,
          se = NA,
          p = NA,
          n = 0
        ))      
    }
  }
}

library(ggplot2)
library(cowplot)

ggplot(cor_res, aes(x=gene, y=r , fill = pgs_method)) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

# Restrict to genes available for all methods
cor_res <- cor_res[complete.cases(cor_res),]
cor_res_complete <- NULL
for(i in names(pgs)){
    pgs_i <- pgs[[i]]
    if(all(pgs_methods %in% names(pgs_i))){
      cor_res_complete <- rbind(cor_res_complete, cor_res[cor_res$gene == i,])
    }
}

cor_res <- cor_res_complete
# Compute medians using aggregate()
medians <- aggregate(r ~ pgs_method, data = cor_res, FUN = median, na.rm = TRUE)

ggplot(cor_res, aes(x = pgs_method, y = r, fill = pgs_method)) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = pgs_method, y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

```

The predicted-observed correlations are highly concordant between GenoPred and eqtl_to_twas methods. Benefits of GenoPred are incusion of megaprs and quickprs, integration with wider GenoPred pipeline. Benefits of eqtl_to_twas are that it includes the top1 and susie is probably faster and requires fewer job submissions. I kind of want to keep developing GenoPred as seems more fun and better integrated with GenoPred ecosystem. We can include top1 and susie models if we feel necessary, and can think of ways to speed up target scoring.

***

## OTTERS

We will need to run OTTERS ourselves to generate models restricted to HapMap3. Some OTTERS code already in otters_comp.

***

### Install

```{bash}
# Install OTTERS v1.0.0
cd /users/k1806347/oliverpainfel/Software
git clone https://github.com/daiqile96/OTTERS.git 
cd OTTERS
git checkout v1.0.0

# Install tabix and bgzip
cd /users/k1806347/oliverpainfel/Software
wget https://sourceforge.net/projects/samtools/files/tabix/tabix-0.2.6.tar.bz2
tar jxvf tabix-0.2.6.tar.bz2
cd tabix-0.2.6
make

# copy the binary bgzip and tabix bin folder so can be called from PATH
mkdir /users/k1806347/oliverpainfel/Software/bin
cp bgzip tabix /users/k1806347/oliverpainfel/Software/bin

# copy the plink1.9 binary and add to PATH
cp /users/k1806347/oliverpainfel/Software/plink1.9/plink /users/k1806347/oliverpainfel/Software/bin/
export PATH=/users/k1806347/oliverpainfel/Software/bin:$PATH 

# Create a conda environment to run otters
cd /users/k1806347/oliverpainfel/Software/OTTERS
mamba env create --file otters_env.yaml
conda activate otters

# Install lassosum
Rscript -e 'remotes::install_github("tshmak/lassosum@v0.4.5")'

# Install SDPR
cd /users/k1806347/oliverpainfel/Software/bin
git clone https://github.com/eldronzhou/SDPR.git
cd SDPR
git checkout v0.9.1
```

***

### Prepare input

```{r}
# Get the list of genes tested above
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

##
# Annotation file (CHROM GeneStart GeneEnd TargetID)
##
pos<-fread('/users/k1806347/oliverpainfel/Data/eQTLGen/eQTL_to_TWAS/eQTLGen.eQTL.pos')

pos <- pos[pos$ID %in% genes,]
annot <- data.table(
  CHROM = pos$CHR,
  GeneStart = pos$P0,
  GeneEnd = pos$P1,
  TargetID = pos$ID
)

annot<-annot[order(annot$CHROM, annot$GeneStart),]

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/input', recursive = T)
fwrite(annot, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/input/annot.txt', sep='\t', na='NA', quote=F)

##
# eQTL sumstats (CHROM POS A1 A2 Z TargetID N)
##

library(data.table)

cis<-fread('~/oliverpainfel/Data/eQTLGen/full/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt.gz')
cis$GenePos <- NULL
cis$NrCohorts <- NULL
cis$FDR <- NULL
cis$BonferroniP <- NULL
cis$GeneChr <- NULL
cis<-cis[cis$Gene %in% genes,]

eqtl_ss<-data.table(
  CHROM = cis$SNPChr,
  POS = cis$SNPPos,
  A1 = cis$AssessedAllele,
  A2 = cis$OtherAllele,
  Z = cis$Zscore,
  TargetID = cis$Gene,
  N = cis$NrSamples
)

eqtl_ss<-eqtl_ss[order(eqtl_ss$CHROM, eqtl_ss$POS),]

fwrite(eqtl_ss, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/input/eqtl_ss.txt', sep='\t', na='NA', quote=F)

```

***

### Run OTTERS

```{bash}
conda activate otters

# set number of threads to be used
N_THREADS=3

# set up my OTTERS directory and SDPR directory
OTTERS_DIR=/users/k1806347/oliverpainfel/Software/OTTERS
SDPR_DIR=/users/k1806347/oliverpainfel/Software/bin/SDPR

# Load plink
module add plink

# make sure the dynamic libraries of SDPR are not changed (For SDPR)
export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:${SDPR_DIR}/MKL/lib
export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:${SDPR_DIR}/gsl/lib

# prevent automatically using  all available cores on a compute node (For SDPR and PRS-CS)
export MKL_NUM_THREADS=$N_THREADS
export NUMEXPR_NUM_THREADS=$N_THREADS
export OMP_NUM_THREADS=$N_THREADS

# Input for OTTERS STAGE I 
# Annotation File 
exp_anno=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/input/annot.txt
# eQTL summary statistics 
sst_file=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/input/eqtl_ss.txt

# Set LD-clumping threshold in STAGE I
clump_r2=0.99

# STAGE I
# train eQTL weights using P+T, lassosum, SDPR and PRS-CS. 
# It may take several minutes to complete.
for chr in $(seq 22 22); do
  # Genotype data from LD reference panel
  geno_dir=/users/k1806347/oliverpainfel/Data/1KG/FUSION/1000G.EUR.${chr}

  # Set output directory for STAGE I
  mkdir -p /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/output_chr${chr}
  out_dir=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/output_chr${chr}

  sbatch -p neurohack_cpu,cpu --mem 20G -n 3 -t 200 --wrap="
    python3 ${OTTERS_DIR}/training.py \
      --OTTERS_dir=${OTTERS_DIR} \
      --SDPR_dir=${SDPR_DIR} \
      --anno_dir=${exp_anno} \
      --geno_dir=${geno_dir} \
      --sst_file=${sst_file} \
      --out_dir=${out_dir} \
      --chrom=${chr} \
      --r2=${clump_r2} \
      --models=PT,lassosum,SDPR,PRScs \
      --thread=$N_THREADS \
      --lassosum_ld_blocks EUR.hg19"
done
```

```{r}
# Combine output
models <- NULL
for(i in 1:22){
  if(file.exists(paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/output_chr', i,'/lassosum.txt'))){
    for(j in list.files(paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/output_chr', i))){
      models_i_j <-
        fread(
          paste0(
            '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/output_chr',
            i,
            '/',
            j
          )
        )
      names(models_i_j)[names(models_i_j) == 'ES']<-'BETA'
      models_i_j$method <- gsub('.txt','',j)
      models <- rbind(models, models_i_j)
    }
  }
}

# Convert into score files
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/score_files', recursive = T)
for(i in unique(models$TargetID)){
  for(j in unique(models$method[models$TargetID == i])){
    models_i_j <- models[models$TargetID == i & models$method == j,]
    
    weights<-data.table(
      chr_name = models_i_j$CHROM,
      chr_position = models_i_j$POS,
      effect_allele = models_i_j$A1,
      other_allele = models_i_j$A2,
      effect_weight = models_i_j$BETA
    )

    weights <- weights[weights$effect_weight != 0 & !is.na(weights$effect_weight),]
    
    # Save to a .score file
    write.table(
      weights,
      file = paste0(
        "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/score_files/",
        i,
        '.',
        j,
        ".score"
      ),
      sep = "\t",
      quote = FALSE,
      row.names = FALSE)
  }
}
```

***

### Predict expression

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_otters.yaml',
  'score_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list_otters.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_otters.yaml', col.names = F, row.names = F, quote = F)

# Create gwas_list
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1
methods <- c("lassosum", "P0.001", "P0.05", "PRScs", "SDPR")

score_list <- data.table(
  name = paste0(rep(genes, each = length(methods)), "_otters_", rep(methods, times = length(genes))),
  path = paste0(
    "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/score_files/",
    rep(genes, each = length(methods)),
    ".",
    rep(methods, times = length(genes)),
    ".score"
  ),
  label = paste0(rep(genes, each = length(methods)), "_otters_", rep(methods, times = length(genes)))
)

write.table(
  score_list,
  '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list_otters.txt',
  col.names = T,
  row.names = F,
  quote = F,
  sep = ' '
)

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_otters.yaml \
  target_pgs -n 
```

***

### Evaluate PGS

```{r}
# Compare predictions from GenoPred and OTTERS directly
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

###
# GenoPred
###

setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
pgs_genopred <- read_pgs(config = config, pop = 'EUR', pseudo_only = T)$gtex_v8$EUR

# Get a list of genes
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

###
# OTTERS
###

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_otters.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
score_files <- list_score_files(config = config)
pgs_otters <- read_pgs(config = config, pop = 'EUR')$gtex_v8$EUR

pgs_otters_new<-list()
for(i in genes){
  pgs_otters_new[[i]] <-
    lapply(pgs_otters[which(grepl(i, names(pgs_otters)))], function(x) {
      x[[1]]
    })
  names(pgs_otters_new[[i]])<-gsub('.*_','',names(pgs_otters_new[[i]]))
}
pgs_otters <- pgs_otters_new
rm(pgs_otters_new)

pgs_all <- list()
pgs_all[['genopred']] <- pgs_genopred
pgs_all[['otters']] <- pgs_otters

###
# Observed
###

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in both_genes){
  for(j in c('genopred', 'otters')){
    pgs_i_j <- pgs_all[[j]][[i]]
    for(k in names(pgs_i_j)){
      obs_resid_i <- obs_resid[, c('ID', i), with = F]
      names(obs_resid_i)<-c('ID','obs')
    
      pgs_i_j_k <- pgs_i_j[[k]]
      pgs_i_j_k$FID<-NULL
      if(ncol(pgs_i_j_k) == 1){
        next
      }
      names(pgs_i_j_k)<-c('ID','pred')
      
      both_i_j <- merge(obs_resid_i, pgs_i_j_k, by = 'ID')
      both_i_j <- both_i_j[complete.cases(both_i_j),]
      
      if(nrow(both_i_j) > 0){
        mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
        
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = coef(mod)[2,1],
            se = coef(mod)[2,2],
            p = coef(mod)[2,4],
            n = nrow(both_i_j),
            pgs_method = k
          ))
      } else {
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = NA,
            se = NA,
            p = NA,
            n = 0,
            pgs_method = k
          ))      
      }
    }
  }
}

library(ggplot2)
library(cowplot)

ggplot(cor_res, aes(x=gene, y=r , fill = interaction(software, pgs_method))) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

ggplot(cor_res[cor_res$pgs_method %in% c('sdpr','SDPR'),], aes(x=gene, y=r , fill = interaction(software, pgs_method))) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")


# Compute medians using aggregate()
medians <- aggregate(r ~ pgs_method + software, data = cor_res, FUN = median, na.rm = TRUE)

counts <- aggregate(r ~ pgs_method + software, data = cor_res, FUN = function(x) sum(!is.na(x)))
names(counts)[3] <- "n"
counts$group <- interaction(counts$software, counts$pgs_method)

ggplot(cor_res, aes(x = interaction(software, pgs_method), y = r, fill = interaction(software, pgs_method))) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = interaction(software, pgs_method), y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  geom_text(
    data = counts,
    aes(x = group, y = 0.55, label = paste0("n=", n)),  # adjust y as needed
    inherit.aes = FALSE,
    size = 3
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

# Check the correlation between the predicted expression levels
pred_cor<-list()
for(i in genes){
  pgs_all_i_genopred <- pgs_all[['genopred']][[i]]
  pgs_all_i_genopred_tmp <- do.call(cbind, lapply(pgs_all_i_genopred, function(x) {
    if(ncol(x) == 3){
      tmp <- x[,3]
    } else {
      tmp <- NA
    }
    return(tmp)
  }))
  names(pgs_all_i_genopred_tmp)<-paste0('genopred_', names(pgs_all_i_genopred))
  
  pgs_all_i_otters <- pgs_all[['otters']][[i]]
  pgs_all_i_otters_tmp <- do.call(cbind, lapply(pgs_all_i_otters, function(x) {
    tmp <- x[,3]
  }))
  names(pgs_all_i_otters_tmp)<-paste0('otters_', names(pgs_all_i_otters))

  pred_cor[[i]] <- cor(cbind(pgs_all_i_genopred_tmp, pgs_all_i_otters_tmp))
}

# Stack all matrices into an array
bad_entries <- sapply(pred_cor, function(x) !is.matrix(x) || any(dim(x) != dim(pred_cor[[1]])))
names(pred_cor)[bad_entries]
pred_cor_clean <- pred_cor[!bad_entries]
cor_array <- simplify2array(pred_cor_clean)

# Take the mean across the third dimension (genes)
avg_cor <- apply(cor_array, c(1, 2), mean, na.rm = TRUE)
      
library(reshape2)
library(ggplot2)

melted_cor <- melt(avg_cor)

ggplot(melted_cor, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", limits = c(0.5, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "", fill = "Avg Cor")
```

As expected the results of OTTERS and GenoPred are highly concordant.

***

## EXPRESSO

### Published models

```{bash}
# Download EXPRESSO repo
cd /users/k1806347/oliverpainfel/Software
git clone https://github.com/LidaWangPSU/EXPRESSO.git

# Examine published wieghts from eQTLGen
cd /users/k1806347/oliverpainfel/Software/EXPRESSO/trained_model
unzip EXPRESSO_Whole_Blood_eQTLGen_weights_hg19.txt.zip
```

```{r}
library(data.table)
weights<-fread('/users/k1806347/oliverpainfel/Software/EXPRESSO/trained_model/EXPRESSO_Whole_Blood_eQTLGen_weights_hg19.txt')

length(unique(weights$gene_id))
# 12280

nrow(weights)
# 540457

weights$heritable <- NULL

# Split the SNP column into components
snp_parts <- do.call(rbind, strsplit(weights$snp, "_"))
weights$chr_name <- snp_parts[,1]
weights$chr_position <- snp_parts[,2]
weights$effect_allele <- snp_parts[,3]
weights$other_allele <- snp_parts[,4]

# Rename 'weight' column to match PGS format
weights$effect_weight <- weights$weight

dir.create('~/oliverpainfel/Data/EXPRESSO/published_models', recursive = T)
for(i in unique(weights$gene_id)){
  # Select and reorder columns
  pgs_score_i<-weights[weights$gene_id == i,]
  pgs_score <- pgs_score_i[, c("chr_name", "chr_position", "effect_allele", "other_allele", "effect_weight")]

  # Save to a .score file
  write.table(
    pgs_score,
    file = paste0(
      "~/oliverpainfel/Data/EXPRESSO/published_models/eQTLGen_",
      i,
      ".score"
    ),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE)
}
```

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

# Create config file
conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml',
  'gwas_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  'score_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list.txt',
  "pgs_methods: ['ptclump','sbayesrc','sbayesr','quickprs','lassosum','ldpred2','prscs','dbslmm','megaprs']",
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']",
  "lassosum_pseudo_only: true",
  "megaprs_pseudo_only: true",
  "prscs_phi: ['auto']",
  "dbslmm_h2f: ['1']",
  "cores_prep_pgs: 1",
  "quickprs_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/quickprs/hm3",
  "sbayesrc_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/sbayesrc/hm3"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml', col.names = F, row.names = F, quote = F)

conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso_pub.yaml',
  'score_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso_pub.yaml', col.names = F, row.names = F, quote = F)

# Create gwas_list
genes <- gsub('.gz', '', list.files('~/oliverpainfel/Data/eQTLGen/full/cis_trans_combined'))

score_list <- data.table(
  name = paste0(genes,'_expresso_pub'),
  path = paste0(
      "/users/k1806347/oliverpainfel/Data/EXPRESSO/published_models/eQTLGen_",
      genes,
      ".score"),
  label = paste0(genes,'_expresso_pub'))

write.table(score_list, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list.txt', col.names = T, row.names = F, quote = F, sep = ' ')

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso_pub.yaml \
  target_pgs -n 
```

This doesn't work as they did not restrict to HapMap3 variants, so poor overlap with GenoPred reference. Run EXPRESSO to generate weights using data harmonised with the GenoPred reference to improve comparability to other approaches. We should probably consider expanding the SNP-list at some point to see if it makes an improvement.

***

### Generate models

```{r}
# Prepare input
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

# eQTL sumstats
library(data.table)

cis<-fread('~/oliverpainfel/Data/eQTLGen/full/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt.gz')
cis$GenePos <- NULL
cis$NrCohorts <- NULL
cis$FDR <- NULL
cis$BonferroniP <- NULL
cis$GeneChr <- NULL

cis<-cis[cis$SNPChr == 22,]

# Read in EUR MAF
frq<-NULL
for(i in 1:22){
  frq<-rbind(frq, fread(paste0('/users/k1806347/oliverpainfel/Data/1KG/FUSION//1000G.EUR.',i,'.frq')))
}
names(frq)[names(frq) == 'MAF']<-'FREQ'

setkey(cis, Gene)
setkey(frq, SNP)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/sumstats/', recursive = T)

genes <- unique(cis$Gene)
for(i in genes[1:50]){
  print(i)
  cis_i<-cis[.(i)]
  cis_i <- cis[cis$Gene == i,]

  names(cis_i)[names(cis_i) == 'AssessedAllele']<-'A1'
  names(cis_i)[names(cis_i) == 'OtherAllele']<-'A2'
  
  # Insert FREQ from EUR reference
  # There don't seem to be any strand flips
  frq_tmp<-frq[.(cis_i$SNP)]
  cis_i_match<-merge(cis_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by=c('SNP','A1','A2'))
  cis_i_flip<-merge(cis_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by.x=c('SNP','A1','A2'), by.y=c('SNP','A2','A1'))
  cis_i_flip$FREQ<-1-cis_i_flip$FREQ
  cis_i<-rbind(cis_i_match, cis_i_flip)
  
  # Approximate BETA, SE, and P
  cis_i$BETA<-cis_i$Z/sqrt((2*cis_i$FREQ)*(1-cis_i$FREQ)*(cis_i$NrSamples+(cis_i$Z^2)))
  cis_i$SE<-abs(cis_i$BETA)/abs(cis_i$Z)
  
  cis_i <- cis_i[, c('SNPChr','SNPPos','SNP','A1','A2','BETA','SE','Pvalue','NrSamples','GeneSymbol','Gene','FREQ')]
  names(cis_i) <- c('CHR','BP','SNP','A1','A2','BETA','SE','P','N','SYMBOL','GENE','FREQ')

  cis_i<-cis_i[,c('GENE','SYMBOL','CHR','BP','A1','A2','FREQ','BETA','N','SNP'), with=F]
  names(cis_i)<-c('gene','symbol','chr','pos','ref','alt','af','beta.est','n','snp_id')
  
  fwrite(cis_i, paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/sumstats/', i,'.txt'), sep='\t', na='NA', quote=F)
}

# Annotation file
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

pvar <- read_pvar('resources/data/ref/ref.chr')
pvar <- pvar[, c('SNP','CHR','BP','A1','A2'), with=F]
names(pvar)<-c('snp','chr','pos','ref','alt')
pvar$overlap<-0

fwrite(pvar, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/annot.txt', sep='\t', na='NA', quote=F)

```

Make an R script to run EXPRESSO for each gene in parallel.

```{r}
#!/usr/bin/env Rscript
args <- commandArgs(trailingOnly = TRUE)
gene <- args[1]

# load libraries & objects
library(EXPRESSO)
library(fast.lasso)
library(rareGWAMA)
library(MASS)
library(data.table)

# assume annoFile, windowFile, refFile are set here or via a config
refFile <- "/users/k1806347/oliverpainfel/Data/1KG/FUSION/1000G.EUR.22"
annoFile <- "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/annot.txt"
sumstatFile<- "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/eqtl_sumstats.txt"

windowFile <- c( "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_1mb_chr22.txt",
                 "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_250kb_chr22.txt",
                 "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_loop_chr22.txt",
                 "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_domain_chr22.txt",
                 "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_tad_chr22.txt",
                 "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_pchic_chr22.txt")

indir <- paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/sumstats/', gene,'.txt')
outdir  <- paste0("/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/output/", gene)

EXPRESSO(
  indir,
  annoFile,
  windowFile,
  refFile,
  outdir,
  minMaf  = 0.05,
  maxIter = 1000,
  append  = FALSE
)
```

```{bash}
#!/bin/bash
#SBATCH --job-name=EXPRESSO
#SBATCH --output=~/logs/expresso_%A_%a.out
#SBATCH --error=~/logs/expresso_%A_%a.err
#SBATCH --time=12:00:00
#SBATCH --mem=10G
#SBATCH --ntasks=1
#SBATCH --partition=neurohack_cpu
#SBATCH --array=1-50%50    # adjust 1–N to your #genes, %50 = max concurrent

# build the gene list into a Bash array
mapfile -t GENES < ~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt

# SLURM_ARRAY_TASK_ID runs from 1..N
gene=${GENES[$SLURM_ARRAY_TASK_ID-1]}

# call the R wrapper
./run_expresso.R "$gene"
```

```{bash}
cd /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO
chmod a+x /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/run_expresso.R
chmod a+x /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/run_expresso.sh
sbatch run_expresso.sh
```

Format into score files for GenoPred.

```{r}
library(data.table)

genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/score_files', recursive = T)
for(model in c('mse','pseudo')){
  for(i in genes){
    weights <-
      fread(
        paste0(
          '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/output/',
          i,
          '-weight.', model, '.txt'
        ), fill = T)
    
    weights <- weights[, c('snp','weight'), with = F]
    
    # Insert A1 and A2 from eqtl sumstats (not the ref)
    eqtl_ss<-fread(paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/sumstats/', i,'.txt'))
    
    weights <- merge(weights, eqtl_ss[,c('snp_id','ref','alt'), with = F], by.x = 'snp', by.y = 'snp_id', all.x = T)
    
    score_file<-data.table(
      rsID = weights$snp,
      effect_allele = weights$ref,
      other_allele = weights$alt,
      effect_weight = weights$weight
    )

    # Save to a .score file
    write.table(
      score_file,
      file = paste0(
        "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/score_files/",
        i,
        '.',
        model,
        ".score"
      ),
      sep = "\t",
      quote = FALSE,
      row.names = FALSE)
  }
}
```

***

### Predict expression

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso.yaml',
  'score_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list_expresso.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso.yaml', col.names = F, row.names = F, quote = F)

# Create gwas_list
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1
methods <- c("mse","pseudo")

score_list <- data.table(
  name = paste0(rep(genes, each = length(methods)), "_expresso_", rep(methods, times = length(genes))),
  path = paste0(
    "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/score_files/",
    rep(genes, each = length(methods)),
    ".",
    rep(methods, times = length(genes)),
    ".score"
  ),
  label = paste0(rep(genes, each = length(methods)), "_expresso_", rep(methods, times = length(genes)))
)

write.table(
  score_list,
  '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list_expresso.txt',
  col.names = T,
  row.names = F,
  quote = F,
  sep = ' '
)

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso.yaml \
  target_pgs -n 
```

***

### Evaluate PGS

```{r}
# Compare predictions from GenoPred and EXPRESSO directly
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

###
# GenoPred
###

setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
pgs_genopred <- read_pgs(config = config, pop = 'EUR')$gtex_v8$EUR

# Split ptclump scores into seperate methods
pgs_genopred <- lapply(pgs_genopred, function(x){
  for(i in names(x$ptclump)[-1:-2]){
    x[[paste0('ptclump_', gsub('.*_', '', i))]] <- x$ptclump[, c('FID','IID',i), with=F]
  }
  x$ptclump <-NULL
  return(x)
})

# Get a list of genes
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

###
# EXPRESSO
###

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
score_files <- list_score_files(config = config)
pgs_expresso <- read_pgs(config = config, pop = 'EUR')$gtex_v8$EUR

pgs_expresso_new<-list()
for(i in genes){
  pgs_expresso_new[[i]] <-
    lapply(pgs_expresso[which(grepl(i, names(pgs_expresso)))], function(x) {
      x[[1]]
    })
  names(pgs_expresso_new[[i]])<-gsub('.*_','',names(pgs_expresso_new[[i]]))
}
pgs_expresso <- pgs_expresso_new
rm(pgs_expresso_new)

pgs_all <- list()
pgs_all[['genopred']] <- pgs_genopred
pgs_all[['expresso']] <- pgs_expresso

###
# Observed
###

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in both_genes){
  for(j in c('genopred', 'expresso')){
    pgs_i_j <- pgs_all[[j]][[i]]
    for(k in names(pgs_i_j)){
      obs_resid_i <- obs_resid[, c('ID', i), with = F]
      names(obs_resid_i)<-c('ID','obs')
    
      pgs_i_j_k <- pgs_i_j[[k]]
      pgs_i_j_k$FID<-NULL
      if(ncol(pgs_i_j_k) == 1){
        next
      }
      names(pgs_i_j_k)<-c('ID','pred')
      
      both_i_j <- merge(obs_resid_i, pgs_i_j_k, by = 'ID')
      both_i_j <- both_i_j[complete.cases(both_i_j),]
      
      if(nrow(both_i_j) > 0){
        mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
        
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = coef(mod)[2,1],
            se = coef(mod)[2,2],
            p = coef(mod)[2,4],
            n = nrow(both_i_j),
            pgs_method = k
          ))
      } else {
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = NA,
            se = NA,
            p = NA,
            n = 0,
            pgs_method = k
          ))      
      }
    }
  }
}

library(ggplot2)
library(cowplot)

ggplot(cor_res, aes(x=gene, y=r , fill = interaction(software, pgs_method))) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

ggplot(cor_res[cor_res$pgs_method %in% c('pseudo','mse','prscs'),], aes(x=gene, y=r , fill = interaction(software, pgs_method))) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

# Compute medians using aggregate()
cor_res_comp <- cor_res[complete.cases(cor_res),]
gene_n<-data.frame(table(cor_res_comp$gene))
cor_res_comp <- cor_res_comp[cor_res_comp$gene %in% gene_n$Var1[gene_n$Freq == 12],]

medians <- aggregate(r ~ pgs_method + software, data = cor_res_comp, FUN = median, na.rm = TRUE)

counts <- aggregate(r ~ pgs_method + software, data = cor_res_comp, FUN = function(x) sum(!is.na(x)))
names(counts)[3] <- "n"
counts$group <- interaction(counts$software, counts$pgs_method)

ggplot(cor_res_comp, aes(x = interaction(software, pgs_method), y = r, fill = interaction(software, pgs_method))) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = interaction(software, pgs_method), y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  geom_text(
    data = counts,
    aes(x = group, y = 0.55, label = paste0("n=", n)),  # adjust y as needed
    inherit.aes = FALSE,
    size = 3
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

# Check the correlation between the predicted expression levels
pred_cor<-list()
for(i in genes){
  pgs_all_i_genopred <- pgs_all[['genopred']][[i]]
  pgs_all_i_genopred_tmp <- do.call(cbind, lapply(pgs_all_i_genopred, function(x) {
    if(ncol(x) == 3){
      tmp <- x[,3]
    } else {
      tmp <- NA
    }
    return(tmp)
  }))
  names(pgs_all_i_genopred_tmp)<-paste0('genopred_', names(pgs_all_i_genopred))
  
  pgs_all_i_expresso <- pgs_all[['expresso']][[i]]
  pgs_all_i_expresso_tmp <- do.call(cbind, lapply(pgs_all_i_expresso, function(x) {
    tmp <- x[,3]
  }))
  names(pgs_all_i_expresso_tmp)<-paste0('expresso_', names(pgs_all_i_expresso))

  pred_cor[[i]] <- cor(cbind(pgs_all_i_genopred_tmp, pgs_all_i_expresso_tmp))
}

# Stack all matrices into an array
bad_entries <- sapply(pred_cor, function(x) !is.matrix(x) || any(dim(x) != dim(pred_cor[[1]])))
names(pred_cor)[bad_entries]
pred_cor_clean <- pred_cor[!bad_entries]
cor_array <- simplify2array(pred_cor_clean)

# Take the mean across the third dimension (genes)
avg_cor <- apply(cor_array, c(1, 2), mean, na.rm = TRUE)
      
library(reshape2)
library(ggplot2)

melted_cor <- melt(avg_cor)

ggplot(melted_cor, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", limits = c(0.5, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "", fill = "Avg Cor")
```

***


## SUMMIT-FA

No point is looking at SUMMIT models since these superseed SUMMIT. Download published models.

***

### Download format models

Download SUMMIT-FA_Models.zip from https://osf.io/xnkcz. I have stored this file in: /users/k1806347/oliverpainfel/Data/SUMMIT-FA

```{bash}
cd /users/k1806347/oliverpainfel/Data/SUMMIT-FA
unzip SUMMIT-FA_Models.zip
```

```{r}
##
# Reformat models as score files
##

# Get the list of genes tested above
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/SUMMIT-FA/score_files', recursive = T)
for(i in genes){
  if(file.exists(paste0('/users/k1806347/oliverpainfel/Data/SUMMIT-FA/WeightsMACIE/TWASPreppedWeights/SUMMIT-M.Whole-Blood.',i,'.wgt.rds'))){
  
    weights_i <- readRDS(paste0('/users/k1806347/oliverpainfel/Data/SUMMIT-FA/WeightsMACIE/TWASPreppedWeights/SUMMIT-M.Whole-Blood.',i,'.wgt.rds'))
    for(j in names(weights_i)){
      names(weights_i[[j]][[3]])[names(weights_i[[j]][[3]]) == 'a1'] <- 'allele1'
      names(weights_i[[j]][[3]])[names(weights_i[[j]][[3]]) == 'a2'] <- 'allele2'

      weights_i_j <- data.table(rsID = weights_i[[j]][[3]]$SNP,
                                effect_allele = weights_i[[j]][[3]]$allele1,
                                other_allele = weights_i[[j]][[3]]$allele2,
                                effect_weight = as.numeric(weights_i[[j]][[2]]))
  
      if(ncol(weights_i_j) != 4){
        print(i)
      }
      
      if(var(weights_i_j$effect_weight) != 0){
        # Save to a .score file
        write.table(
          weights_i_j,
          file = paste0(
            "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/SUMMIT-FA/score_files/",
            i,
            '.',
            j,
            ".score"
          ),
          sep = "\t",
          quote = FALSE,
          row.names = FALSE)
      }
    }
  }
}
```

***

### Predict expression

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_summitfa.yaml',
  'score_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list_summitfa.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_summitfa.yaml', col.names = F, row.names = F, quote = F)

# Create gwas_list
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1
methods <- c("mse","pseudo")

score_file_list <- list.files('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/SUMMIT-FA/score_files')
score_file_list <- score_file_list[grepl('.score', score_file_list)]
score_file_list <- score_file_list[grepl(paste0(genes, collapse='|'), score_file_list)]

score_list <- data.table(
  name = gsub('\\.','_',gsub('.score', '', score_file_list)),
  path = paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/SUMMIT-FA/score_files/', score_file_list),
  label = gsub('\\.','_',gsub('.score', '', score_file_list)))

write.table(
  score_list,
  '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list_summitfa.txt',
  col.names = T,
  row.names = F,
  quote = F,
  sep = ' '
)

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_summitfa.yaml \
  target_pgs -n 
```

### Evaluate PGS

```{r}
# Compare predictions from GenoPred and EXPRESSO directly
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

###
# GenoPred
###

setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
pgs_genopred <- read_pgs(config = config, pop = 'EUR')$gtex_v8$EUR

# Split ptclump scores into seperate methods
pgs_genopred <- lapply(pgs_genopred, function(x){
  for(i in names(x$ptclump)[-1:-2]){
    x[[paste0('ptclump_', gsub('.*_', '', i))]] <- x$ptclump[, c('FID','IID',i), with=F]
  }
  x$ptclump <-NULL
  return(x)
})

# Get a list of genes
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

###
# EXPRESSO
###

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_summitfa.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
score_files <- list_score_files(config = config)
pgs_summitfa <- read_pgs(config = config, pop = 'EUR')$gtex_v8$EUR

pgs_summitfa_new<-list()
for(i in genes){
  pgs_summitfa_new[[i]] <-
    lapply(pgs_summitfa[which(grepl(i, names(pgs_summitfa)))], function(x) {
      x[[1]]
    })
  names(pgs_summitfa_new[[i]])<-gsub('.*_','',names(pgs_summitfa_new[[i]]))
}
pgs_summitfa <- pgs_summitfa_new
rm(pgs_summitfa_new)

pgs_all <- list()
pgs_all[['genopred']] <- pgs_genopred
pgs_all[['summitfa']] <- pgs_summitfa

###
# Observed
###

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in both_genes){
  for(j in c('genopred', 'summitfa')){
    pgs_i_j <- pgs_all[[j]][[i]]
    for(k in names(pgs_i_j)){
      obs_resid_i <- obs_resid[, c('ID', i), with = F]
      names(obs_resid_i)<-c('ID','obs')
    
      pgs_i_j_k <- pgs_i_j[[k]]
      pgs_i_j_k$FID<-NULL
      if(ncol(pgs_i_j_k) == 1){
        next
      }
      names(pgs_i_j_k)<-c('ID','pred')
      
      both_i_j <- merge(obs_resid_i, pgs_i_j_k, by = 'ID')
      both_i_j <- both_i_j[complete.cases(both_i_j),]
      
      if(nrow(both_i_j) > 0){
        mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
        
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = coef(mod)[2,1],
            se = coef(mod)[2,2],
            p = coef(mod)[2,4],
            n = nrow(both_i_j),
            pgs_method = k
          ))
      } else {
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = NA,
            se = NA,
            p = NA,
            n = 0,
            pgs_method = k
          ))      
      }
    }
  }
}

library(ggplot2)
library(cowplot)

ggplot(cor_res, aes(x=gene, y=r , fill = interaction(software, pgs_method))) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

ggplot(cor_res[cor_res$pgs_method == 'prscs' | cor_res$software == 'summitfa',], aes(x=gene, y=r , fill = interaction(software, pgs_method))) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

# Compute medians using aggregate()
cor_res_comp <- cor_res[complete.cases(cor_res),]
gene_n<-data.frame(table(cor_res_comp$gene))
cor_res_comp <- cor_res_comp[cor_res_comp$gene %in% gene_n$Var1[gene_n$Freq == 19],]

medians <- aggregate(r ~ pgs_method + software, data = cor_res_comp, FUN = median, na.rm = TRUE)

counts <- aggregate(r ~ pgs_method + software, data = cor_res_comp, FUN = function(x) sum(!is.na(x)))
names(counts)[3] <- "n"
counts$group <- interaction(counts$software, counts$pgs_method)

ggplot(cor_res_comp, aes(x = interaction(software, pgs_method), y = r, fill = interaction(software, pgs_method))) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = interaction(software, pgs_method), y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  geom_text(
    data = counts,
    aes(x = group, y = 0.55, label = paste0("n=", n)),  # adjust y as needed
    inherit.aes = FALSE,
    size = 3
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

# Check the correlation between the predicted expression levels
pred_cor<-list()
for(i in genes){
  pgs_all_i_genopred <- pgs_all[['genopred']][[i]]
  pgs_all_i_genopred_tmp <- do.call(cbind, lapply(pgs_all_i_genopred, function(x) {
    if(ncol(x) == 3){
      tmp <- x[,3]
    } else {
      tmp <- NA
    }
    return(tmp)
  }))
  names(pgs_all_i_genopred_tmp)<-paste0('genopred_', names(pgs_all_i_genopred))
  
  pgs_all_i_summitfa <- pgs_all[['summitfa']][[i]]
  pgs_all_i_summitfa_tmp <- do.call(cbind, lapply(pgs_all_i_summitfa, function(x) {
    tmp <- x[,3]
  }))
  names(pgs_all_i_summitfa_tmp)<-paste0('summitfa_', names(pgs_all_i_summitfa))

  pred_cor[[i]] <- cor(cbind(pgs_all_i_genopred_tmp, pgs_all_i_summitfa_tmp))
}

# Stack all matrices into an array
bad_entries <- sapply(pred_cor, function(x) !is.matrix(x) || any(dim(x) != dim(pred_cor[[1]])))
names(pred_cor)[bad_entries]
pred_cor_clean <- pred_cor[!bad_entries]
cor_array <- simplify2array(pred_cor_clean)

# Take the mean across the third dimension (genes)
avg_cor <- apply(cor_array, c(1, 2), mean, na.rm = TRUE)
      
library(reshape2)
library(ggplot2)

melted_cor <- melt(avg_cor)

ggplot(melted_cor, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", limits = c(0.5, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "", fill = "Avg Cor")
```

Several models within SUMMIT-FA actually seem to perform quite well. 

***

## Compare GenoPred, EXPRESSO, and SUMMIT-FA

We don't need to include OTTERS as all the methods it runs are in GenoPred, and we have already shown they are consistent.

```{r}
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

###
# GenoPred
###

setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
pgs_genopred <- read_pgs(config = config, pop = 'EUR')$gtex_v8$EUR

# Split ptclump scores into seperate methods
pgs_genopred <- lapply(pgs_genopred, function(x){
  for(i in names(x$ptclump)[-1:-2]){
    x[[paste0('ptclump_', gsub('.*_', '', i))]] <- x$ptclump[, c('FID','IID',i), with=F]
  }
  x$ptclump <-NULL
  return(x)
})

# Get a list of genes
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

###
# EXPRESSO
###

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
score_files <- list_score_files(config = config)
pgs_expresso <- read_pgs(config = config, pop = 'EUR')$gtex_v8$EUR

pgs_expresso_new<-list()
for(i in genes){
  pgs_expresso_new[[i]] <-
    lapply(pgs_expresso[which(grepl(i, names(pgs_expresso)))], function(x) {
      x[[1]]
    })
  names(pgs_expresso_new[[i]])<-gsub('.*_','',names(pgs_expresso_new[[i]]))
}
pgs_expresso <- pgs_expresso_new
rm(pgs_expresso_new)

###
# SUMMIT-FA
###

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_summitfa.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
score_files <- list_score_files(config = config)
pgs_summitfa <- read_pgs(config = config, pop = 'EUR')$gtex_v8$EUR

pgs_summitfa_new<-list()
for(i in genes){
  pgs_summitfa_new[[i]] <-
    lapply(pgs_summitfa[which(grepl(i, names(pgs_summitfa)))], function(x) {
      x[[1]]
    })
  names(pgs_summitfa_new[[i]])<-gsub('.*_','',names(pgs_summitfa_new[[i]]))
}
pgs_summitfa <- pgs_summitfa_new
rm(pgs_summitfa_new)

pgs_all <- list()
pgs_all[['genopred']] <- pgs_genopred
pgs_all[['summitfa']] <- pgs_summitfa
pgs_all[['expresso']] <- pgs_expresso

###
# Observed
###

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in both_genes){
  for(j in c('genopred', 'expresso', 'summitfa')){
    pgs_i_j <- pgs_all[[j]][[i]]
    for(k in names(pgs_i_j)){
      obs_resid_i <- obs_resid[, c('ID', i), with = F]
      names(obs_resid_i)<-c('ID','obs')
    
      pgs_i_j_k <- pgs_i_j[[k]]
      pgs_i_j_k$FID<-NULL
      if(ncol(pgs_i_j_k) == 1){
        next
      }
      names(pgs_i_j_k)<-c('ID','pred')
      
      both_i_j <- merge(obs_resid_i, pgs_i_j_k, by = 'ID')
      both_i_j <- both_i_j[complete.cases(both_i_j),]
      
      if(nrow(both_i_j) > 0){
        mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
        
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = coef(mod)[2,1],
            se = coef(mod)[2,2],
            p = coef(mod)[2,4],
            n = nrow(both_i_j),
            pgs_method = k
          ))
      } else {
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = NA,
            se = NA,
            p = NA,
            n = 0,
            pgs_method = k
          ))      
      }
    }
  }
}

library(ggplot2)
library(cowplot)

ggplot(cor_res, aes(x=gene, y=r , fill = pgs_method)) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    panel_border() +
    background_grid(major = 'y', minor = 'y') +
    facet_grid(software ~.) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

ggplot(cor_res, aes(x = gene, y = r, fill = software)) +
  geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
  geom_boxplot(width = 0.6, position = position_dodge(width = 0.9), outlier.shape = NA) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  panel_border() +
  background_grid(major = 'y', minor = 'y') +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )

# Compute medians using aggregate()
cor_res_comp <- cor_res[complete.cases(cor_res),]
gene_n<-data.frame(table(cor_res_comp$gene))
cor_res_comp <- cor_res_comp[cor_res_comp$gene %in% gene_n$Var1[gene_n$Freq == length(unique(cor_res$pgs_method))],]

medians <- aggregate(r ~ pgs_method + software, data = cor_res_comp, FUN = median, na.rm = TRUE)

counts <- aggregate(r ~ pgs_method + software, data = cor_res_comp, FUN = function(x) sum(!is.na(x)))
names(counts)[3] <- "n"
counts$pgs_method <- counts$pgs_method

ggplot(cor_res_comp, aes(x = pgs_method, y = r, fill = pgs_method)) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = pgs_method, y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  geom_text(
    data = counts,
    aes(x = pgs_method, y = 0.55, label = paste0("n=", n)),  # adjust y as needed
    inherit.aes = FALSE,
    size = 3
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  panel_border() +
  background_grid(major = 'y', minor = 'y') +
  facet_grid(. ~ software, scales = 'free_x', space = 'free_x') +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

# Check the correlation between the predicted expression levels
pred_cor<-list()
for(i in genes){
  pgs_all_i_genopred <- pgs_all[['genopred']][[i]]
  pgs_all_i_genopred_tmp <- do.call(cbind, lapply(pgs_all_i_genopred, function(x) {
    if(ncol(x) == 3){
      tmp <- x[,3]
    } else {
      tmp <- NA
    }
    return(tmp)
  }))
  names(pgs_all_i_genopred_tmp)<-paste0('genopred_', names(pgs_all_i_genopred))
  
  pgs_all_i_summitfa <- pgs_all[['summitfa']][[i]]
  pgs_all_i_summitfa_tmp <- do.call(cbind, lapply(pgs_all_i_summitfa, function(x) {
    tmp <- x[,3]
  }))
  names(pgs_all_i_summitfa_tmp)<-paste0('summitfa_', names(pgs_all_i_summitfa))

  pred_cor[[i]] <- cor(cbind(pgs_all_i_genopred_tmp, pgs_all_i_summitfa_tmp))
}

# Stack all matrices into an array
bad_entries <- sapply(pred_cor, function(x) !is.matrix(x) || any(dim(x) != dim(pred_cor[[1]])))
names(pred_cor)[bad_entries]
pred_cor_clean <- pred_cor[!bad_entries]
cor_array <- simplify2array(pred_cor_clean)

# Take the mean across the third dimension (genes)
avg_cor <- apply(cor_array, c(1, 2), mean, na.rm = TRUE)
      
library(reshape2)
library(ggplot2)

melted_cor <- melt(avg_cor)

ggplot(melted_cor, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", limits = c(0.5, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "", fill = "Avg Cor")
```

***

## Ensemble approach

We want to see whether an elastic net model combining predicted expression across methods improves prediction over the top model. We can use one of the model_builder scripts to achieve this.

```{r}
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# List score files
score_files <- list_score_files(config = config)

# Get a list of genes
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

for(i in both_genes){
  score_files_i <- score_files[score_files$name == i,]
  
  # Split ptclump by p-value threshold
  ptclump_i <-
    fread(paste0(
      outdir,
      '/gtex_v8/pgs/EUR/ptclump/',
      i,
      '/gtex_v8-',
      i,
      '-EUR.profiles'
    ))
  
  for(j in names(ptclump_i)[-1:-2]){
    ptclump_i_j <- ptclump_i[, c('FID','IID',j), with=F]
    score_files_i <- rbind(
      score_files_i,
      data.frame(
        name = i,
        method = paste0('ptclump_', gsub('.*_','',j))
      ))
    
    dir.create(paste0(outdir, '/gtex_v8/pgs/EUR/ptclump_', gsub('.*_','',j),'/',i), recursive = T)
    
    fwrite(
      ptclump_i_j,
      paste0(
        outdir,
        '/gtex_v8/pgs/EUR/ptclump_', gsub('.*_','',j),'/',
        i,
        '/gtex_v8-',
        i,
        '-EUR.profiles'
      ),
      sep = ' ',
      na = NA,
      quote=F)
  }
  
  score_files_i <- score_files_i[score_files_i$method != 'ptclump',]
  
  score_files_i$predictor <- paste0(
      outdir,
      '/gtex_v8/pgs/EUR/',
      score_files_i$method,
      '/',
      i,
      '/gtex_v8-',
      i,
      '-EUR.profiles'
    )
  
  score_files_i$group <- score_files_i$method
  score_files_i <- score_files_i[, c('predictor', 'group'), with = F]
  
  dir.create(paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/ensemble_test/', i), recursive = T)
  write.table(score_files_i, paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/ensemble_test/', i,'/predictor_list.txt'), col.names=T, row.names=F, quote=F)
  
  obs_resid_i <- obs_resid[, c('ID',i), with=F]
  obs_resid_i$FID <- obs_resid_i$ID
  obs_resid_i$IID <- obs_resid_i$ID
  obs_resid_i<-obs_resid_i[, c('FID','IID',i), with=F]
  
  write.table(obs_resid_i, paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/ensemble_test/', i,'/outcome.txt'), col.names=T, row.names=F, quote=F)
}

write.table(both_genes, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/ensemble_test/gene_list.txt', col.names=F, row.names = F, quote=F)

```

```{bash}
cd /users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline
conda activate model_builder

for gene in $(cat /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/ensemble_test/gene_list.txt); do
  if [ ! -f "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/ensemble_test/${gene}/res.pred_comp.txt" ]; then
    sbatch --mem 10G -n 1 --exclude=erc-hpc-comp058 -p neurohack_cpu,interruptible_cpu -t 1:00:00 --wrap="Rscript ../Scripts/model_builder/model_builder.R \
      --outcome /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/ensemble_test/${gene}/outcome.txt \
      --predictors /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/ensemble_test/${gene}/predictor_list.txt \
      --out /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/ensemble_test/${gene}/res \
      --n_core 1"
  fi
done

```

```{r}
# Read in the results see average performance of each model vs the elastic net.
genes <- fread('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/ensemble_test/gene_list.txt', header=F)$V1

pred_eval <- NULL
for(i in genes){
  pred_eval_i <- fread(paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/ensemble_test/', i,'/res.pred_eval.txt'))
  pred_eval_i$Gene <- i
  pred_eval<-rbind(pred_eval, pred_eval_i)
}

library(ggplot2)
library(cowplot)

ggplot(pred_eval, aes(x=Gene, y=R , fill = Group)) +
    geom_errorbar(aes(ymin = R - SE, ymax = R + SE),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

medians <- aggregate(R ~ Group, data = pred_eval, FUN = median, na.rm = TRUE)

ggplot(pred_eval, aes(x = Group, y = R, fill = Group)) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = Group, y = 0.65, label = round(R, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

# Having a brief look, it appears the target sample size is too small. It is smaller than they report in the SUMMIT study interestingly.
# Probably best to move over to using the eQTLGen sumstats that exclude GTEx entirely, and then can evaluate in all 
```










***

```{bash}
mkdir /users/k1806347/oliverpainfel/Data/AdipoExpress
cd /users/k1806347/oliverpainfel/Data/AdipoExpress
wget https://zenodo.org/records/13845120/files/marginal_byChr_EURonly.tar.gz?download=1
mv 'marginal_byChr_EURonly.tar.gz?download=1' marginal_byChr_EURonly.tar.gz
tar -xvzf marginal_byChr_EURonly.tar.gz
```

***

# eQTLGen excl. GTEx

```{r, eval=F, echo=T}
library(data.table)

# Read in relevent columns from the sumstats
sumstats<-fread('/users/k1806347/oliverpainfel/Data/eQTLGen/excl_GTEx/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-WoGTEx.small.txt.gz')

sumstats<-sumstats[sumstats$SNPChr == 22,]

# Extract data for each gene
genes<-unique(sumstats$ProbeName)

# Read in EUR MAF
frq<-NULL
for(i in 1:22){
  frq<-rbind(frq, fread(paste0('/users/k1806347/oliverpainfel/Data/1KG/FUSION//1000G.EUR.',i,'.frq')))
}
names(frq)[names(frq) == 'MAF']<-'FREQ'

setkey(sumstats, ProbeName)
setkey(frq, SNP)

# Process eQTL sumstats for each gene
dir.create('~/oliverpainfel/Data/eQTLGen/excl_gtex/cis_only', recursive = T)
for(i in 1:50){
  print(i)
  
  tmp<-sumstats[.(genes[i])]
  
  # Create A1 and A2 columns and update column names
  tmp$A1<-gsub('.*/','',tmp$SNPType)
  tmp$A2<-gsub('/.*','',tmp$SNPType)
  tmp$A2[tmp$A1 != tmp$AlleleAssessed]<-gsub('.*/','',tmp$SNPType[tmp$A1 != tmp$AlleleAssessed])
  tmp$A1[tmp$A1 != tmp$AlleleAssessed]<-gsub('/.*','',tmp$SNPType[tmp$A1 != tmp$AlleleAssessed])

  tmp<-tmp[,c('ProbeName','SNPName','SNPChr','SNPChrPos', 'A1', 'A2','PValue','OverallZScore','SumNumberOfSamples'), with=F]
  names(tmp)<-c('Gene','SNP','CHR','BP','A1','A2','P','Z','N')

  frq_tmp<-frq[.(tmp$SNP)]
  
  # Insert FREQ from EUR reference
  # There don't seem to be any strand flips
  tmp_match<-merge(tmp, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by=c('SNP','A1','A2'))
  tmp_flip<-merge(tmp, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by.x=c('SNP','A1','A2'), by.y=c('SNP','A2','A1'))
  tmp_flip$FREQ<-1-tmp_flip$FREQ
  tmp<-rbind(tmp_match, tmp_flip)
  
  # Approximate BETA, SE, and P
  tmp$P<-2*pnorm(-abs(tmp$Z))
  tmp$BETA<-tmp$Z/sqrt((2*tmp$FREQ)*(1-tmp$FREQ)*(tmp$N+(tmp$Z^2)))
  tmp$SE<-abs(tmp$BETA)/abs(tmp$Z)
  
  tmp$GENE<-genes[i]
    
  tmp <- tmp[, c('CHR','BP','SNP','A1','A2','BETA','SE','P','N')]

  fwrite(tmp, paste0('~/oliverpainfel/Data/eQTLGen/excl_gtex/cis_only/',i,'.gz'), quote = F, sep = ' ', na= 'NA')
}

write.table(genes[1:50], '~/oliverpainfel/Data/eQTLGen/excl_gtex/cis_only/gene_list.txt', row.names=F, col.names=F, quote=F)


```

***

## GenoPred

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

# Create config file
conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only_excl_gtex.yaml',
  'gwas_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list_cis_only.txt_excl_gtex',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_methods: ['ptclump','quickprs','lassosum','ldpred2','prscs','dbslmm','megaprs','sbayesr','sdpr']",
  "pgs_scaling: ['discrete']",
  "ptclump_pts: ['0.001','0.05']",
  "target_populations: ['EUR']",
  "lassosum_pseudo_only: true",
  "ldpred2_model: ['auto']",
  "megaprs_pseudo_only: true",
  "prscs_phi: ['auto']",
  "dbslmm_h2f: ['1']",
  "h2_method: 'sbayesr'",
  "cores_prep_pgs: 1",
  "quickprs_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/quickprs/hm3",
  "sbayesrc_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/sbayesrc/hm3"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only_excl_gtex.yaml', col.names = F, row.names = F, quote = F)

# Create gwas_list
genes <- fread('~/oliverpainfel/Data/eQTLGen/excl_gtex/cis_only/gene_list.txt', header=F)$V1

gwas_list <- data.table(
  name = paste0(genes, 'excl_gtex'),
  path = paste0('/users/k1806347/oliverpainfel/Data/eQTLGen/excl_gtex/cis_only/', genes,'.gz'),
  population = 'EUR',
  n = NA,
  sampling = NA,
  prevalence = NA,
  mean = NA,
  sd = NA,
  label = genes)

write.table(gwas_list, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list_cis_only_excl_gtex.txt', col.names = T, row.names = F, quote = F, sep = ' ')

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only_excl_gtex.yaml \
  target_pgs -n 
```

***

# To do

- Extend benchmark of methods to 200 genes to get a better indication of performance.
- Run GenoPred using eQTLGen excluding GTEx for testing ensemble approach and performance in AFR

- Make SBayesRC more efficient when using locus specific sumstats
- Compare prediction when using models from SUMMIT-FA, OTTERS and EXPRESSO
- Look into AdipoExpress meta-analysis sumstats
- Look into liver eQTL meta-analysis: https://mohlke.web.unc.edu/data/signal-based-liver-eqtl-meta-analysis-broadaway-et-al-ajhg-2024/
- Brain single cell meta-analysis: https://pmc.ncbi.nlm.nih.gov/articles/PMC11908325/#S22
- SABR - study. New eQTL dataset for AFR ancestry


- What about cross tissue extension?

- I've got some questions for the EXPRESSO authors:
-- What is the PUMAS method? This is a tuning approach, not a PGS method
-- Were the auto models only used for PGS methods, like PRS-CS and LDpred2
-- What LD reference data was used for each method?
-- Did you run SUMMIT yourself, or download the weights. There are multiple models for each gene, how did you combine them?
