---
title: "Comparison with OTTERS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

library(knitr)
library(data.table)

```

***

This document is an attempt of simplifying the message: See what methods most accurately predicts gene expression. We will using the eQTLGen data (excluding GTEx) as the discovery sample, and GTEx v8 as the target sample.

Methods to include are:

- SUMMIT
- SUMMIT-FA
- EXPRESSO
- OTTERS
- BGW-TWAS

***

# Literature review

## SUMMIT and SUMMIT-FA

SUMMIT and SUMMIT-FA are not flexible - they requires a specific set of files with required file names and formats. It is going to be hard to implement. They release the models, so perhaps we can compare the R2 in some independent data.

These methods restrict to HapMap3 variants. They also output models in format resembling FUSION format.

They trained the models using the full eQTLGen sumstats, evaluating the 309 independent individuals in GTEx v8.

We can download the model from SUMMIT and SUMMIT-FA can be download from OSF. Do this and predict expression into independent GTEx v8.

***

## EXPRESSO

This software and the manuscript is more impressive. Really comprehensive comparison of methods using sumstats and individual-level data. EXPRESSO appears to be the best method. They also developed a TWAS-based drug repurposing pipeline. They apply their models to multi-ancestry GWAS using TESLA. They don't include SBayesRC or MegaPRS, which are some of the best performing PGS methods.

The authors position EXPRESSO as just one method for generating the models, and that it can be added in combination with the methods in OTTERS for improved power in TWAS, since OTTERS combines across models using Cauchy distribution. 

I could compare SBayesRC to EXPRESSO, and my pipeline to OTTERS. I could also include multi-ancestry eQTL data if there is any.

***

## BGW-TWAS

Develops approach for modelling cis and trans eQTLs. They use ROSMAP to develop the models, and evaluate in MCADGS.

It is novel in its use of trans-eQTL data, but performance is not great.

It would be interesting to see whether we can go beyond this, by applying some of the PGS methods to genome-wide eQTL data - eQTLGen have released this now.

***

# Analysis Phase 1: GenoPred with cis and trans

Try combining the cis- and trans-eQTL sumstats for eQTLGen, splitting by gene, and then using GenoPred to generate the prediction models, and then evaluate in independent GTEx participants

***

## Prepare eQTL sumstats

```{r}
library(data.table)

cis<-fread('~/oliverpainfel/Data/eQTLGen/full/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt.gz')
cis$GenePos <- NULL
cis$NrCohorts <- NULL
cis$FDR <- NULL
cis$BonferroniP <- NULL
cis$GeneChr <- NULL

trans<-fread('~/oliverpainfel/Data/eQTLGen/full/2018-09-04-trans-eQTLsFDR-CohortInfoRemoved-BonferroniAdded.txt.gz')
trans$GenePos <- NULL
trans$NrCohorts <- NULL
trans$FDR <- NULL
trans$BonferroniP <- NULL
trans$GeneChr <- NULL

# Read in EUR MAF
frq<-NULL
for(i in 1:22){
  frq<-rbind(frq, fread(paste0('/users/k1806347/oliverpainfel/Data/1KG/FUSION//1000G.EUR.',i,'.frq')))
}
names(frq)[names(frq) == 'MAF']<-'FREQ'

setkey(cis, Gene)
setkey(trans, Gene)
setkey(frq, SNP)

dir.create('~/oliverpainfel/Data/eQTLGen/full/cis_trans_combined')
genes <- unique(cis$Gene)
for(i in genes[1:50]){
  print(i)
  cis_i<-cis[.(i)]
  trans_i<-trans[.(i)]

  cis_i <- cis[cis$Gene == i,]
  trans_i <- trans[trans$Gene == i,]
  
  both_i<-rbind(cis_i, trans_i)
  
  names(both_i)[names(both_i) == 'AssessedAllele']<-'A1'
  names(both_i)[names(both_i) == 'OtherAllele']<-'A2'
  
  # Insert FREQ from EUR reference
  # There don't seem to be any strand flips
  frq_tmp<-frq[.(both_i$SNP)]
  both_i_match<-merge(both_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by=c('SNP','A1','A2'))
  both_i_flip<-merge(both_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by.x=c('SNP','A1','A2'), by.y=c('SNP','A2','A1'))
  both_i_flip$FREQ<-1-both_i_flip$FREQ
  both_i<-rbind(both_i_match, both_i_flip)
  
  # Approximate BETA, SE, and P
  both_i$BETA<-both_i$Z/sqrt((2*both_i$FREQ)*(1-both_i$FREQ)*(both_i$NrSamples+(both_i$Z^2)))
  both_i$SE<-abs(both_i$BETA)/abs(both_i$Z)
  
  both_i <- both_i[, c('SNPChr','SNPPos','SNP','A1','A2','BETA','SE','Pvalue','NrSamples')]
  names(both_i) <- c('CHR','BP','SNP','A1','A2','BETA','SE','P','N')
  
  fwrite(both_i, paste0('~/oliverpainfel/Data/eQTLGen/full/cis_trans_combined/',i,'.gz'), quote = F, sep = ' ', na= 'NA')
}

############################

library(data.table)

cis<-fread('~/oliverpainfel/Data/eQTLGen/full/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt.gz')
cis$GenePos <- NULL
cis$NrCohorts <- NULL
cis$FDR <- NULL
cis$BonferroniP <- NULL
cis$GeneChr <- NULL

cis<-cis[cis$SNPChr == 22,]

# Read in EUR MAF
frq<-NULL
for(i in 1:22){
  frq<-rbind(frq, fread(paste0('/users/k1806347/oliverpainfel/Data/1KG/FUSION//1000G.EUR.',i,'.frq')))
}
names(frq)[names(frq) == 'MAF']<-'FREQ'

setkey(cis, Gene)
setkey(frq, SNP)

dir.create('~/oliverpainfel/Data/eQTLGen/full/cis_only')
genes <- unique(cis$Gene)
for(i in genes[1:50]){
  print(i)
  cis_i<-cis[.(i)]
  cis_i <- cis[cis$Gene == i,]

  names(cis_i)[names(cis_i) == 'AssessedAllele']<-'A1'
  names(cis_i)[names(cis_i) == 'OtherAllele']<-'A2'
  
  # Insert FREQ from EUR reference
  # There don't seem to be any strand flips
  frq_tmp<-frq[.(cis_i$SNP)]
  cis_i_match<-merge(cis_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by=c('SNP','A1','A2'))
  cis_i_flip<-merge(cis_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by.x=c('SNP','A1','A2'), by.y=c('SNP','A2','A1'))
  cis_i_flip$FREQ<-1-cis_i_flip$FREQ
  cis_i<-rbind(cis_i_match, cis_i_flip)
  
  # Approximate BETA, SE, and P
  cis_i$BETA<-cis_i$Z/sqrt((2*cis_i$FREQ)*(1-cis_i$FREQ)*(cis_i$NrSamples+(cis_i$Z^2)))
  cis_i$SE<-abs(cis_i$BETA)/abs(cis_i$Z)
  
  cis_i <- cis_i[, c('SNPChr','SNPPos','SNP','A1','A2','BETA','SE','Pvalue','NrSamples')]
  names(cis_i) <- c('CHR','BP','SNP','A1','A2','BETA','SE','P','N')
  
  fwrite(cis_i, paste0('~/oliverpainfel/Data/eQTLGen/full/cis_only/',i,'.gz'), quote = F, sep = ' ', na= 'NA')
}

write.table(genes[1:50], '~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', row.names=F, col.names=F, quote=F)
```

***

## Run GenoPred

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

# Create config file
conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml',
  'gwas_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_methods: ['ptclump','sbayesrc','sbayesr','quickprs','lassosum','ldpred2','prscs','dbslmm','megaprs']",
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']",
  "lassosum_pseudo_only: true",
  "megaprs_pseudo_only: true",
  "prscs_phi: ['auto']",
  "dbslmm_h2f: ['1']",
  "cores_prep_pgs: 1",
  "quickprs_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/quickprs/hm3",
  "sbayesrc_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/sbayesrc/hm3"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml', col.names = F, row.names = F, quote = F)

# Create target_list
target_list <- data.table(name = 'gtex_v8',
                          path = '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/processed/gtex_v8.snp.maf001.hm3',
                          type = 'plink2',
                          indiv_report = F)

write.table(
  target_list,
  '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  col.names = T,
  row.names = F,
  quote = F,
  sep = ' '
)

# Create gwas_list
genes <- gsub('.gz', '', list.files('~/oliverpainfel/Data/eQTLGen/full/cis_trans_combined'))

gwas_list <- data.table(
  name = genes,
  path = paste0('/users/k1806347/oliverpainfel/Data/eQTLGen/full/cis_trans_combined/', genes,'.gz'),
  population = 'EUR',
  n = NA,
  sampling = NA,
  prevalence = NA,
  mean = NA,
  sd = NA,
  label = genes)

write.table(gwas_list, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list.txt', col.names = T, row.names = F, quote = F, sep = ' ')

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml \
  target_pgs -n 
```

This is probably an inefficient way of running these PGS methods. If the sumstats were truly genome-wide then it would be perfect, but eQTLGen trans QTLs are restricted to 10K variants. Several methods do run in a fairly efficient manor (ldpred2, lassosum, quickprs, ptclump). quickprs and sbayesrc will have issues due to poor overlap with the reference. dbslmm and prs-cs are run per chromosome so won't work.

Two options, find truly genome-wide trans-eQTL dataset, or continue with cis-eQTLs only.

For cis-eQTL only analysis, we can make GenoPred more efficient by ensuring only the relevant LD reference data is used. We just need to subset the reference to include eQTL sumstats for given gene, and adjsut CHROMS to relevant chromosome. SBayesRC we can subset the reference to the relevant blocks. The only extra overhead of running in GenoPred would be that there is seperate job for every method, but shouldn't be much of an overhead. It will also track jobs completing better. WuickPRS can't be so easily remedied as the reference data assumes a certain overlap with the sumstats. MegaPRS would be fine though. I can make an R function for packaging models in FUSION format for TWAS.

Note. I am running into an issue regarding reference data for sbayesr and sbayesrc 

DBSLMM and LDpred2 require SNP-h2 estimate. Currently this comes from LDSC, which is inaccurate when using subsets of variants. It would be better if this was done using sbayesr or sbayesrc. Problem is SBayesRC is taking a while to run when on subset of variants. We could test when using cis-eQTLs only, or using SBayesR which I know if fairly accurate and quick. We would have to do this prior to running DBSLMM and LDpred2, and provide an option to eithe estimate SNP-h2 using LDSC or SBayesR.

***

# Analysis Phase 2: GenoPred with cis only

## Run GenoPred

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

# Create config file
conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml',
  'gwas_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list_cis_only.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_methods: ['ptclump','quickprs','lassosum','ldpred2','prscs','dbslmm','megaprs','sbayesr','sdpr']",
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']",
  "lassosum_pseudo_only: true",
  "ldpred2_model: ['auto']",
  "megaprs_pseudo_only: true",
  "prscs_phi: ['auto']",
  "dbslmm_h2f: ['1']",
  "h2_method: 'sbayesr'",
  "cores_prep_pgs: 1",
  "quickprs_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/quickprs/hm3",
  "sbayesrc_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/sbayesrc/hm3"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml', col.names = F, row.names = F, quote = F)

# Create gwas_list
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

gwas_list <- data.table(
  name = genes,
  path = paste0('/users/k1806347/oliverpainfel/Data/eQTLGen/full/cis_only/', genes,'.gz'),
  population = 'EUR',
  n = NA,
  sampling = NA,
  prevalence = NA,
  mean = NA,
  sd = NA,
  label = genes)

write.table(gwas_list, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list_cis_only.txt', col.names = T, row.names = F, quote = F, sep = ' ')

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml \
  target_pgs -n 
```

Testing using ptclump. Check it runs with other methods. 

sbayesrc very slow due to using all 591 LD blocks despite sumstats only including one locus.

target_scoring: Maybe we can speed this up by only include chr for each PGS that have >1 non-zero effect? Haven't tested yet so maybe not worth it. Or we can just export the weights into FUSION format and parse to FeaturePred...

***

## Evaluate PGS

```{r}
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
pgs <- read_pgs(config = config, pop = 'EUR', pseudo_only = T)$gtex_v8$EUR

# Get a list of genes
score_files <- list_score_files(config = config)
genes <- unique(score_files$name)

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in names(pgs)){
  obs_resid_i <- obs_resid[, c('ID', i), with = F]
  names(obs_resid_i)<-c('ID','obs')
  pgs_i <- pgs[[i]]
  for(j in names(pgs_i)){
    pgs_i_j <- pgs[[i]][[j]]
    pgs_i_j$FID<-NULL
    if(ncol(pgs_i_j) == 1){
      next
    }
    names(pgs_i_j)<-c('ID','pred')
    
    both_i_j <- merge(obs_resid_i, pgs_i_j, by = 'ID')
    both_i_j <- both_i_j[complete.cases(both_i_j),]
    
    if(nrow(both_i_j) > 0){
      mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
      
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = coef(mod)[2,1],
          se = coef(mod)[2,2],
          p = coef(mod)[2,4],
          n = nrow(both_i_j)
        ))
    } else {
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = NA,
          se = NA,
          p = NA,
          n = 0
        ))      
    }
  }
}

library(ggplot2)
library(cowplot)

# Restrict to genes available for all methods
cor_res <- cor_res[complete.cases(cor_res),]
cor_res_complete <- NULL
for(i in names(pgs)){
    pgs_i <- pgs[[i]]
    if(all(pgs_methods %in% names(pgs_i))){
      cor_res_complete <- rbind(cor_res_complete, cor_res[cor_res$gene == i,])
    }
}
cor_res <- cor_res_complete

ggplot(cor_res, aes(x=gene, y=r , fill = pgs_method)) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

# Compute medians using aggregate()
medians <- aggregate(r ~ pgs_method, data = cor_res, FUN = median, na.rm = TRUE)

ggplot(cor_res, aes(x = pgs_method, y = r, fill = pgs_method)) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = pgs_method, y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

```

This looks good. It shows different methods perform best for each gene. SBayesR and DBSLMM perform well.

***

# Analysis Phase 3: eqtl_to_twas with cis only

Lets run the same analysis, but derive the models using the eqtl_to_twas code to see how it compares in terms of predictive utility and computational efficiency.

We have generated models using the script previously, so use those models. Use FeaturePred to predict expression into the GTEx sample.

```{bash}

conda activate feature_pred
sbatch -p neurohack_cpu,cpu --mem 30G -n 40 -t 200 --wrap="
  Rscript /users/k1806347/oliverpainfel/Software/MyGit/Predicting-TWAS-features/FeaturePred.V2.0.R \
  	--PLINK_prefix_chr /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/processed/gtex_v8.snp.maf001.hm3.chr \
  	--ref_ld_chr /users/k1806347/oliverpainfel/Data/1KG/FUSION/1000G.EUR. \
  	--ref_maf /users/k1806347/oliverpainfel/Data/1KG/FUSION/1000G.EUR. \
  	--weights /users/k1806347/oliverpainfel/Data/eQTLGen/eQTL_to_TWAS/eQTLGen.eQTL.pos \
  	--weights_dir /users/k1806347/oliverpainfel/Data/eQTLGen/eQTL_to_TWAS/ \
  	--plink /users/k1806347/oliverpainfel/Software/plink1.9.sh \
  	--n_cores 40 \
  	--memory 10000 \
  	--all_mod T \
  	--pigz /users/k1806347/oliverpainfel/Software/pigz \
  	--output /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/eqtl_to_twas/output/gtex_v8.eqtlgen
"

```

***

## Evaluate PGS

```{r}
library(data.table)

# Get the list of genes tested above
genes <- gsub('.gz', '', list.files('~/oliverpainfel/Data/eQTLGen/full/cis_only'))

# Read in predicted expression by eqtl_to_twas
pred_exp <- fread('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/eqtl_to_twas/output/gtex_v8.eqtlgen/FeaturePredictions_eQTLGen.eQTL.txt.gz')
pred_exp<-pred_exp[, c(1, which(grepl(paste(genes, collapse = '|'), names(pred_exp)))), with=F]
names(pred_exp)[1] <- 'ID'

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in genes){
  pred_exp_i <- pred_exp[, c(1, which(grepl(paste0('\\.', i, '\\.'), names(pred_exp)))), with = F]
  obs_resid_i <- obs_resid[, c('ID', i), with = F]
  names(obs_resid_i)<-c('ID','obs')
  methods <- gsub('.*\\.', '', names(pred_exp_i)[-1])
  for(j in methods){
    pred_exp_i_j <- pred_exp_i[, c(1, which(grepl(paste0('\\.', j, '$'), names(pred_exp_i)))), with = F]
    if(ncol(pgs_i_j) == 1){
      next
    }
    names(pred_exp_i_j)<-c('ID','pred')
    
    both_i_j <- merge(obs_resid_i, pred_exp_i_j, by = 'ID')
    both_i_j <- both_i_j[complete.cases(both_i_j),]
    
    if(nrow(both_i_j) > 0){
      mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
      
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = coef(mod)[2,1],
          se = coef(mod)[2,2],
          p = coef(mod)[2,4],
          n = nrow(both_i_j)
        ))
    } else {
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = NA,
          se = NA,
          p = NA,
          n = 0
        ))      
    }
  }
}

library(ggplot2)
library(cowplot)

ggplot(cor_res, aes(x=gene, y=r , fill = pgs_method)) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

# Restrict to genes available for all methods
cor_res <- cor_res[complete.cases(cor_res),]
cor_res_complete <- NULL
for(i in names(pgs)){
    pgs_i <- pgs[[i]]
    if(all(pgs_methods %in% names(pgs_i))){
      cor_res_complete <- rbind(cor_res_complete, cor_res[cor_res$gene == i,])
    }
}

cor_res <- cor_res_complete
# Compute medians using aggregate()
medians <- aggregate(r ~ pgs_method, data = cor_res, FUN = median, na.rm = TRUE)

ggplot(cor_res, aes(x = pgs_method, y = r, fill = pgs_method)) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = pgs_method, y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

```

The predicted-observed correlations are highly concordant between GenoPred and eqtl_to_twas methods. Benefits of GenoPred are incusion of megaprs and quickprs, integration with wider GenoPred pipeline. Benefits of eqtl_to_twas are that it includes the top1 and susie is probably faster and requires fewer job submissions. I kind of want to keep developing GenoPred as seems more fun and better integrated with GenoPred ecosystem. We can include top1 and susie models if we feel necessary, and can think of ways to speed up target scoring.

***

# Analysis Phase 4: EXPRESSO

## Published models

```{bash}
# Download EXPRESSO repo
cd /users/k1806347/oliverpainfel/Software
git clone https://github.com/LidaWangPSU/EXPRESSO.git

# Examine published wieghts from eQTLGen
cd /users/k1806347/oliverpainfel/Software/EXPRESSO/trained_model
unzip EXPRESSO_Whole_Blood_eQTLGen_weights_hg19.txt.zip
```

```{r}
library(data.table)
weights<-fread('/users/k1806347/oliverpainfel/Software/EXPRESSO/trained_model/EXPRESSO_Whole_Blood_eQTLGen_weights_hg19.txt')

length(unique(weights$gene_id))
# 12280

nrow(weights)
# 540457

weights$heritable <- NULL

# Split the SNP column into components
snp_parts <- do.call(rbind, strsplit(weights$snp, "_"))
weights$chr_name <- snp_parts[,1]
weights$chr_position <- snp_parts[,2]
weights$effect_allele <- snp_parts[,3]
weights$other_allele <- snp_parts[,4]

# Rename 'weight' column to match PGS format
weights$effect_weight <- weights$weight

dir.create('~/oliverpainfel/Data/EXPRESSO/published_models', recursive = T)
for(i in unique(weights$gene_id)){
  # Select and reorder columns
  pgs_score_i<-weights[weights$gene_id == i,]
  pgs_score <- pgs_score_i[, c("chr_name", "chr_position", "effect_allele", "other_allele", "effect_weight")]

  # Save to a .score file
  write.table(
    pgs_score,
    file = paste0(
      "~/oliverpainfel/Data/EXPRESSO/published_models/eQTLGen_",
      i,
      ".score"
    ),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE)
}
```

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

# Create config file
conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml',
  'gwas_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/gwas_list.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  'score_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list.txt',
  "pgs_methods: ['ptclump','sbayesrc','sbayesr','quickprs','lassosum','ldpred2','prscs','dbslmm','megaprs']",
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']",
  "lassosum_pseudo_only: true",
  "megaprs_pseudo_only: true",
  "prscs_phi: ['auto']",
  "dbslmm_h2f: ['1']",
  "cores_prep_pgs: 1",
  "quickprs_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/quickprs/hm3",
  "sbayesrc_ldref: /users/k1806347/oliverpainfel/Data/hgdp_1kg/sbayesrc/hm3"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config.yaml', col.names = F, row.names = F, quote = F)

conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso_pub.yaml',
  'score_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso_pub.yaml', col.names = F, row.names = F, quote = F)

# Create gwas_list
genes <- gsub('.gz', '', list.files('~/oliverpainfel/Data/eQTLGen/full/cis_trans_combined'))

score_list <- data.table(
  name = paste0(genes,'_expresso_pub'),
  path = paste0(
      "/users/k1806347/oliverpainfel/Data/EXPRESSO/published_models/eQTLGen_",
      genes,
      ".score"),
  label = paste0(genes,'_expresso_pub'))

write.table(score_list, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list.txt', col.names = T, row.names = F, quote = F, sep = ' ')

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso_pub.yaml \
  target_pgs -n 
```

This doesn't work as they did not restrict to HapMap3 variants, so poor overlap with GenoPred reference. Run EXPRESSO to generate weights using data harmonised with the GenoPred reference to improve comparability to other approaches. We should probably consider expanding the SNP-list at some point to see if it makes an improvement.

***

## Generate models

```{r}
# Prepare input
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

# eQTL sumstats
library(data.table)

cis<-fread('~/oliverpainfel/Data/eQTLGen/full/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt.gz')
cis$GenePos <- NULL
cis$NrCohorts <- NULL
cis$FDR <- NULL
cis$BonferroniP <- NULL
cis$GeneChr <- NULL

cis<-cis[cis$SNPChr == 22,]

# Read in EUR MAF
frq<-NULL
for(i in 1:22){
  frq<-rbind(frq, fread(paste0('/users/k1806347/oliverpainfel/Data/1KG/FUSION//1000G.EUR.',i,'.frq')))
}
names(frq)[names(frq) == 'MAF']<-'FREQ'

setkey(cis, Gene)
setkey(frq, SNP)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/sumstats/', recursive = T)

genes <- unique(cis$Gene)
for(i in genes[1:50]){
  print(i)
  cis_i<-cis[.(i)]
  cis_i <- cis[cis$Gene == i,]

  names(cis_i)[names(cis_i) == 'AssessedAllele']<-'A1'
  names(cis_i)[names(cis_i) == 'OtherAllele']<-'A2'
  
  # Insert FREQ from EUR reference
  # There don't seem to be any strand flips
  frq_tmp<-frq[.(cis_i$SNP)]
  cis_i_match<-merge(cis_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by=c('SNP','A1','A2'))
  cis_i_flip<-merge(cis_i, frq_tmp[,c('SNP','A1','A2','FREQ'),with=F], by.x=c('SNP','A1','A2'), by.y=c('SNP','A2','A1'))
  cis_i_flip$FREQ<-1-cis_i_flip$FREQ
  cis_i<-rbind(cis_i_match, cis_i_flip)
  
  # Approximate BETA, SE, and P
  cis_i$BETA<-cis_i$Z/sqrt((2*cis_i$FREQ)*(1-cis_i$FREQ)*(cis_i$NrSamples+(cis_i$Z^2)))
  cis_i$SE<-abs(cis_i$BETA)/abs(cis_i$Z)
  
  cis_i <- cis_i[, c('SNPChr','SNPPos','SNP','A1','A2','BETA','SE','Pvalue','NrSamples','GeneSymbol','Gene','FREQ')]
  names(cis_i) <- c('CHR','BP','SNP','A1','A2','BETA','SE','P','N','SYMBOL','GENE','FREQ')

  cis_i<-cis_i[,c('GENE','SYMBOL','CHR','BP','A1','A2','FREQ','BETA','N','SNP'), with=F]
  names(cis_i)<-c('gene','symbol','chr','pos','ref','alt','af','beta.est','n','snp_id')
  
  fwrite(cis_i, paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/sumstats/', i,'.txt'), sep='\t', na='NA', quote=F)
}

# Annotation file
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

pvar <- read_pvar('resources/data/ref/ref.chr')
pvar <- pvar[, c('SNP','CHR','BP','A1','A2'), with=F]
names(pvar)<-c('snp','chr','pos','ref','alt')
pvar$overlap<-0

fwrite(pvar, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/annot.txt', sep='\t', na='NA', quote=F)

```

Make an R script to run EXPRESSO for each gene in parallel.

```{r}
#!/usr/bin/env Rscript
args <- commandArgs(trailingOnly = TRUE)
gene <- args[1]

# load libraries & objects
library(EXPRESSO)
library(fast.lasso)
library(rareGWAMA)
library(MASS)
library(data.table)

# assume annoFile, windowFile, refFile are set here or via a config
refFile <- "/users/k1806347/oliverpainfel/Data/1KG/FUSION/1000G.EUR.22"
annoFile <- "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/annot.txt"
sumstatFile<- "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/eqtl_sumstats.txt"

windowFile <- c( "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_1mb_chr22.txt",
                 "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_250kb_chr22.txt",
                 "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_loop_chr22.txt",
                 "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_domain_chr22.txt",
                 "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_tad_chr22.txt",
                 "/users/k1806347/oliverpainfel/Software/EXPRESSO/example_data/Whole_Blood_pchic_chr22.txt")

indir <- paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/sumstats/', gene,'.txt')
outdir  <- paste0("/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/output/", gene)

EXPRESSO(
  indir,
  annoFile,
  windowFile,
  refFile,
  outdir,
  minMaf  = 0.05,
  maxIter = 1000,
  append  = FALSE
)
```

```{bash}
#!/bin/bash
#SBATCH --job-name=EXPRESSO
#SBATCH --output=~/logs/expresso_%A_%a.out
#SBATCH --error=~/logs/expresso_%A_%a.err
#SBATCH --time=12:00:00
#SBATCH --mem=10G
#SBATCH --ntasks=1
#SBATCH --partition=neurohack_cpu
#SBATCH --array=1-50%50    # adjust 1–N to your #genes, %50 = max concurrent

# build the gene list into a Bash array
mapfile -t GENES < ~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt

# SLURM_ARRAY_TASK_ID runs from 1..N
gene=${GENES[$SLURM_ARRAY_TASK_ID-1]}

# call the R wrapper
./run_expresso.R "$gene"
```

```{bash}
cd /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO
chmod a+x /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/run_expresso.R
chmod a+x /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/run_expresso.sh
sbatch run_expresso.sh
```

Format into score files for GenoPred.

```{r}
library(data.table)

genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/score_files', recursive = T)
for(model in c('mse','pseudo')){
  for(i in genes){
    weights <-
      fread(
        paste0(
          '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/output/',
          i,
          '-weight.', model, '.txt'
        ), fill = T)
    
    weights <- weights[, c('snp','weight'), with = F]
    
    # Insert A1 and A2 from eqtl sumstats (not the ref)
    eqtl_ss<-fread(paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/input/sumstats/', i,'.txt'))
    
    weights <- merge(weights, eqtl_ss[,c('snp_id','ref','alt'), with = F], by.x = 'snp', by.y = 'snp_id', all.x = T)
    
    score_file<-data.table(
      rsID = weights$snp,
      effect_allele = weights$ref,
      other_allele = weights$alt,
      effect_weight = weights$weight
    )

    # Save to a .score file
    write.table(
      score_file,
      file = paste0(
        "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/score_files/",
        i,
        '.',
        model,
        ".score"
      ),
      sep = "\t",
      quote = FALSE,
      row.names = FALSE)
  }
}
```

***

## Predict expression

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso.yaml',
  'score_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list_expresso.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso.yaml', col.names = F, row.names = F, quote = F)

# Create gwas_list
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1
methods <- c("mse","pseudo")

score_list <- data.table(
  name = paste0(rep(genes, each = length(methods)), "_expresso_", rep(methods, times = length(genes))),
  path = paste0(
    "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/EXPRESSO/score_files/",
    rep(genes, each = length(methods)),
    ".",
    rep(methods, times = length(genes)),
    ".score"
  ),
  label = paste0(rep(genes, each = length(methods)), "_expresso_", rep(methods, times = length(genes)))
)

write.table(
  score_list,
  '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list_expresso.txt',
  col.names = T,
  row.names = F,
  quote = F,
  sep = ' '
)

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso.yaml \
  target_pgs -n 
```

***

## Evaluate PGS

```{r}
# Compare predictions from GenoPred and EXPRESSO directly
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

###
# GenoPred
###

setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
pgs_genopred <- read_pgs(config = config, pop = 'EUR', pseudo_only = T)$gtex_v8$EUR

# Get a list of genes
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

###
# EXPRESSO
###

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_expresso.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
score_files <- list_score_files(config = config)
pgs_expresso <- read_pgs(config = config, pop = 'EUR')$gtex_v8$EUR

pgs_expresso_new<-list()
for(i in genes){
  pgs_expresso_new[[i]] <-
    lapply(pgs_expresso[which(grepl(i, names(pgs_expresso)))], function(x) {
      x[[1]]
    })
  names(pgs_expresso_new[[i]])<-gsub('.*_','',names(pgs_expresso_new[[i]]))
}
pgs_expresso <- pgs_expresso_new
rm(pgs_expresso_new)

pgs_all <- list()
pgs_all[['genopred']] <- pgs_genopred
pgs_all[['expresso']] <- pgs_expresso

###
# Observed
###

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in both_genes){
  for(j in c('genopred', 'expresso')){
    pgs_i_j <- pgs_all[[j]][[i]]
    for(k in names(pgs_i_j)){
      obs_resid_i <- obs_resid[, c('ID', i), with = F]
      names(obs_resid_i)<-c('ID','obs')
    
      pgs_i_j_k <- pgs_i_j[[k]]
      pgs_i_j_k$FID<-NULL
      if(ncol(pgs_i_j_k) == 1){
        next
      }
      names(pgs_i_j_k)<-c('ID','pred')
      
      both_i_j <- merge(obs_resid_i, pgs_i_j_k, by = 'ID')
      both_i_j <- both_i_j[complete.cases(both_i_j),]
      
      if(nrow(both_i_j) > 0){
        mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
        
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = coef(mod)[2,1],
            se = coef(mod)[2,2],
            p = coef(mod)[2,4],
            n = nrow(both_i_j),
            pgs_method = k
          ))
      } else {
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = NA,
            se = NA,
            p = NA,
            n = 0,
            pgs_method = k
          ))      
      }
    }
  }
}

library(ggplot2)
library(cowplot)

ggplot(cor_res, aes(x=gene, y=r , fill = interaction(software, pgs_method))) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

ggplot(cor_res[cor_res$pgs_method %in% c('pseudo','mse','prscs'),], aes(x=gene, y=r , fill = interaction(software, pgs_method))) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

# GO TO HERE: 8735 is particularly bad compared to other methods. See what is going on. Could indicate alleles are being mixed up. ALSO, the target scoring step did not detect the external score files had been updated.

# Compute medians using aggregate()
cor_res_comp <- cor_res[complete.cases(cor_res),]
gene_n<-data.frame(table(cor_res_comp$gene))
cor_res_comp <- cor_res_comp[cor_res_comp$gene %in% gene_n$Var1[gene_n$Freq == 11],]

medians <- aggregate(r ~ pgs_method + software, data = cor_res_comp, FUN = median, na.rm = TRUE)

counts <- aggregate(r ~ pgs_method + software, data = cor_res_comp, FUN = function(x) sum(!is.na(x)))
names(counts)[3] <- "n"
counts$group <- interaction(counts$software, counts$pgs_method)

ggplot(cor_res_comp, aes(x = interaction(software, pgs_method), y = r, fill = interaction(software, pgs_method))) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = interaction(software, pgs_method), y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  geom_text(
    data = counts,
    aes(x = group, y = 0.55, label = paste0("n=", n)),  # adjust y as needed
    inherit.aes = FALSE,
    size = 3
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

# Check the correlation between the predicted expression levels
pred_cor<-list()
for(i in genes){
  pgs_all_i_genopred <- pgs_all[['genopred']][[i]]
  pgs_all_i_genopred_tmp <- do.call(cbind, lapply(pgs_all_i_genopred, function(x) {
    if(ncol(x) == 3){
      tmp <- x[,3]
    } else {
      tmp <- NA
    }
    return(tmp)
  }))
  names(pgs_all_i_genopred_tmp)<-paste0('genopred_', names(pgs_all_i_genopred))
  
  pgs_all_i_expresso <- pgs_all[['expresso']][[i]]
  pgs_all_i_expresso_tmp <- do.call(cbind, lapply(pgs_all_i_expresso, function(x) {
    tmp <- x[,3]
  }))
  names(pgs_all_i_expresso_tmp)<-paste0('expresso_', names(pgs_all_i_expresso))

  pred_cor[[i]] <- cor(cbind(pgs_all_i_genopred_tmp, pgs_all_i_expresso_tmp))
}

# Stack all matrices into an array
bad_entries <- sapply(pred_cor, function(x) !is.matrix(x) || any(dim(x) != dim(pred_cor[[1]])))
names(pred_cor)[bad_entries]
pred_cor_clean <- pred_cor[!bad_entries]
cor_array <- simplify2array(pred_cor_clean)

# Take the mean across the third dimension (genes)
avg_cor <- apply(cor_array, c(1, 2), mean, na.rm = TRUE)
      
library(reshape2)
library(ggplot2)

melted_cor <- melt(avg_cor)

ggplot(melted_cor, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", limits = c(0.5, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "", fill = "Avg Cor")
```

***

# Analysis Phase 5: OTTERS

We will need to run OTTERS ourselves to generate models restricted to HapMap3. Some OTTERS code already in otters_comp.

***

## Install

```{bash}
# Install OTTERS v1.0.0
cd /users/k1806347/oliverpainfel/Software
git clone https://github.com/daiqile96/OTTERS.git 
cd OTTERS
git checkout v1.0.0

# Install tabix and bgzip
cd /users/k1806347/oliverpainfel/Software
wget https://sourceforge.net/projects/samtools/files/tabix/tabix-0.2.6.tar.bz2
tar jxvf tabix-0.2.6.tar.bz2
cd tabix-0.2.6
make

# copy the binary bgzip and tabix bin folder so can be called from PATH
mkdir /users/k1806347/oliverpainfel/Software/bin
cp bgzip tabix /users/k1806347/oliverpainfel/Software/bin

# copy the plink1.9 binary and add to PATH
cp /users/k1806347/oliverpainfel/Software/plink1.9/plink /users/k1806347/oliverpainfel/Software/bin/
export PATH=/users/k1806347/oliverpainfel/Software/bin:$PATH 

# Create a conda environment to run otters
cd /users/k1806347/oliverpainfel/Software/OTTERS
mamba env create --file otters_env.yaml
conda activate otters

# Install lassosum
Rscript -e 'remotes::install_github("tshmak/lassosum@v0.4.5")'

# Install SDPR
cd /users/k1806347/oliverpainfel/Software/bin
git clone https://github.com/eldronzhou/SDPR.git
cd SDPR
git checkout v0.9.1
```

***

## Prepare input

```{r}
# Get the list of genes tested above
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

##
# Annotation file (CHROM GeneStart GeneEnd TargetID)
##
pos<-fread('/users/k1806347/oliverpainfel/Data/eQTLGen/eQTL_to_TWAS/eQTLGen.eQTL.pos')

pos <- pos[pos$ID %in% genes,]
annot <- data.table(
  CHROM = pos$CHR,
  GeneStart = pos$P0,
  GeneEnd = pos$P1,
  TargetID = pos$ID
)

annot<-annot[order(annot$CHROM, annot$GeneStart),]

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/input', recursive = T)
fwrite(annot, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/input/annot.txt', sep='\t', na='NA', quote=F)

##
# eQTL sumstats (CHROM POS A1 A2 Z TargetID N)
##

library(data.table)

cis<-fread('~/oliverpainfel/Data/eQTLGen/full/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt.gz')
cis$GenePos <- NULL
cis$NrCohorts <- NULL
cis$FDR <- NULL
cis$BonferroniP <- NULL
cis$GeneChr <- NULL
cis<-cis[cis$Gene %in% genes,]

eqtl_ss<-data.table(
  CHROM = cis$SNPChr,
  POS = cis$SNPPos,
  A1 = cis$AssessedAllele,
  A2 = cis$OtherAllele,
  Z = cis$Zscore,
  TargetID = cis$Gene,
  N = cis$NrSamples
)

eqtl_ss<-eqtl_ss[order(eqtl_ss$CHROM, eqtl_ss$POS),]

fwrite(eqtl_ss, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/input/eqtl_ss.txt', sep='\t', na='NA', quote=F)

```

***

## Run OTTERS Stage 1

```{bash}
conda activate otters

# set number of threads to be used
N_THREADS=3

# set up my OTTERS directory and SDPR directory
OTTERS_DIR=/users/k1806347/oliverpainfel/Software/OTTERS
SDPR_DIR=/users/k1806347/oliverpainfel/Software/bin/SDPR

# Load plink
module add plink

# make sure the dynamic libraries of SDPR are not changed (For SDPR)
export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:${SDPR_DIR}/MKL/lib
export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:${SDPR_DIR}/gsl/lib

# prevent automatically using  all available cores on a compute node (For SDPR and PRS-CS)
export MKL_NUM_THREADS=$N_THREADS
export NUMEXPR_NUM_THREADS=$N_THREADS
export OMP_NUM_THREADS=$N_THREADS

# Input for OTTERS STAGE I 
# Annotation File 
exp_anno=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/input/annot.txt
# eQTL summary statistics 
sst_file=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/input/eqtl_ss.txt

# Set LD-clumping threshold in STAGE I
clump_r2=0.99

# STAGE I
# train eQTL weights using P+T, lassosum, SDPR and PRS-CS. 
# It may take several minutes to complete.
for chr in $(seq 22 22); do
  # Genotype data from LD reference panel
  geno_dir=/users/k1806347/oliverpainfel/Data/1KG/FUSION/1000G.EUR.${chr}

  # Set output directory for STAGE I
  mkdir -p /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/output_chr${chr}
  out_dir=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/output_chr${chr}

  sbatch -p neurohack_cpu,cpu --mem 20G -n 3 -t 200 --wrap="
    python3 ${OTTERS_DIR}/training.py \
      --OTTERS_dir=${OTTERS_DIR} \
      --SDPR_dir=${SDPR_DIR} \
      --anno_dir=${exp_anno} \
      --geno_dir=${geno_dir} \
      --sst_file=${sst_file} \
      --out_dir=${out_dir} \
      --chrom=${chr} \
      --r2=${clump_r2} \
      --models=PT,lassosum,SDPR,PRScs \
      --thread=$N_THREADS \
      --lassosum_ld_blocks EUR.hg19"
done
```

```{r}
# Combine output
models <- NULL
for(i in 1:22){
  if(file.exists(paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/output_chr', i,'/lassosum.txt'))){
    for(j in list.files(paste0('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/output_chr', i))){
      models_i_j <-
        fread(
          paste0(
            '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/output_chr',
            i,
            '/',
            j
          )
        )
      names(models_i_j)[names(models_i_j) == 'ES']<-'BETA'
      models_i_j$method <- gsub('.txt','',j)
      models <- rbind(models, models_i_j)
    }
  }
}

# Convert into score files
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/score_files', recursive = T)
for(i in unique(models$TargetID)){
  for(j in unique(models$method[models$TargetID == i])){
    models_i_j <- models[models$TargetID == i & models$method == j,]
    
    weights<-data.table(
      chr_name = models_i_j$CHROM,
      chr_position = models_i_j$POS,
      effect_allele = models_i_j$A1,
      other_allele = models_i_j$A2,
      effect_weight = models_i_j$BETA
    )

    weights <- weights[weights$effect_weight != 0 & !is.na(weights$effect_weight),]
    
    # Save to a .score file
    write.table(
      weights,
      file = paste0(
        "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/score_files/",
        i,
        '.',
        j,
        ".score"
      ),
      sep = "\t",
      quote = FALSE,
      row.names = FALSE)
  }
}
```

***

## Predict expression

```{r}
setwd('/users/k1806347/oliverpainfel/Software/MyGit/GenoPred/pipeline')
library(data.table)

dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config', recursive = T)
dir.create('/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output', recursive = T)

conf <- c(
  'outdir: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output',
  'config_file: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_otters.yaml',
  'score_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list_otters.txt',
  'target_list: /scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/target_list.txt',
  "pgs_scaling: ['discrete']",
  "target_populations: ['EUR']"
)

write.table(conf, '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_otters.yaml', col.names = F, row.names = F, quote = F)

# Create gwas_list
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1
methods <- c("lassosum", "P0.001", "P0.05", "PRScs", "SDPR")

score_list <- data.table(
  name = paste0(rep(genes, each = length(methods)), "_otters_", rep(methods, times = length(genes))),
  path = paste0(
    "/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/OTTERS/score_files/",
    rep(genes, each = length(methods)),
    ".",
    rep(methods, times = length(genes)),
    ".score"
  ),
  label = paste0(rep(genes, each = length(methods)), "_otters_", rep(methods, times = length(genes)))
)

write.table(
  score_list,
  '/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/score_list_otters.txt',
  col.names = T,
  row.names = F,
  quote = F,
  sep = ' '
)

```

```{bash}
snakemake \
  --profile slurm \
  --use-conda \
  --configfile=/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_otters.yaml \
  target_pgs -n 
```

***

## Evaluate PGS

```{r}
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_otters.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
pgs <- read_pgs(config = config, pop = 'EUR', pseudo_only = T)$gtex_v8$EUR

# Get a list of genes
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in genes){
  obs_resid_i <- obs_resid[, c('ID', i), with = F]
  names(obs_resid_i)<-c('ID','obs')
  for(j in c('lassosum', 'P0.001', 'P0.05', 'PRScs', 'SDPR')){
    pgs_i_j <- pgs[[paste0(i,'_otters_',j)]]$external
    pgs_i_j$FID<-NULL
    if(ncol(pgs_i_j) == 1){
      next
    }
    names(pgs_i_j)<-c('ID','pred')
    
    both_i_j <- merge(obs_resid_i, pgs_i_j, by = 'ID')
    both_i_j <- both_i_j[complete.cases(both_i_j),]
    
    if(nrow(both_i_j) > 0){
      mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
      
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = coef(mod)[2,1],
          se = coef(mod)[2,2],
          p = coef(mod)[2,4],
          n = nrow(both_i_j)
        ))
    } else {
      cor_res <- rbind(
        cor_res,
        data.frame(
          gene = i,
          pgs_method = j,
          r = NA,
          se = NA,
          p = NA,
          n = 0
        ))      
    }
  }
}

library(ggplot2)
library(cowplot)

# Restrict to genes available for all methods
cor_res <- cor_res[complete.cases(cor_res),]
cor_res_complete <- NULL
for(i in names(pgs)){
    pgs_i <- pgs[[i]]
    if(all(c('lassosum', 'P0.001', 'P0.05', 'PRScs', 'SDPR') %in% names(pgs_i))){
      cor_res_complete <- rbind(cor_res_complete, cor_res[cor_res$gene == i,])
    }
}
cor_res <- cor_res_complete

ggplot(cor_res, aes(x=gene, y=r , fill = pgs_method)) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

# Compute medians using aggregate()
medians <- aggregate(r ~ pgs_method, data = cor_res, FUN = median, na.rm = TRUE)

ggplot(cor_res, aes(x = pgs_method, y = r, fill = pgs_method)) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = pgs_method, y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

```

The results are generally consistent with GenoPred and eQTL_to_TWAS. I have noticed that there are a few genes where the lassosum model performs poorly in GenoPred but not eQTL_to_TWAS, and the average performance of lassosum is better on average in OTTERS than in GenoPred and eQTL_to_TWAS. Look into this.

OTTERS uses a different approach when providing lassosum with the correlations for each SNP - It using Z/sqrt(N) and then scales between -1 and 1. This goes against the advice from lassosum authors, but seems to work well.

```{r}
# Compare predictions from GenoPred and OTTERS directly
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

###
# GenoPred
###

setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
pgs_genopred <- read_pgs(config = config, pop = 'EUR', pseudo_only = T)$gtex_v8$EUR

# Get a list of genes
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1

###
# OTTERS
###

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_otters.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
score_files <- list_score_files(config = config)
pgs_otters <- read_pgs(config = config, pop = 'EUR')$gtex_v8$EUR

pgs_otters_new<-list()
for(i in genes){
  pgs_otters_new[[i]] <-
    lapply(pgs_otters[which(grepl(i, names(pgs_otters)))], function(x) {
      x[[1]]
    })
  names(pgs_otters_new[[i]])<-gsub('.*_','',names(pgs_otters_new[[i]]))
}
pgs_otters <- pgs_otters_new
rm(pgs_otters_new)

pgs_all <- list()
pgs_all[['genopred']] <- pgs_genopred
pgs_all[['otters']] <- pgs_otters

###
# Observed
###

# Read in observed expression data
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)

cor_res<-NULL
for(i in both_genes){
  for(j in c('genopred', 'otters')){
    pgs_i_j <- pgs_all[[j]][[i]]
    for(k in names(pgs_i_j)){
      obs_resid_i <- obs_resid[, c('ID', i), with = F]
      names(obs_resid_i)<-c('ID','obs')
    
      pgs_i_j_k <- pgs_i_j[[k]]
      pgs_i_j_k$FID<-NULL
      if(ncol(pgs_i_j_k) == 1){
        next
      }
      names(pgs_i_j_k)<-c('ID','pred')
      
      both_i_j <- merge(obs_resid_i, pgs_i_j_k, by = 'ID')
      both_i_j <- both_i_j[complete.cases(both_i_j),]
      
      if(nrow(both_i_j) > 0){
        mod <- summary(lm(scale(obs) ~ scale(pred), data = both_i_j))
        
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = coef(mod)[2,1],
            se = coef(mod)[2,2],
            p = coef(mod)[2,4],
            n = nrow(both_i_j),
            pgs_method = k
          ))
      } else {
        cor_res <- rbind(
          cor_res,
          data.frame(
            gene = i,
            software = j,
            r = NA,
            se = NA,
            p = NA,
            n = 0,
            pgs_method = k
          ))      
      }
    }
  }
}

library(ggplot2)
library(cowplot)

ggplot(cor_res, aes(x=gene, y=r , fill = interaction(software, pgs_method))) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")

ggplot(cor_res[cor_res$pgs_method %in% c('sdpr','SDPR'),], aes(x=gene, y=r , fill = interaction(software, pgs_method))) +
    geom_errorbar(aes(ymin = r - se, ymax = r + se),
                  width = 0,
                  position = position_dodge(width = 1)) +
    geom_point(stat="identity", position=position_dodge(1), size=3, shape=23) +
    geom_vline(xintercept = seq(1.5, length(unique(cor_res$gene))), linetype="dotted") +
    labs(y = "R (SE)", x='Gene', fill = 'Method') +
    theme_half_open() +
    background_grid(major = 'y', minor = 'y') + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          legend.position = "top",
          legend.key.spacing.x = unit(1, "cm"),
          legend.justification = "center")


# Compute medians using aggregate()
medians <- aggregate(r ~ pgs_method + software, data = cor_res, FUN = median, na.rm = TRUE)

counts <- aggregate(r ~ pgs_method + software, data = cor_res, FUN = function(x) sum(!is.na(x)))
names(counts)[3] <- "n"
counts$group <- interaction(counts$software, counts$pgs_method)

ggplot(cor_res, aes(x = interaction(software, pgs_method), y = r, fill = interaction(software, pgs_method))) +
  geom_violin(position = position_dodge(width = 0.9), alpha = 0.6, width = 0.9) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9), outlier.shape = NA) +
  geom_text(
    data = medians,
    aes(x = interaction(software, pgs_method), y = 0.65, label = round(r, 3)),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  geom_text(
    data = counts,
    aes(x = group, y = 0.55, label = paste0("n=", n)),  # adjust y as needed
    inherit.aes = FALSE,
    size = 3
  ) +
  labs(y = "R", x = "Method") +
  theme_half_open() +
  background_grid(major = 'y', minor = 'y') + 
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'none'
  )

# Check the correlation between the predicted expression levels
pred_cor<-list()
for(i in genes){
  pgs_all_i_genopred <- pgs_all[['genopred']][[i]]
  pgs_all_i_genopred_tmp <- do.call(cbind, lapply(pgs_all_i_genopred, function(x) {
    if(ncol(x) == 3){
      tmp <- x[,3]
    } else {
      tmp <- NA
    }
    return(tmp)
  }))
  names(pgs_all_i_genopred_tmp)<-paste0('genopred_', names(pgs_all_i_genopred))
  
  pgs_all_i_otters <- pgs_all[['otters']][[i]]
  pgs_all_i_otters_tmp <- do.call(cbind, lapply(pgs_all_i_otters, function(x) {
    tmp <- x[,3]
  }))
  names(pgs_all_i_otters_tmp)<-paste0('otters_', names(pgs_all_i_otters))

  pred_cor[[i]] <- cor(cbind(pgs_all_i_genopred_tmp, pgs_all_i_otters_tmp))
}

# Stack all matrices into an array
bad_entries <- sapply(pred_cor, function(x) !is.matrix(x) || any(dim(x) != dim(pred_cor[[1]])))
names(pred_cor)[bad_entries]
pred_cor_clean <- pred_cor[!bad_entries]
cor_array <- simplify2array(pred_cor_clean)

# Take the mean across the third dimension (genes)
avg_cor <- apply(cor_array, c(1, 2), mean, na.rm = TRUE)
      
library(reshape2)
library(ggplot2)

melted_cor <- melt(avg_cor)

ggplot(melted_cor, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", limits = c(0.5, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "", fill = "Avg Cor")
```

As expected the results of OTTERS and GenoPred are highly concordant.

OTTERS has two p-value thresh models. I could specify these thresholds for ptclump. It also includes the SDPR method. 

***

# Analysis Phase 6: SUMMIT

Download published models.

***

# Analysis Phase 7: SUMMIT-FA

Download published models.

***

To do:
- Fix SBayesRC reference issue - I think it is fixed but reference data download is slow - 
- Make SBayesRC more efficient when using locus specific sumstats
- Updating analysis to 50 genes in chr22, so we can then run EXPRESSO with the data they provide.
- Compare prediction when using models from SUMMIT, SUMMIT-FA, OTTERS and EXPRESSO
- Look into AdipoExpress meta-analysis sumstats
- Look into liver eQTL meta-analysis: https://mohlke.web.unc.edu/data/signal-based-liver-eqtl-meta-analysis-broadaway-et-al-ajhg-2024/
- Could compare to models from OMICSPRED - They have shared sumstats
- Brain single cell meta-analysis: https://pmc.ncbi.nlm.nih.gov/articles/PMC11908325/#S22
- SABR - study. New eQTL dataset for AFR ancestry


Thoughts overall:
- GenoPred provides very little over OTTERS as it stands. The main advantage over OTTERS is its scalability to genome-wide QTL data. Another advance could come from modelling multiple PGS methods together.

***

```{bash}
mkdir /users/k1806347/oliverpainfel/Data/AdipoExpress
cd /users/k1806347/oliverpainfel/Data/AdipoExpress
wget https://zenodo.org/records/13845120/files/marginal_byChr_EURonly.tar.gz?download=1
mv 'marginal_byChr_EURonly.tar.gz?download=1' marginal_byChr_EURonly.tar.gz
tar -xvzf marginal_byChr_EURonly.tar.gz
```

```{r}
library(data.table)
ss <- fread('/users/k1806347/oliverpainfel/Data/AdipoExpress/marginal_byChr_EURonly/EURonly_marginal_local_eQTL_meta_chr10.txt')

table(ss$ENSG)

export MKL_NUM_THREADS=1;     export NUMEXPR_NUM_THREADS=1;     export OMP_NUM_THREADS=1;     export OPENBLAS_NUM_THREADS=1;
    export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:resources/software/sdpr/MKL/lib;     export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:resources/software/sdpr/gsl/lib


opt$ref_plink_chr<-'resources/data/ref/ref.chr'
opt$ref_keep<-'resources/data/ref/keep_files/EUR.keep'
opt$gwas_pop<-'EUR'
opt$ref_pcs<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output/reference/pc_score_files/TRANS/ref-TRANS-pcs.profiles'
opt$sumstats<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output/reference/gwas_sumstat/ENSG00000002549/ENSG00000002549-cleaned.gz'
opt$output<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/output/reference/pgs_score_files/sdpr/ENSG00000002549/ref-ENSG00000002549'
opt$n_cores<-1
opt$pop_data<-'resources/data/ref/ref.pop.txt'
opt$sdpr<-'resources/software/sdpr/SDPR'
opt$test<-NA 

```

***

# Analysis Phase 8: Ensemble approach

```{r}
library(glmnet)
library(fastshap)
library(tibble)
library(data.table)
library(dplyr)
library(psych)

# Read in observed expression
obs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_expression_matrices/Whole_Blood.v8.normalized_expression.bed.gz')
obs$gene_id<-gsub('\\..*','',obs$gene_id)
obs<-t(obs[,-1:-3, with = F])
colnames(obs)<-obs[1,]
obs<-obs[-1,]
obs<-cbind(row.names(obs),obs)
colnames(obs)[1]<-'ID'
obs<-data.table(obs)
obs<-cbind(obs[,1],data.frame(lapply(obs[,-1], function(x) as.numeric(as.character(x)))))

both_genes<-intersect(names(obs)[-1], genes)

# Read in covariates
covs<-fread('/users/k1806347/oliverpainfel/Data/GTeX/v8/GTEx_Analysis_v8_eQTL_covariates/Whole_Blood.v8.covariates.txt')
covs<-t(covs)
colnames(covs)<-covs[1,]
covs<-covs[-1,]
covs<-cbind(row.names(covs),covs)
colnames(covs)[1]<-'ID'
covs<-data.table(covs)
covs<-cbind(covs[,1],data.frame(lapply(covs[,-1], function(x) as.numeric(as.character(x)))))

# Merge observed expression and covariates
obs<-obs[,!duplicated(names(obs)),with=F]
obs_covs<-merge(obs, covs, by='ID')

# Remove GTEx v7 individuals
gtex_v7 <- fread('/users/k1806347/oliverpainfel/Data/GTeX/GTEx_v7_Annotations_SampleAttributesDS.txt')
gtex_v7$ID <- sub("^([^-]+-[^-]+)-.*$", "\\1", gtex_v7$SAMPID)
obs_covs <- obs_covs[!(obs_covs$ID %in% gtex_v7$ID),]

# Residualise the covariates
obs_resid<-data.frame(ID=obs_covs$ID, stringsAsFactors=F)
for(i in both_genes){
  obs_resid[[i]]<-as.numeric(scale(resid(lm(as.formula(paste0(i,' ~ ', paste(names(covs)[-1], collapse=' + '))), data=obs_covs))))
}
obs_resid<-data.table(obs_resid)
expression_mat <- obs_resid

###
# Read in prediction expression
###

# Compare predictions from GenoPred and EXPRESSO directly
setwd('~/oliverpainfel/Software/MyGit/GenoPred/pipeline/')
source('../functions/misc.R')
source_all('../functions')
library(data.table)

config<-'/scratch/prj/bcn_gtex_eqtl_twas/scratch_tmp/GenoPred/config/config_cis_only.yaml'
pgs_methods <- read_param(config = config, param = 'pgs_methods', return_obj = F)
outdir <- read_param(config = config, param = 'outdir', return_obj = F)

# Read in PGS
pgs_genopred <- read_pgs(config = config, pop = 'EUR', pseudo_only = T)$gtex_v8$EUR

pred_list <- lapply(pgs_genopred, function(x) {
  tmp <- Reduce(function(dtf1, dtf2) merge(dtf1, dtf2, by = c('FID','IID'), all.x = TRUE), x)
  names(tmp) <- c('FID','ID', names(x))
  tmp$FID<-NULL
  tmp
})

# Get a list of genes
genes <- fread('~/oliverpainfel/Data/eQTLGen/full/cis_only/gene_list.txt', header=F)$V1
genes <- genes[genes %in% names(expression_mat)]

#–––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––
# 2) Loop over genes, fit elastic‐net & compute SHAP
#–––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––

# 1) Define the testing function
test_gene_williams <- function(y, X, alpha = 0.5) {
  X_df <- as.data.frame(X)
  stopifnot(!any(is.null(colnames(X_df))))
  
  # Fit elastic‐net stacking
  cvf <- cv.glmnet(
    x           = as.matrix(X_df), 
    y           = as.numeric(y), 
    alpha       = alpha, 
    standardize = TRUE
  )
  pred_ens <- predict(cvf$glmnet.fit, 
                      newx = as.matrix(X_df), 
                      s    = cvf$lambda.min)[,1]
  
  # If pred_ens is constant, correlations undefined → NAs
  if (sd(pred_ens) == 0) {
    return(
      tibble(
        method = colnames(X_df),
        r_ens  = NA_real_,
        r_m    = NA_real_,
        r_pred = NA_real_,
        p_val  = NA_real_
      )
    )
  }
  
  # Compute ensemble correlation
  r_ens <- cor(y, pred_ens, use = "complete.obs")
  n     <- length(y)
  
  # Build results per method
  out <- tibble(method = colnames(X_df)) %>%
    rowwise() %>%
    mutate(
      r_m    = cor(y,        X_df[[method]],   use = "complete.obs"),
      r_pred = cor(pred_ens, X_df[[method]],   use = "complete.obs"),
      p_val  = {
        if (is.na(r_m) || is.na(r_ens) || is.na(r_pred)) {
          NA_real_
        } else {
          psych::paired.r(
            xy       = r_ens,
            yz       = r_m,
            xz       = r_pred,
            n         = n,
            twotailed = FALSE
          )$p
        }
      }
    ) %>%
    ungroup() %>%
    select(method, r_ens, r_m, r_pred, p_val)
  
  return(out)
}


test_gene_williams <- function(y, X, alpha = 0.5) {
  # 1) Ensure y is a plain numeric vector
  y_vec <- as.numeric(y)
  stopifnot(length(y_vec) > 1)
  
  # 2) Ensure X is a data.frame with colnames
  X_df <- as.data.frame(X)
  stopifnot(!is.null(colnames(X_df)))
  
  # 3) Fit elastic‐net on the full data
  cvf <- cv.glmnet(
    x           = as.matrix(X_df),
    y           = y_vec,
    alpha       = alpha,
    standardize = TRUE
  )
  preds_ens <- predict(
    cvf$glmnet.fit,
    newx = as.matrix(X_df),
    s    = cvf$lambda.min
  )[,1]
  
  # 4) If ensemble is constant → return NAs
  if (sd(preds_ens) == 0) {
    return(
      tibble(
        method = colnames(X_df),
        r_ens  = NA_real_,
        r_m    = NA_real_,
        r_pred = NA_real_,
        p_val  = NA_real_
      )
    )
  }
  
  # 5) Compute ensemble correlation
  r_ens <- cor(y_vec, preds_ens, use = "complete.obs")
  n     <- length(y_vec)
  
  # 6) Build per‐method stats
  out <- tibble(method = colnames(X_df)) %>%
    rowwise() %>%
    mutate(
      # correlation of this method with truth
      r_m    = cor(y_vec, X_df[[method]], use = "complete.obs"),
      # correlation of this method with the ensemble
      r_pred = cor(preds_ens, X_df[[method]], use = "complete.obs"),
      # Williams test (one‐sided: ensemble > method)
      p_val  = if (any(is.na(c(r_ens, r_m, r_pred)))) {
                 NA_real_
               } else {
                 psych::paired.r(
                  xy       = r_ens,
                  yz       = r_m,
                  xz       = r_pred,
                   n         = n,
                   twotailed = FALSE
                 )$p
               }
    ) %>%
    ungroup() %>%
    # 7) add the ensemble correlation as a column
    mutate(r_ens = r_ens) %>%
    # 8) pick and order your columns
    select(method, r_ens, r_m, r_pred, p_val)
  
  return(out)
}

set.seed(123)  # for reproducibility of SHAP sampling
importance_list <- vector("list", length(pred_list))
names(importance_list) <- names(pred_list)
cor_res<-NULL

for (gene in genes[-1:-12]) {
  print(which(genes == gene))
  
  merged <- merge(expression_mat, pred_list[[gene]], by = 'ID')
  # a) Response & predictors for this gene
  y <- merged[, gene, with=F]
  X <- merged[, names(pred_list[[gene]])[-1], with = F]
  X[is.na(X)] <- 0
  
  # b) Fit Elastic Net (α = 0.5; you can tune α if desired)
  cvfit      <- cv.glmnet(as.matrix(X), as.matrix(y), alpha = 0.5)
  fit        <- cvfit$glmnet.fit
  lambda_min <- cvfit$lambda.min
  
  # Estimate correlation between ensemble model predictions and observed
  preds_ens <- predict(
    cvfit$glmnet.fit,
    newx = as.matrix(X),
    s    = cvfit$lambda.min
  )[,1]
  
  cor_res <- rbind(cor_res,
                   data.frame(gene = gene,
                              r_ens = cor(preds_ens, as.numeric(y))[1]))
    
  # c) Wrap the predict() call so fastshap knows how to score new data
  pred_wrapper <- function(object, newdata) {
    predict(object, newx = as.matrix(newdata), s = lambda_min)[,1]
  }
  
  # d) Compute SHAP values (nsim = Monte‐Carlo draws; increase for stability)
  phi <- fastshap::explain(
    object       = fit,
    X            = as.data.frame(X),
    pred_wrapper = pred_wrapper,
    nsim         = 200
  )
  
  # e) Aggregate to global importances
  global_imp <- colMeans(abs(phi))
  
  # f) Store as a tibble
  importance_list[[gene]] <- 
    tibble(
      gene       = gene,
      method     = names(global_imp),
      importance = as.numeric(global_imp)
    )
}

#–––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––
# 3) Combine results across genes
#–––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––
importance_df <- bind_rows(importance_list)

# Now you have a data.frame with columns: gene, method, importance.
# You can, for example, compute mean importance per method across all genes:
avg_imp_by_method <- 
  importance_df %>% 
  group_by(method) %>% 
  summarize(avg_importance = mean(importance)) %>% 
  arrange(desc(avg_importance))

print(avg_imp_by_method)

# Ensure methods are ordered by decreasing importance
avg_imp_by_method <- avg_imp_by_method %>%
  dplyr::arrange(desc(avg_importance)) %>%
  dplyr::mutate(method = factor(method, levels = method))

# Plot
library(ggplot2)
ggplot(avg_imp_by_method, aes(x = method, y = avg_importance, group = 1)) +
  geom_line() +
  geom_point(size = 3) +
  labs(
    x = "PGS Method (ranked)",
    y = "Average SHAP Importance",
    title = "Elbow Plot of PGS Method Importances"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


set.seed(123)  # for reproducibility of SHAP sampling
cor_res<-NULL

for (gene in genes) {
  print(which(genes == gene))
  
  merged <- merge(expression_mat, pred_list[[gene]], by = 'ID')
  # a) Response & predictors for this gene
  y <- merged[, gene, with=F]
  X <- merged[, names(pred_list[[gene]])[-1], with = F]
  X[is.na(X)] <- 0
  
  # b) Fit Elastic Net (α = 0.5; you can tune α if desired)
  cvfit      <- cv.glmnet(as.matrix(X), as.matrix(y), alpha = 0.5)
  fit        <- cvfit$glmnet.fit
  lambda_min <- cvfit$lambda.min
  
  # Estimate correlation between ensemble model predictions and observed
  preds_ens <- predict(
    cvfit$glmnet.fit,
    newx = as.matrix(X),
    s    = cvfit$lambda.min
  )[,1]
  
  cor_res <- rbind(cor_res,
                   data.frame(gene = gene,
                              r_ens = cor(preds_ens, y)[1]))
    
}

# This shows me that the gain from the multi model is not large, but it is always best. Not suprising given this is cross validation. It would be more convincing if we could show an improvement in an independent dataset, or do the tuning with sumstats alone.

```