---
title: "Comparison with OTTERS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

library(knitr)
library(data.table)

```

***

In this document we will make comparisons to the OTTERS software.

***

# Install

```{bash}
# Install OTTERS v1.0.0
cd /users/k1806347/oliverpainfel/Software
git clone https://github.com/daiqile96/OTTERS.git 
cd OTTERS
git checkout v1.0.0

# Install tabix and bgzip
cd /users/k1806347/oliverpainfel/Software
wget https://sourceforge.net/projects/samtools/files/tabix/tabix-0.2.6.tar.bz2
tar jxvf tabix-0.2.6.tar.bz2
cd tabix-0.2.6
make

# copy the binary bgzip and tabix bin folder so can be called from PATH
mkdir /users/k1806347/oliverpainfel/Software/bin
cp bgzip tabix /users/k1806347/oliverpainfel/Software/bin

# copy the plink1.9 binary and add to PATH
cp /users/k1806347/oliverpainfel/Software/plink1.9/plink /users/k1806347/oliverpainfel/Software/bin/
export PATH=/users/k1806347/oliverpainfel/Software/bin:$PATH 

# Create a conda environment to run otters
cd /users/k1806347/oliverpainfel/Software/OTTERS
mamba env create --file otters_env.yaml
conda activate otters

# Install lassosum
Rscript -e 'remotes::install_github("tshmak/lassosum@v0.4.5")'

# Install SDPR
cd /users/k1806347/oliverpainfel/Software/bin
git clone https://github.com/eldronzhou/SDPR.git
cd SDPR
git checkout v0.9.1
```

***

# Run test data

```{bash}
# Run using test data
conda activate otters

# set number of threads to be used
N_THREADS=1

# set up my OTTERS directory and SDPR directory
OTTERS_DIR=/users/k1806347/oliverpainfel/Software/OTTERS
SDPR_DIR=/users/k1806347/oliverpainfel/Software/bin/SDPR

# make sure the dynamic libraries of SDPR are not changed (For SDPR)
export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:${SDPR_DIR}/MKL/lib
export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:${SDPR_DIR}/gsl/lib

# prevent automatically using  all available cores on a compute node (For SDPR and PRS-CS)
export MKL_NUM_THREADS=$N_THREADS
export NUMEXPR_NUM_THREADS=$N_THREADS
export OMP_NUM_THREADS=$N_THREADS

# Start to run OTTERS
cd ${OTTERS_DIR}/Example

# Input for OTTERS STAGE I 
# Annotation File 
exp_anno=exp_anno.txt
# Genotype data from LD reference panel
geno_dir=Exp_geno
# eQTL summary statistics 
sst_file=Exp_eQTLSumStats.txt
# Input for OTTERS STAGE II
# GWAS summary statistics 
gwas_sst_file=Exp_GWASSumStats.txt

# Set chromosome number (The example is for chromosome 4)
chr=4
# Set LD-clumping threshold in STAGE I
clump_r2=0.99
# Set output directory for STAGE I
out_dir=Results

# STAGE I
# train eQTL weights using P+T, lassosum, SDPR and PRS-CS. 
# It may take several minutes to complete.
python3 ${OTTERS_DIR}/training.py \
--OTTERS_dir=${OTTERS_DIR} \
--SDPR_dir=${SDPR_DIR} \
--anno_dir=${exp_anno} \
--geno_dir=${geno_dir} \
--sst_file=${sst_file} \
--out_dir=${out_dir} \
--chrom=${chr} \
--r2=${clump_r2} \
--models=PT,lassosum,SDPR,PRScs \
--thread=$N_THREADS

# It has completed with the test data
# Do a speed and memory benchmark for 100 genes from YFS eQTL data
```

***

# Replicate OTTERS paper

Try to replicate the results provided by OTTERS for cardiovascular disease.

***

## Download eQTLGen weights

```{bash}
# Download the eQTLGen weights and results in the OTTERS paper: https://www.synapse.org/#!Synapse:syn51009573/files/
mkdir -p /users/k1806347/oliverpainfel/Data/OTTERS/eQTLGen
cd /users/k1806347/oliverpainfel/Data/OTTERS/eQTLGen
```

```{python}
import shutil
import os
from synapseclient import Synapse

# Read in auth_token
with open('/users/k1806347/oliverpainfel/Data/synapse/auth_token.txt', 'r') as file:
    content = file.read()

# Initialize the Synapse client and log in
syn = Synapse()
syn.login(authToken=content)

# List of Synapse IDs you want to download and move
syn_ids = [
    'syn51015448',
    'syn51015449',
    'syn51015450',
    'syn51015456',
    'syn51015458',
    'syn51015464'
]

# Define the new directory where you want to store the files
new_directory = '/users/k1806347/oliverpainfel/Data/OTTERS/eQTLGen'

for syn_id in syn_ids:
    # Download the file from Synapse
    entity = syn.get(entity=syn_id, version=1)
    
    # Get the original path and filename
    original_path = entity.path
    filename = os.path.basename(original_path)
    
    # Construct the new path in the desired directory with the same filename
    new_path = os.path.join(new_directory, filename)
    
    # Move the file
    shutil.move(original_path, new_path)
    
    print(f"Moved {filename} to {new_path}")
    
    
```

```{bash}
# Decompress the eQTLGen weights
cd /users/k1806347/oliverpainfel/Data/OTTERS/eQTLGen 

gunzip lassosum.txt.gz
gunzip OTTERS_TWAS_SumStat_CARDIOVASCULAR_UKBB.txt
gunzip P0.001.txt.gz
gunzip P0.05.txt.gz
gunzip PRScs.txt.gz
gunzip SDPR.txt.gz

# Note the weights appear to be GRCh38
# We need to liftover the reference genotype data and GWAS sumstats
```

***

## Prepare input files

```{r}
# Create annotation file using re
library(data.table)
res<-fread('/users/k1806347/oliverpainfel/Data/OTTERS/eQTLGen/OTTERS_TWAS_SumStat_CARDIOVASCULAR_UKBB.txt')
annot<-res[, c('CHROM','GeneStart','GeneEnd','TargetID'), with=F]
fwrite(annot, '/users/k1806347/oliverpainfel/Data/OTTERS/eQTLGen/annot.txt', sep='\t', quote=F, na='NA')
```

```{bash}
# Download and format the GWAS sumstats
# The sumstats were downloaded using gsutils from the Akes Price google cloud: gs://broad-alkesgroup-public-requester-pays/UKBB/disease_CARDIOVASCULAR.sumstats.gz
```

```{r}
# Liftover to GRCh38
library(data.table)

ss<-fread('/users/k1806347/oliverpainfel/Data/OTTERS/gwas_sumstats/disease_CARDIOVASCULAR.sumstats.gz')

# Calculate Z from P value using sign of BETA
ss$Z<-sign(ss$Beta)*qnorm(ss$P / 2, lower.tail=F)

ss<-ss[, c('CHR','POS','A1','A2','Z'), with=F]
names(ss)<-c('chr','pos','A1','A2','Z')

# Create snp_modifyBuild_offline
make_executable <- function(exe) {
  Sys.chmod(exe, mode = (file.info(exe)$mode | "111"))
}

snp_modifyBuild_offline<-function (info_snp, liftOver, chain, from = "hg18", to = "hg19"){
  tmp_folder<-tempdir()

  if (!all(c("chr", "pos") %in% names(info_snp)))
    stop2("Please use proper names for variables in 'info_snp'. Expected %s.",
          "'chr' and 'pos'")
  liftOver <- normalizePath(liftOver)
  make_executable(liftOver)
  BED <- tempfile(fileext = ".BED")
  info_BED <- with(info_snp, data.frame(paste0("chr", chr),
                                        pos0 = pos - 1L, pos, id = seq_len(nrow(info_snp))))
  fwrite(info_BED, BED, col.names = FALSE, sep = " ")
  lifted<-paste0(tmp_folder,'/tmp.lifted')
  unmapped<-paste0(tmp_folder,'/tmp.unmapped')
  system(paste(liftOver, BED, chain, lifted, unmapped))
  new_pos <- fread(lifted)
  bad <- grep("^#", readLines(unmapped), value = TRUE, invert = TRUE)
  print(paste0(length(bad)," variants have not been mapped."))
  info_snp$pos <- NA
  info_snp$pos[new_pos$V4] <- new_pos$V3
  info_snp
}

ss<-snp_modifyBuild_offline(
  info_snp=ss,
  liftOver='/users/k1806347/oliverpainfel/Software/liftOver',
  chain = '/users/k1806347/oliverpainfel/Software/MyGit/GenoDisc/pipeline/resources/data/liftover/hg19ToHg38.over.chain.gz',
  from = "hg19",
  to = "hg38"
)

names(ss)<-c('CHROM','POS','A1','A2','Zscore')
ss<-ss[, c('CHROM','POS','A1','A2','Zscore'), with=F]
ss<-ss[complete.cases(ss),]

fwrite(ss, '/users/k1806347/oliverpainfel/Data/OTTERS/gwas_sumstats/disease_CARDIOVASCULAR_buildGRCh38.reformat.sumstats.tsv', quote=F, sep='\t', na='NA')
```

***

## Run OTTERS

```{bash}
conda activate otters
# set number of threads to be used
N_THREADS=1

# set up my OTTERS directory and SDPR directory
OTTERS_DIR=/users/k1806347/oliverpainfel/Software/OTTERS
SDPR_DIR=/users/k1806347/oliverpainfel/Software/bin/SDPR

# Input for OTTERS STAGE I 
# Annotation File 
exp_anno=/users/k1806347/oliverpainfel/Data/OTTERS/eQTLGen/annot.txt

# Input for OTTERS STAGE II
# GWAS summary statistics 
gwas_sst_file=/users/k1806347/oliverpainfel/Data/OTTERS/gwas_sumstats/disease_CARDIOVASCULAR_buildGRCh38.reformat.sumstats.tsv

for model in $(echo P0.001 P0.05 lassosum SDPR PRScs); do
for chr in $(seq 1 22); do
  sbatch -p neurohack_cpu,cpu -t120 --mem 10G --wrap="

# make sure the dynamic libraries of SDPR are not changed (For SDPR)
export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:${SDPR_DIR}/MKL/lib
export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:${SDPR_DIR}/gsl/lib
export PATH=/users/k1806347/oliverpainfel/Software/bin:$PATH 

# prevent automatically using  all available cores on a compute node (For SDPR and PRS-CS)
export MKL_NUM_THREADS=$N_THREADS
export NUMEXPR_NUM_THREADS=$N_THREADS
export OMP_NUM_THREADS=$N_THREADS

# STAGE II
# gene-based association test using eQTL-weight trained from P+T, lassosum, SDPR and PRS-CS.
mkdir -p /users/k1806347/oliverpainfel/Analyses/eQTL_to_TWAS/data/OTTERS/gtex_ref/chr${chr}/${model}
python3 ${OTTERS_DIR}/testing.py \
  --OTTERS_dir=${OTTERS_DIR} \
  --weight_dir=/users/k1806347/oliverpainfel/Data/OTTERS/eQTLGen \
  --models=${model} \
  --anno_dir=${exp_anno} \
  --geno_dir=/users/k1806347/oliverpainfel/Data/dbGAP/GTEx/processed/gtex_v8.snp.maf001.chr${chr} \
  --out_dir=/users/k1806347/oliverpainfel/Analyses/eQTL_to_TWAS/data/OTTERS/gtex_ref/chr${chr}/${model} \
  --gwas_file=${gwas_sst_file} \
  --chrom=${chr} \
  --thread=$N_THREADS
"
done
done

```

***

## Compare with original results

```{r}
library(data.table)
library(ACAT)

# Create function performing genomic control
# This is what they did in the OTTERS paper
genomic_control<-function(z){
  Lambda = median(z^2, na.rm = T)/qchisq(0.5, df = 1)
  P = pchisq(z^2/Lambda, df = 1, lower.tail = F)
}

# Read in results
res_gtex <- NULL
for (j in c('P0.001', 'P0.05', 'lassosum', 'SDPR', 'PRScs')) {
  for (i in 22:22) {
    tmp <- fread(
        paste0(
          '/users/k1806347/oliverpainfel/Analyses/eQTL_to_TWAS/data/OTTERS/gtex_ref/chr',
          i,
          '/',
          j,
          '/',
          j,
          '.txt'
        )
      )
  }
  tmp$MODEL<-j
  tmp$FUSION_PVAL_GC<-genomic_control(tmp$FUSION_Z)
  res_gtex <- rbind(res_gtex, tmp)
}

write.csv(res_gtex, '/users/k1806347/oliverpainfel/Analyses/eQTL_to_TWAS/data/OTTERS/rep_res_gtex_long.csv', row.names=F)

# Perform ACAT.O analysis
res_gtex<-res_gtex[complete.cases(res_gtex),]
res_gtex_acat<-data.table(aggregate(res_gtex$FUSION_PVAL_GC, list(res_gtex$TargetID), FUN=ACAT))
names(res_gtex_acat)<-c('TargetID', 'OTTERS')

library(reshape2)
res_gtex_wide <- dcast(res_gtex, TargetID ~ MODEL, value.var = "FUSION_PVAL_GC")
res_gtex_wide<-merge(res_gtex_wide, res_gtex_acat, by = 'TargetID')

write.csv(res_gtex_wide, '/users/k1806347/oliverpainfel/Analyses/eQTL_to_TWAS/data/OTTERS/rep_res_gtex_wide.csv', row.names=F)

###############

# Read in original results
orig<-fread('/users/k1806347/oliverpainfel/Data/OTTERS/eQTLGen/OTTERS_TWAS_SumStat_CARDIOVASCULAR_UKBB.txt')

orig<-orig[, names(res_gtex_wide), with=F]

both<-Reduce(function(dtf1, dtf2) merge(dtf1, dtf2, by = "TargetID", all=T), list(orig,res_gtex_wide))

# Convert to Z scores, and check correlations between each test
both_z<-apply(both[,-1], 2, function(x) qnorm(x))
cor(both_z, use='p')
summary(both_z)

# There are missing values in the original results which are not missing in the replication
# For those that are present, the correlation is <1, probably due to missing SNPs and different LD reference.

```

Limitations of OTTERS:
- Slow
- Requires consistent builds across files (not RSID based)
- Not restricted to HapMap3
- Format doesn't integrate well with existing downstream TWAS methodology
- No direction of effect in ACAT result

***

# Compare to eQTL_to_TWAS

OK, now we have replicated the original results in the OTTERS paper, lets compare the results to those generated using our approach.

Run TWAS using eQTLGen weights I generated and the same sumstats as the OTTERS paper.

```{r}
library(data.table)
# Munge the sumstats
ss<-fread('/users/k1806347/oliverpainfel/Data/OTTERS/gwas_sumstats/disease_CARDIOVASCULAR.sumstats.gz')
# Calculate Z from P value using sign of BETA
ss$Z<-sign(ss$Beta)*qnorm(ss$P / 2, lower.tail=F)
ss<-ss[, c('SNP','A1','A2','Z'), with=F]
fwrite(ss, '/users/k1806347/oliverpainfel/Data/OTTERS/gwas_sumstats/disease_CARDIOVASCULAR.sumstats.munged.tsv.gz', quote=F, sep='\t', na='NA')
```

```{bash}
# Run FUSION
mkdir /users/k1806347/oliverpainfel/Analyses/eQTL_to_TWAS/data/OTTERS/eQTL_to_TWAS
conda activate eqtl_to_twas
for chr in $(seq 22 22); do
    sbatch -p neurohack_cpu,cpu -t120 --mem 10G --wrap="
    Rscript /users/k1806347/oliverpainfel/Software/MyGit/eQTL_to_TWAS/FUSION.assoc_test.edit.R \
    --sumstats /users/k1806347/oliverpainfel/Data/OTTERS/gwas_sumstats/disease_CARDIOVASCULAR.sumstats.munged.tsv.gz \
    --weights /users/k1806347/oliverpainfel/Data/eQTLGen/eQTL_to_TWAS/eQTLGen.eQTL.pos \
    --weights_dir /users/k1806347/oliverpainfel/Data/eQTLGen/eQTL_to_TWAS \
    --ref_ld_chr /users/k1806347/oliverpainfel/Data/1KG/FUSION/1000G.EUR. \
    --out /users/k1806347/oliverpainfel/Analyses/eQTL_to_TWAS/data/OTTERS/eQTL_to_TWAS/disease_CARDIOVASCULAR.eQTLGen.chr${chr} \
    --chr ${chr} \
    --GWASN 459324
    "
done
```

```{r}
library(data.table)
library(ACAT)

# Read in the OTTERS replication results
otters<-fread('/users/k1806347/oliverpainfel/Analyses/eQTL_to_TWAS/data/OTTERS/rep_res_gtex_long.csv')
otters<-otters[,c('TargetID','MODEL','FUSION_Z','FUSION_PVAL'), with=F]
names(otters)<-c('ID','MODEL','TWAS.Z', 'TWAS.P')

# Read in the eQTL_to_TWAS results
eqt_to_twas<-fread('/users/k1806347/oliverpainfel/Analyses/eQTL_to_TWAS/data/OTTERS/eQTL_to_TWAS/GCST90029019.eQTLGen.chr22')
eqt_to_twas<-eqt_to_twas[, c('ID','MODEL','TWAS.Z', 'TWAS.P'), with=F]

# Estimate ACAT P value
otters<-otters[complete.cases(otters),]
otters_acat<-data.table(aggregate(otters$TWAS.P, list(otters$ID), FUN=ACAT))
names(otters_acat)<-c('ID', 'ACAT')

eqt_to_twas<-eqt_to_twas[complete.cases(eqt_to_twas),]
eqt_to_twas_acat<-data.table(aggregate(eqt_to_twas$TWAS.P, list(eqt_to_twas$ID), FUN=ACAT))
names(eqt_to_twas_acat)<-c('ID', 'ACAT')

# Compare the ACAT P-values between OTTERS and eqtl_to_twas
all<-Reduce(function(dtf1, dtf2)
  merge(dtf1, dtf2, by = "ID", all=T),
  list(otters_acat, eqt_to_twas_acat))

all_z<-apply(all[,-1], 2, function(x) qnorm(x))
cor(all_z, use='p')
summary(all_z)

# Check how many are p<0.001
sum(all$ACAT.x < 0.001, na.rm=T)
sum(all$ACAT.y < 0.001, na.rm=T)

table(all$ACAT.x < 0.001, all$ACAT.y < 0.001, useNA = 'ifany')

```




