---
title: "Comparison of Methods"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

# Introduction

Here, we will evaluate methods for deriving TWAS SNP-weights based on eQTL summary statistics.

There are two stages to this process, 1) Determining which genes to derive TWAS SNP-weights for, and 2) Derive SNP-weights predicting gene expression.

***

## Determining which genes to derive TWAS SNP-weights for

Existing approaches:

- For SMR this is done by analysing any gene with a genome-wide significant eQTL
- For FUSION this is done by analysing any gene with cis-SNPh2 p < 0.01
- For MetaXcan this is done by analysing any gene with significant positive R2 using SNP-weights

The FUSION and MetaXcan approach is better for determining genes to include in a GWAS as it better reflects the power that TWAS would have. We may not always have a target sample to test the variance explained by the SNP-weights, so the MetaXcan approach is less applicable. So, I think we should use the same criteria as FUSION, except we must use a summary statistic based approach to estimate SNP-based heritability. We should compare summary statistic approaches to estimates from GREML as reported in FUSION released SNP-weights. The approach also need to be fast as it will need to be implemented for each gene seperately.

***

## Derive SNP-weights predicting gene expression

SMR uses single eQTL summary statistics, making it applicable to a wider range of datasets and results from meta-analyses. TWAS based methods are not currently relevent when only eQTL summary statistics are available, as the advantage of these methods over SMR is their ability to incorporate the effects of multiple SNPs on gene expression. Currently, TWAS methods use SNP-weights derived using individual level data, which means TWAS cannot use meta-analysis results for eQTLs, and often TWAS weights are unavailable for the latest eQTL datasets. So, here we will compare a range of summary statistic approaches to derive TWAS weights. Again, the method will need to be fast since it will need to be applied to each genes seperately. Summary statistic based polygenic scoring methods will likely be of use, though given the lack of target samples for validation, a pseudovalidation approach will be necessary, which estimates the optimal parameters without a target sample for validation.

Possible methods for deriving SNP-weights:
- SBayesR
- DBSLMM
- PRScs auto
- Mega-PRS

An alternative method suggested is to use SuSiE posteriors as SNP-weights.

***

# Methods

***

## Comparison using GTEx v7 eQTL data and FUSION SNP-weights

Comparison to FUSION weights is not ideal as by chance different leads SNP could be selected. Ideally we would compare summary statistic-based and observed TWAS weights to observed expression. However, lets start with this as it will allow comparison of heritability estimates and be a good opputunity to design the study.

We will use whole blood for the comparison. eQTL summary statistics are downloaded from [here](https://storage.googleapis.com/gtex_analysis_v7/single_tissue_eqtl_data/all_snp_gene_associations/Whole_Blood.allpairs.txt.gz). FUSION TWAS SNP-weights were downloaded from [here](http://gusevlab.org/projects/fusion/weights/GTEx.Whole_Blood.P01.tar.bz2).

***

### Compare SNP-h2 estimates

We will compare different version of GCTB Bayes models and LDSC. If LDSC doesn't work, then LDAK model is unlikely to make a difference. To avoid unessecary compuational burden, only estimate hsq for 100 genes on chromosome 22.

```{bash, eval=F, echo=T}

for chr in $(seq 1 22);do
pattern=$(echo ^${chr}_)
awk -v pat="$pattern" '$2 ~ pat { print }' /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/Whole_Blood.allpairs.txt > /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/Whole_Blood.allpairs.chr${chr}.txt_tmp
cat <(head -n1 /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/Whole_Blood.allpairs.txt) /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/Whole_Blood.allpairs.chr${chr}.txt_tmp > /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/Whole_Blood.allpairs.chr${chr}.txt
rm /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/Whole_Blood.allpairs.chr${chr}.txt_tmp
done

```

```{r, eval=F, echo=T}
library(data.table)

# Make output directory
dir.create('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR')
dir.create('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesS')
dir.create('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/LDSC')
dir.create('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR/RDat_files')

# Read in HapMap3 SNP-list
hm3_snp<-fread('/users/k1806347/brc_scratch/Data/ldsc/w_hm3.snplist')

sbayesr_h2<-NULL
sbayesr_robust_h2<-NULL
sbayess_h2<-NULL
ldsc_h2<-NULL

# Run across each chromosome seperately
for(chr_i in 22){
  # Read in eQTL data
  eqtl<-fread(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/Whole_Blood.allpairs.chr',chr_i,'.txt'))
  
  # Create CHR, BP, A1, A2, and Build columns
  var_id<-data.table(do.call(rbind, strsplit(eqtl$variant_id, '_')))
  names(var_id)<-c('CHR','BP','A1','A2','Build')
  var_id$CHR<-as.numeric(var_id$CHR)
  var_id$BP<-as.numeric(var_id$BP)
  
  eqtl<-cbind(eqtl, var_id)
  
  # Insert IUPAC codes for SNPs
  eqtl$IUPAC[eqtl$A1 == 'A' & eqtl$A2 =='T' | eqtl$A1 == 'T' & eqtl$A2 =='A']<-'W'
  eqtl$IUPAC[eqtl$A1 == 'C' & eqtl$A2 =='G' | eqtl$A1 == 'G' & eqtl$A2 =='C']<-'S'
  eqtl$IUPAC[eqtl$A1 == 'A' & eqtl$A2 =='G' | eqtl$A1 == 'G' & eqtl$A2 =='A']<-'R'
  eqtl$IUPAC[eqtl$A1 == 'C' & eqtl$A2 =='T' | eqtl$A1 == 'T' & eqtl$A2 =='C']<-'Y'
  eqtl$IUPAC[eqtl$A1 == 'G' & eqtl$A2 =='T' | eqtl$A1 == 'T' & eqtl$A2 =='G']<-'K'
  eqtl$IUPAC[eqtl$A1 == 'A' & eqtl$A2 =='C' | eqtl$A1 == 'C' & eqtl$A2 =='A']<-'M'
  
  # Remove INDELS
  eqtl<-eqtl[!is.na(eqtl$IUPAC),]
  
  # Calculate N for each association
  eqtl$N<-round((eqtl$ma_count/eqtl$maf)/2,1)
  
  # Check the build
  table(eqtl$Build) 
  # They are all b37
  
  ###
  # Insert RSIDs
  ###
  
  # Read in 1KG reference SNP data
  ref_bim_i<-fread(paste0('/mnt/lustre/groups/biomarkers-brc-mh/Reference_data/1KG_Phase3/PLINK/all_phase3.chr',chr_i,'.bim'))
  names(ref_bim_i)<-c('CHR','SNP','POS','BP','A1','A2')
  
  # Extract hm3 SNPs
  ref_bim_i<-ref_bim_i[ref_bim_i$SNP %in% hm3_snp$SNP,]
  
  # Insert IUPAC codes
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'A' & ref_bim_i$A2 =='T' | ref_bim_i$A1 == 'T' & ref_bim_i$A2 =='A']<-'W'
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'C' & ref_bim_i$A2 =='G' | ref_bim_i$A1 == 'G' & ref_bim_i$A2 =='C']<-'S'
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'A' & ref_bim_i$A2 =='G' | ref_bim_i$A1 == 'G' & ref_bim_i$A2 =='A']<-'R'
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'C' & ref_bim_i$A2 =='T' | ref_bim_i$A1 == 'T' & ref_bim_i$A2 =='C']<-'Y'
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'G' & ref_bim_i$A2 =='T' | ref_bim_i$A1 == 'T' & ref_bim_i$A2 =='G']<-'K'
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'A' & ref_bim_i$A2 =='C' | ref_bim_i$A1 == 'C' & ref_bim_i$A2 =='A']<-'M'

  # Merge target and reference based on position
  ref_target<-merge(eqtl, ref_bim_i, by=c('CHR','BP'))
  
  # I have used the GTEx reference data as well, and the same number is returned. 
  # Note: The A1 allele is the ref allele.
  # Small number of SNPs remaining highlights the limitation of using HapMap3 SNPs only.
  
  # Identify SNPs for which alleles need to be flipped
  flip_tmp<-ref_target[(ref_target$IUPAC.x == 'R' & ref_target$IUPAC.y == 'Y' | 
                          ref_target$IUPAC.x == 'Y' & ref_target$IUPAC.y == 'R' | 
                          ref_target$IUPAC.x == 'K' & ref_target$IUPAC.y == 'M' |
                          ref_target$IUPAC.x == 'M' & ref_target$IUPAC.y == 'K'),]
  
  # Idenitfy SNPs which match the reference alleles
  incl<-ref_target[ ref_target$IUPAC.x == 'R' & ref_target$IUPAC.y == 'R' | 
                      ref_target$IUPAC.x == 'Y' & ref_target$IUPAC.y == 'Y' | 
                      ref_target$IUPAC.x == 'K' & ref_target$IUPAC.y == 'K' |
                      ref_target$IUPAC.x == 'M' & ref_target$IUPAC.y == 'M' ]
  
  # If a SNP that needs to be flipped has a duplicate that is on the correct strand, remove it.
  flip<-flip_tmp[!(flip_tmp$SNP %in% incl$SNP)]
  
  # Flip alleles where necessary
  flip_tmp$A1_flipped<-flip_tmp$A1.x
  flip_tmp$A1_flipped[flip_tmp$A1.x == 'A']<-'T'
  flip_tmp$A1_flipped[flip_tmp$A1.x == 'T']<-'C'
  flip_tmp$A1_flipped[flip_tmp$A1.x == 'G']<-'C'
  flip_tmp$A1_flipped[flip_tmp$A1.x == 'C']<-'G'
  flip_tmp$A1.x<-flip_tmp$A1_flipped
  flip_tmp$A1_flipped<-NULL
  
  flip_tmp$A2_flipped<-flip_tmp$A2.x
  flip_tmp$A2_flipped[flip_tmp$A2.x == 'A']<-'T'
  flip_tmp$A2_flipped[flip_tmp$A2.x == 'T']<-'C'
  flip_tmp$A2_flipped[flip_tmp$A2.x == 'G']<-'C'
  flip_tmp$A2_flipped[flip_tmp$A2.x == 'C']<-'G'
  flip_tmp$A2.x<-flip_tmp$A2_flipped
  flip_tmp$A2_flipped<-NULL
  
  # Recombine and format
  eqtl_harm<-rbind(flip_tmp, incl)
  eqtl_harm<-eqtl_harm[,c('CHR','SNP','BP','A1.x','A2.x',"gene_id","variant_id","tss_distance","ma_samples","ma_count","maf","pval_nominal","slope","slope_se","N"), with=F]
  names(eqtl_harm)[names(eqtl_harm) == 'A1.x']<-'A1'
  names(eqtl_harm)[names(eqtl_harm) == 'A2.x']<-'A2'

  # Remove variants with MAF < 1%
  eqtl_harm<-eqtl_harm[eqtl_harm$maf >= 0.01,]
  
  eqtl<-eqtl_harm
  rm(eqtl_harm)
  
  # Identify unique genes 
  genes<-unique(eqtl$gene_id)
  genes<-genes[1:100]

  # Run for each gene seperately
  for(gene_i in genes){
    print(which(genes == gene_i))

    eqtl_gene_i<-eqtl[eqtl$gene_id == gene_i,]
    
    #######
    # SBayesR
    #######
    
    # Format for SBayesR
    eqtl_gene_i_sbayesr<-eqtl_gene_i[,c('SNP','A1','A2','maf','slope','slope_se','pval_nominal','N'), with=F]
    names(eqtl_gene_i_sbayesr)<-c('SNP','A1','A2','freq','b','se','p','N')
    
    # write in SBayesR format
    fwrite(eqtl_gene_i_sbayesr, paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR/',gene_i,'.txt'), sep=' ', na = "NA", quote=F)
    
    # Run SBayesR
    log<-system(paste0('/users/k1806347/brc_scratch/Software/gctb_2.03beta_Linux/gctb_203.sh --sbayes R --ldm /users/k1806347/brc_scratch/Data/1KG/Phase3/LD_matrix/GCTB_shared/ukbEURu_hm3_shrunk_sparse/ukbEURu_hm3_v3_50k_chr',chr_i,'.ldm.sparse --pi 0.95,0.02,0.02,0.01 --gamma 0.0,0.01,0.1,1 --gwas-summary /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR/',gene_i,'.txt --chain-length 10000 --exclude-mhc --burn-in 2000 --out-freq 1000 --out /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR/',gene_i,'.SBayesR'), intern=T)
    
    # Run SBayesR
    log2<-system(paste0('/users/k1806347/brc_scratch/Software/gctb_2.03beta_Linux/gctb_203.sh --sbayes R --ldm /users/k1806347/brc_scratch/Data/1KG/Phase3/LD_matrix/GCTB_shared/ukbEURu_hm3_shrunk_sparse/ukbEURu_hm3_v3_50k_chr',chr_i,'.ldm.sparse --pi 0.95,0.02,0.02,0.01 --gamma 0.0,0.01,0.1,1 --gwas-summary /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR/',gene_i,'.txt --chain-length 10000 --robust --exclude-mhc --burn-in 2000 --out-freq 1000 --out /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR/',gene_i,'.SBayesR_robust'), intern=T)
    
    #######
    # SBayesS
    #######
    
    # Run SBayesS
    log3<-system(paste0('/users/k1806347/brc_scratch/Software/gctb_2.03beta_Linux/gctb_203.sh --sbayes S --ldm /users/k1806347/brc_scratch/Data/1KG/Phase3/LD_matrix/GCTB_shared/ukbEURu_hm3_shrunk_sparse/ukbEURu_hm3_v3_50k_chr',chr_i,'.ldm.sparse --pi 0.1 --hsq 0.5 --chain-length 25000 --burn-in 5000 --gwas-summary /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR/',gene_i,'.txt --exclude-mhc --out-freq 1000 --out /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesS/',gene_i,'.SBayesS'), intern=T)
    
    #######
    # LDSC
    #######
    
    # Format for LDSC
    eqtl_gene_i_ldsc<-eqtl_gene_i[,c('SNP','A1','A2','slope','slope_se','N'), with=F]
    names(eqtl_gene_i_ldsc)<-c('SNP','A1','A2','BETA','SE','N')
    eqtl_gene_i_ldsc$Z<-eqtl_gene_i_ldsc$BETA/eqtl_gene_i_ldsc$SE
    eqtl_gene_i_ldsc<-eqtl_gene_i_ldsc[,c('SNP','A1','A2','Z','N'), with=F]
    
    # write in LDSC format
    fwrite(eqtl_gene_i_ldsc, paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/LDSC/',gene_i,'.txt'), sep=' ', na = "NA", quote=F)
    
    # Run LDSC
    log4<-system(paste0('/users/k1806347/brc_scratch/Software/ldsc.sh --h2 /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/LDSC/',gene_i,'.txt --ref-ld-chr /users/k1806347/brc_scratch/Data/ldsc/eur_w_ld_chr/ --w-ld-chr /users/k1806347/brc_scratch/Data/ldsc/eur_w_ld_chr/ --out /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/LDSC/',gene_i,'.h2'), intern=T)

    ########
    # Tabulate SNP-h2 results
    ########
    
    # Read SbayesR heritability result
    if(file.exists(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR/',gene_i,'.SBayesR.parRes'))){
      
      par_res_file_i<-fread(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR/',gene_i,'.SBayesR.parRes'))
      par_res_file_i<-par_res_file_i[par_res_file_i$V1 == 'hsq',2:3, with=F]
      par_res_file_i$P<-pnorm(-abs(par_res_file_i$Mean/par_res_file_i$SD))
      par_res_file_i$Gene<-gene_i
      par_res_file_i<-par_res_file_i[,c('Gene','Mean','SD','P'), with=F]
      names(par_res_file_i)<-c('gene','hsq','se','p')
      sbayesr_h2<-rbind(sbayesr_h2, par_res_file_i)
    }

    if(file.exists(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR/',gene_i,'.SBayesR_robust.parRes'))){
      
      par_res_file_i<-fread(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesR/',gene_i,'.SBayesR_robust.parRes'))
      par_res_file_i<-par_res_file_i[par_res_file_i$V1 == 'hsq',2:3, with=F]
      par_res_file_i$P<-pnorm(-abs(par_res_file_i$Mean/par_res_file_i$SD))
      par_res_file_i$Gene<-gene_i
      par_res_file_i<-par_res_file_i[,c('Gene','Mean','SD','P'), with=F]
      names(par_res_file_i)<-c('gene','hsq','se','p')
      sbayesr_robust_h2<-rbind(sbayesr_robust_h2, par_res_file_i)
    }

    # Read SbayesS heritability result
    if(file.exists(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesS/',gene_i,'.SBayesS.parRes'))){
      
      par_res_file_i<-fread(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/SBayesS/',gene_i,'.SBayesS.parRes'))
      par_res_file_i<-par_res_file_i[par_res_file_i$V1 == 'hsq',2:3, with=F]
      par_res_file_i$P<-pnorm(-abs(par_res_file_i$Mean/par_res_file_i$SD))
      par_res_file_i$Gene<-gene_i
      par_res_file_i<-par_res_file_i[,c('Gene','Mean','SD','P'), with=F]
      names(par_res_file_i)<-c('gene','hsq','se','p')
      sbayess_h2<-rbind(sbayess_h2, par_res_file_i)
    }
    
    # Read LDSC heritability result
    if(file.exists(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/LDSC/',gene_i,'.h2.log'))){
      
      par_res_file_i<-readLines(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/LDSC/',gene_i,'.h2.log'))
      par_res_file_i<-gsub('.*: ','',par_res_file_i[grepl('Total Observed scale h2',par_res_file_i)])
      hsq<-as.numeric(gsub(' .*','',par_res_file_i))
      se<-as.numeric(gsub(")",'',gsub(".*\\(",'',par_res_file_i)))
      p<-pnorm(hsq/se, lower.tail = F)
      
      ldsc_h2<-rbind(ldsc_h2, data.frame(gene=gene_i,
                                         hsq=hsq,
                                         se=se,
                                         p=p))
      
    }
  }
}

# Compare hsq estimates across methods
fusion_files<-list.files(path='/scratch/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/Whole_Blood/Whole_Blood/', pattern='.RDat')
fusion_files<-gsub('.wgt.RDat','',gsub('Whole_Blood.','',fusion_files))
fusion_files<-intersect(fusion_files, genes)

greml_h2<-NULL

for(genes_i in fusion_files){
  load(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/Whole_Blood/Whole_Blood/Whole_Blood.',genes_i,'.wgt.RDat'))
  fusion<-data.frame(wgt.matrix)
  fusion$SNP<-row.names(fusion)
  greml<-c(hsq, hsq.pv)

  greml_h2<-rbind(greml_h2, data.frame(gene=genes_i,
                                         hsq=greml[1],
                                         se=greml[2],
                                         p=greml[3]))
}

# Combine results across methods
sbayesr_h2$method<-'sbayesr'
sbayesr_robust_h2$method<-'sbayesr_robust'
sbayess_h2$method<-'sbayess'
ldsc_h2$method<-'ldsc'
greml_h2$method<-'greml'

all_h2<-do.call(rbind, list(sbayesr_h2, 
                            sbayesr_robust_h2, 
                            sbayess_h2, 
                            ldsc_h2, 
                            greml_h2))

write.table(all_h2, '/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/h2_estimates.txt', col.names=T, row.names=F, quote=F)

```

```{r, eval=T, echo=T}
library(data.table)

all_h2<-fread('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/h2_estimates.txt')

library(UpSetR)

# Remove estimates which did not converge
all_h2<-all_h2[all_h2$se != 0,]

# Merge estimates an check correlation
all_h2_full_list<-split(all_h2[,c('gene','hsq'),with=F], f = all_h2$method)

for(i in names(all_h2_full_list)){
  names(all_h2_full_list[[which(names(all_h2_full_list) == i)]])<-c('gene',i)
}

all_h2_full_dat<-Reduce(function(...) merge(..., by='gene', all=T), all_h2_full_list)

library("ggplot2")
library("GGally")   

ggpairs(all_h2_full_dat[,-1])  
# Note. LDSC estimates are very inaccurate and do not correlated well with other methods. SBayesR correlates best wit GREML, and SBayesR and SBayesR with robust setting are highly correlated. The SBayesS method correlates less well with other methods, but this doesn't mean it is inaccurate as we don't know the truth.

# Check the number of genes that are significant in each method and how they overlap across methods.

# Extract significant estimates
all_h2_sig<-all_h2[all_h2$p < 0.01,]
table(all_h2_sig$method)
# The number is similar across methods, but SBayesR has the highest.
all_h2_sig_list<-split(all_h2_sig[['gene']], f = all_h2_sig$method)

# Plot number of genes with valid and significant estimates
upset(fromList(all_h2_sig_list), nsets=10, order.by = "freq")
# Again LDSC is least concordant with other methods.

# Compare number overlapping with GREML list
all_h2_greml_sig_list<-all_h2_sig_list
for(i in names(all_h2_greml_sig_list)){
  tmp<-all_h2_greml_sig_list[[which(names(all_h2_greml_sig_list) == i)]]
  tmp<-tmp[tmp %in% all_h2_greml_sig_list[['greml']]]
  all_h2_greml_sig_list[[which(names(all_h2_greml_sig_list) == i)]]<-tmp
}

for(i in 1:length(all_h2_greml_sig_list)){
  print(names(all_h2_greml_sig_list)[i])
  print(length(all_h2_greml_sig_list[[i]]))
}

# Note. SBayesR has the largest overlap with GREML. Given this, and that SBayesR finds the most significant genes, I think we should use SBayesR without robust parameteristation to estimate heritability.

```

***

### Compare methods for generating SNP-weights

Now we have decided which genes to create SNP-weights for, we will generate SNP-weights for predicting gene expression using a range of methods. Then we will compare these SNP-weights to those derived by FUSION. Again, to avoid computational burden, start by running for the first 100 genes on chromosome 22.

```{r, eval=F, echo=T}

library(data.table)

# Make output directory
dir.create('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp')
dir.create('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SBayesR')
dir.create('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SBayesR_robust')
dir.create('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/DBSLMM')
dir.create('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/PRScs')
dir.create('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SuSiE')
dir.create('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/RDat_files')

# Read in HapMap3 SNP-list
hm3_snp<-fread('/users/k1806347/brc_scratch/Data/ldsc/w_hm3.snplist')

sbayesr_h2<-NULL

# Run across each chromosome seperately
for(chr_i in 22){
  # Read in eQTL data
  eqtl<-fread(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/Whole_Blood.allpairs.chr',chr_i,'.txt'))
  
  # Create CHR, BP, A1, A2, and Build columns
  var_id<-data.table(do.call(rbind, strsplit(eqtl$variant_id, '_')))
  names(var_id)<-c('CHR','BP','A1','A2','Build')
  var_id$CHR<-as.numeric(var_id$CHR)
  var_id$BP<-as.numeric(var_id$BP)
  
  eqtl<-cbind(eqtl, var_id)
  
  # Insert IUPAC codes for SNPs
  eqtl$IUPAC[eqtl$A1 == 'A' & eqtl$A2 =='T' | eqtl$A1 == 'T' & eqtl$A2 =='A']<-'W'
  eqtl$IUPAC[eqtl$A1 == 'C' & eqtl$A2 =='G' | eqtl$A1 == 'G' & eqtl$A2 =='C']<-'S'
  eqtl$IUPAC[eqtl$A1 == 'A' & eqtl$A2 =='G' | eqtl$A1 == 'G' & eqtl$A2 =='A']<-'R'
  eqtl$IUPAC[eqtl$A1 == 'C' & eqtl$A2 =='T' | eqtl$A1 == 'T' & eqtl$A2 =='C']<-'Y'
  eqtl$IUPAC[eqtl$A1 == 'G' & eqtl$A2 =='T' | eqtl$A1 == 'T' & eqtl$A2 =='G']<-'K'
  eqtl$IUPAC[eqtl$A1 == 'A' & eqtl$A2 =='C' | eqtl$A1 == 'C' & eqtl$A2 =='A']<-'M'
  
  # Remove INDELS
  eqtl<-eqtl[!is.na(eqtl$IUPAC),]
  
  # Calculate N for each association
  eqtl$N<-round((eqtl$ma_count/eqtl$maf)/2,1)
  
  # Check the build
  table(eqtl$Build) 
  # They are all b37
  
  ###
  # Insert RSIDs
  ###
  
  # Read in 1KG reference SNP data
  ref_bim_i<-fread(paste0('/mnt/lustre/groups/biomarkers-brc-mh/Reference_data/1KG_Phase3/PLINK/all_phase3.chr',chr_i,'.bim'))
  names(ref_bim_i)<-c('CHR','SNP','POS','BP','A1','A2')
  
  # Extract hm3 SNPs
  ref_bim_i<-ref_bim_i[ref_bim_i$SNP %in% hm3_snp$SNP,]
  
  # Insert IUPAC codes
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'A' & ref_bim_i$A2 =='T' | ref_bim_i$A1 == 'T' & ref_bim_i$A2 =='A']<-'W'
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'C' & ref_bim_i$A2 =='G' | ref_bim_i$A1 == 'G' & ref_bim_i$A2 =='C']<-'S'
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'A' & ref_bim_i$A2 =='G' | ref_bim_i$A1 == 'G' & ref_bim_i$A2 =='A']<-'R'
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'C' & ref_bim_i$A2 =='T' | ref_bim_i$A1 == 'T' & ref_bim_i$A2 =='C']<-'Y'
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'G' & ref_bim_i$A2 =='T' | ref_bim_i$A1 == 'T' & ref_bim_i$A2 =='G']<-'K'
  ref_bim_i$IUPAC[ref_bim_i$A1 == 'A' & ref_bim_i$A2 =='C' | ref_bim_i$A1 == 'C' & ref_bim_i$A2 =='A']<-'M'

  # Merge target and reference based on position
  ref_target<-merge(eqtl, ref_bim_i, by=c('CHR','BP'))
  
  # I have used the GTEx reference data as well, and the same number is returned. 
  # Note: The A1 allele is the ref allele.
  # Small number of SNPs remaining highlights the limitation of using HapMap3 SNPs only.
  
  # Identify SNPs for which alleles need to be flipped
  flip_tmp<-ref_target[(ref_target$IUPAC.x == 'R' & ref_target$IUPAC.y == 'Y' | 
                          ref_target$IUPAC.x == 'Y' & ref_target$IUPAC.y == 'R' | 
                          ref_target$IUPAC.x == 'K' & ref_target$IUPAC.y == 'M' |
                          ref_target$IUPAC.x == 'M' & ref_target$IUPAC.y == 'K'),]
  
  # Idenitfy SNPs which match the reference alleles
  incl<-ref_target[ ref_target$IUPAC.x == 'R' & ref_target$IUPAC.y == 'R' | 
                      ref_target$IUPAC.x == 'Y' & ref_target$IUPAC.y == 'Y' | 
                      ref_target$IUPAC.x == 'K' & ref_target$IUPAC.y == 'K' |
                      ref_target$IUPAC.x == 'M' & ref_target$IUPAC.y == 'M' ]
  
  # If a SNP that needs to be flipped has a duplicate that is on the correct strand, remove it.
  flip<-flip_tmp[!(flip_tmp$SNP %in% incl$SNP)]
  
  # Flip alleles where necessary
  flip_tmp$A1_flipped<-flip_tmp$A1.x
  flip_tmp$A1_flipped[flip_tmp$A1.x == 'A']<-'T'
  flip_tmp$A1_flipped[flip_tmp$A1.x == 'T']<-'C'
  flip_tmp$A1_flipped[flip_tmp$A1.x == 'G']<-'C'
  flip_tmp$A1_flipped[flip_tmp$A1.x == 'C']<-'G'
  flip_tmp$A1.x<-flip_tmp$A1_flipped
  flip_tmp$A1_flipped<-NULL
  
  flip_tmp$A2_flipped<-flip_tmp$A2.x
  flip_tmp$A2_flipped[flip_tmp$A2.x == 'A']<-'T'
  flip_tmp$A2_flipped[flip_tmp$A2.x == 'T']<-'C'
  flip_tmp$A2_flipped[flip_tmp$A2.x == 'G']<-'C'
  flip_tmp$A2_flipped[flip_tmp$A2.x == 'C']<-'G'
  flip_tmp$A2.x<-flip_tmp$A2_flipped
  flip_tmp$A2_flipped<-NULL
  
  # Recombine and format
  eqtl_harm<-rbind(flip_tmp, incl)
  eqtl_harm<-eqtl_harm[,c('CHR','SNP','BP','A1.x','A2.x',"gene_id","variant_id","tss_distance","ma_samples","ma_count","maf","pval_nominal","slope","slope_se","N"), with=F]
  names(eqtl_harm)[names(eqtl_harm) == 'A1.x']<-'A1'
  names(eqtl_harm)[names(eqtl_harm) == 'A2.x']<-'A2'

  # Remove variants with MAF < 1%
  eqtl_harm<-eqtl_harm[eqtl_harm$maf >= 0.01,]
  
  eqtl<-eqtl_harm
  rm(eqtl_harm)
  
  # Identify unique genes 
  genes<-unique(eqtl$gene_id)
  genes<-genes[1:100]

  # Run for each gene seperately
  for(gene_i in genes){
    print(which(genes == gene_i))

    eqtl_gene_i<-eqtl[eqtl$gene_id == gene_i,]
    
    # Filter SNPs to those with N > 80% of max(N)
    eqtl_gene_i<-eqtl_gene_i[eqtl_gene_i$N >= 0.8*max(eqtl_gene_i$N),]
    
    #######
    # SBayesR
    #######
    
    # Format for SBayesR
    eqtl_gene_i_sbayesr<-eqtl_gene_i[,c('SNP','A1','A2','maf','slope','slope_se','pval_nominal','N'), with=F]
    names(eqtl_gene_i_sbayesr)<-c('SNP','A1','A2','freq','b','se','p','N')
    
    # write in SBayesR format
    fwrite(eqtl_gene_i_sbayesr, paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SBayesR/',gene_i,'.txt'), sep=' ', na = "NA", quote=F)
    
    # Run SBayesR
    log<-system(paste0('/users/k1806347/brc_scratch/Software/gctb_2.03beta_Linux/gctb_203.sh --sbayes R --ldm /users/k1806347/brc_scratch/Data/1KG/Phase3/LD_matrix/GCTB_shared/ukbEURu_hm3_shrunk_sparse/ukbEURu_hm3_v3_50k_chr',chr_i,'.ldm.sparse --pi 0.95,0.02,0.02,0.01 --gamma 0.0,0.01,0.1,1 --gwas-summary /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SBayesR/',gene_i,'.txt --chain-length 10000 --exclude-mhc --burn-in 2000 --out-freq 1000 --out /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SBayesR/',gene_i,'.SBayesR'), intern=T)
    
    # Read SbayesR heritability result
    if(file.exists(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SBayesR/',gene_i,'.SBayesR.parRes'))){
      
      par_res_file_i<-fread(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SBayesR/',gene_i,'.SBayesR.parRes'))
      par_res_file_i<-par_res_file_i[par_res_file_i$V1 == 'hsq',2:3, with=F]
      par_res_file_i$P<-pnorm(-abs(par_res_file_i$Mean/par_res_file_i$SD))
      par_res_file_i$Gene<-gene_i
      par_res_file_i<-par_res_file_i[,c('Gene','Mean','SD','P'), with=F]
      names(par_res_file_i)<-c('gene','hsq','se','p')
      sbayesr_h2<-rbind(sbayesr_h2, par_res_file_i)
      
      # If SNP-h2 p < 0.01, generate SNP-weights
      if(par_res_file_i$p < 0.01){
        # Flip the effect of each method to match eqtl sumstats
        ref_tmp<-eqtl_gene_i[, c('SNP','A1','A2'), with=F]
        
        #############
        # SBayesR
        #############
        # SBayesR has already been run, so just read in the SNP-weights
        sbayesr_score<-fread(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SBayesR/',gene_i,'.SBayesR.snpRes'))
        sbayesr_score<-sbayesr_score[,c('Name','A1','A2','A1Effect'), with=F]
        names(sbayesr_score)<-c('SNP','A1','A2','BETA')
        
        # Flip effects so allele match eQTL sumstats
        sbayesr_score<-merge(ref_tmp, sbayesr_score, by='SNP', all=T)
        sbayesr_score$BETA[which(sbayesr_score$A1.x == sbayesr_score$A2.y)]<--sbayesr_score$BETA[which(sbayesr_score$A1.x == sbayesr_score$A2.y)]
        sbayesr_score<-sbayesr_score[,c('SNP','A1.x','A2.x','BETA'), with=F]
        names(sbayesr_score)<-c('SNP','A1','A2','BETA')
        
        #############
        # SBayesR robust
        #############
        # SBayesR format sumstats are already available
        # So just run SBayesR with robust setting
        
        log<-system(paste0('/users/k1806347/brc_scratch/Software/gctb_2.03beta_Linux/gctb_203.sh --sbayes R --ldm /users/k1806347/brc_scratch/Data/1KG/Phase3/LD_matrix/GCTB_shared/ukbEURu_hm3_shrunk_sparse/ukbEURu_hm3_v3_50k_chr',chr_i,'.ldm.sparse --pi 0.95,0.02,0.02,0.01 --gamma 0.0,0.01,0.1,1 --gwas-summary /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SBayesR/',gene_i,'.txt --robust --chain-length 10000 --exclude-mhc --burn-in 2000 --out-freq 1000 --out /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SBayesR_robust/',gene_i,'.SBayesR'), intern=T)

        # Read in the results
        sbayesr_robust_score<-fread(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SBayesR_robust/',gene_i,'.SBayesR.snpRes'))
        sbayesr_robust_score<-sbayesr_robust_score[,c('Name','A1','A2','A1Effect'), with=F]
        names(sbayesr_robust_score)<-c('SNP','A1','A2','BETA')
        
        # Flip effects so allele match eQTL sumstats
        sbayesr_robust_score<-merge(ref_tmp, sbayesr_robust_score, by='SNP', all=T)
        sbayesr_robust_score$BETA[which(sbayesr_robust_score$A1.x == sbayesr_robust_score$A2.y)]<--sbayesr_robust_score$BETA[which(sbayesr_robust_score$A1.x == sbayesr_robust_score$A2.y)]
        sbayesr_robust_score<-sbayesr_robust_score[,c('SNP','A1.x','A2.x','BETA'), with=F]
        names(sbayesr_robust_score)<-c('SNP','A1','A2','BETA')

        #############
        # DBSLMM
        #############
        
        # Convert to GEMMA format
        eqtl_gene_i_dbslmm<-eqtl_gene_i
        eqtl_gene_i_dbslmm$N_MISS<-max(eqtl_gene_i_dbslmm$N)-eqtl_gene_i_dbslmm$N
        eqtl_gene_i_dbslmm<-eqtl_gene_i_dbslmm[,c('CHR','SNP','BP','N_MISS','N','A1','A2','maf','slope','slope_se','pval_nominal'),with=F]
          names(eqtl_gene_i_dbslmm)<-c('chr','rs','ps','n_mis','n_obs','allele1','allele0','af','beta','se','p_wald')
        
        # Match allele1 and 0 with A1 and 2 in reference (DBSLMM calls this allele discrepancy)
        ref_bim_subset<-fread(paste0('/scratch/groups/biomarkers-brc-mh/Reference_data/1KG_Phase3/PLINK/EUR/EUR_phase3.MAF_001.chr',chr_i,'.bim'))
      
        GWAS_match<-merge(eqtl_gene_i_dbslmm, ref_bim_subset[,c('V2','V5','V6'),with=F], by.x=c('rs','allele1','allele0'), by.y=c('V2','V5','V6'))
        GWAS_switch<-merge(eqtl_gene_i_dbslmm, ref_bim_subset[,c('V2','V5','V6'),with=F], by.x=c('rs','allele1','allele0'), by.y=c('V2','V6','V5'))
        GWAS_switch$allele_tmp<-GWAS_switch$allele0
        GWAS_switch$allele0<-GWAS_switch$allele1
        GWAS_switch$allele1<-GWAS_switch$allele_tmp
        GWAS_switch$allele_tmp<-NULL
        GWAS_switch$beta<--GWAS_switch$beta
        GWAS_switch$af<-1-GWAS_switch$af
        GWAS<-rbind(GWAS_match, GWAS_switch)
        
        GWAS<-GWAS[order(GWAS$chr, GWAS$ps),]
        GWAS<-GWAS[,c('chr','rs','ps','n_mis','n_obs','allele1','allele0','af','beta','se','p_wald'),with=F]
        GWAS_N<-mean(GWAS$n_obs)
        nsnp<-nrow(GWAS)
        
        # Write out formatted sumstats
        fwrite(GWAS, paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/DBSLMM/',gene_i,'.DBSLMM'), sep='\t', col.names=F)

        # Run dbslmm
        system('chmod 777 /users/k1806347/brc_scratch/Software/DBSLMM/software/dbslmm')
        system(paste0('/users/k1806347/brc_scratch/Software/Rscript.sh /users/k1806347/brc_scratch/Software/DBSLMM/software/DBSLMM.R --plink /users/k1806347/brc_scratch/Software/plink1.9.sh --block /users/k1806347/brc_scratch/Data/LDetect/EUR/fourier_ls-chr',chr_i,'.bed --dbslmm /users/k1806347/brc_scratch/Software/DBSLMM/software/dbslmm --h2 ',par_res_file_i$hsq,' --ref /scratch/groups/biomarkers-brc-mh/Reference_data/1KG_Phase3/PLINK/EUR/EUR_phase3.MAF_001.chr',chr_i,' --summary /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/DBSLMM/',gene_i,'.DBSLMM --n ',round(GWAS_N,0),' --nsnp ',nsnp,' --outPath /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/DBSLMM/ --thread 1'))

        # Read in the results
        dbslmm_score<-fread(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/DBSLMM/',gsub('\\..*','',gene_i),'.dbslmm.txt'))
        dbslmm_score<-dbslmm_score[,c(1,2,4), with=T]
        names(dbslmm_score)<-c('SNP','A1','BETA')

        # Flip effects so allele match eQTL sumstats
        dbslmm_score<-merge(ref_tmp, dbslmm_score, by='SNP', all=T)
        dbslmm_score$BETA[which(dbslmm_score$A1.x != dbslmm_score$A1.y)]<--dbslmm_score$BETA[which(dbslmm_score$A1.x != dbslmm_score$A1.y)]
        dbslmm_score<-dbslmm_score[,c('SNP','A1.x','A2','BETA'), with=F]
        names(dbslmm_score)<-c('SNP','A1','A2','BETA')

        ##############
        # PRScs
        ##############
        
        # Format for PRScs
        eqtl_gene_i_prscs<-eqtl_gene_i[,c('SNP','A1','A2','slope','pval_nominal'), with=F]
        names(eqtl_gene_i_prscs)<-c('SNP','A1','A2','BETA','P')
        
        # write in PRScs format
        fwrite(eqtl_gene_i_prscs, paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/PRScs/',gene_i,'.txt'), sep=' ', na = "NA", quote=F)

        system(paste0('/users/k1806347/brc_scratch/Software/PRScs.sh --ref_dir=/users/k1806347/brc_scratch/Software/PRScs/ldblk_1kg_eur --bim_prefix=/scratch/groups/biomarkers-brc-mh/Reference_data/1KG_Phase3/PLINK/EUR/EUR_phase3.MAF_001.chr',chr_i,' --sst_file=/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/PRScs/',gene_i,'.txt --n_gwas=',round(GWAS_N,0),' --out_dir=/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/PRScs/',gene_i,' --chrom=',chr_i))

        # Read in the results
        prscs_score<-fread(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/PRScs/',gene_i,'_pst_eff_a1_b0.5_phiauto_chr22.txt'))
        prscs_score<-prscs_score[,c('V2','V4','V6'), with=F]
        names(prscs_score)<-c('SNP','A1','BETA')

        # Flip effects so allele match eQTL sumstats
        prscs_score<-merge(ref_tmp, prscs_score, by='SNP', all=T)
        prscs_score$BETA[which(prscs_score$A1.x != prscs_score$A1.y)]<--prscs_score$BETA[which(prscs_score$A1.x != prscs_score$A1.y)]
        prscs_score<-prscs_score[,c('SNP','A1.x','A2','BETA'), with=F]
        names(prscs_score)<-c('SNP','A1','A2','BETA')
        
        # Note PRScs takes a lot longer than other methods.
        
        ################
        # SuSiE finemapping
        ################
        
        # Read LD estimates for eQTL sumstats
        write.table(eqtl_gene_i$SNP, paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SuSiE/',gene_i,'_snps.txt'), col.names=F, row.names=F, quote=F)
        system(paste0('/users/k1806347/brc_scratch/Software/plink1.9.sh --bfile /scratch/groups/biomarkers-brc-mh/Reference_data/1KG_Phase3/PLINK/EUR/EUR_phase3.MAF_001.chr',chr_i,' --extract /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SuSiE/',gene_i,'_snps.txt --r square --out /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SuSiE/',gene_i))

        ld<-as.matrix(fread(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/SuSiE/',gene_i,'.ld')))

        library(susieR)
        
        tryCatch(fitted_rss <- susie_rss(eqtl_gene_i$slope/eqtl_gene_i$slope_se, ld, L = 10), error = function(e){skip_to_next <<- TRUE})
        
        if(skip_to_next == TRUE){
          susie_score<-data.table(SNP=eqtl_gene_i$SNP,
                                  A1=eqtl_gene_i$A1,
                                  BETA=0)
        } else {
          susie_score<-data.table(SNP=eqtl_gene_i$SNP,
                        A1=eqtl_gene_i$A1,
                        BETA=fitted_rss$pip)
        }
        
        # Flip effects so allele match eQTL sumstats
        susie_score<-merge(ref_tmp, susie_score, by='SNP', all=T)
        susie_score$BETA[which(susie_score$A1.x != susie_score$A1.y)]<--susie_score$BETA[which(susie_score$A1.x != susie_score$A1.y)]
        susie_score<-susie_score[,c('SNP','A1.x','A2','BETA'), with=F]
        names(susie_score)<-c('SNP','A1','A2','BETA')
        
        # Create RDat file for FUSION
        cv.performance<-as.matrix(data.frame(sbayesr=c(NA,NA),
                                             sbayesr_robust=c(NA,NA),
                                             dbslmm=c(NA,NA),
                                             prscs=c(NA,NA),
                                             susie=c(NA,NA),
                                             top1=c(NA,NA), 
                                             row.names=c('rsq','pval')))
        
        # Sort eQTL data into the same order as other score files
        eqtl_gene_i<-eqtl_gene_i[match(sbayesr_score$SNP, eqtl_gene_i$SNP),]
        
        hsq<-c(par_res_file_i$hsq, par_res_file_i$se)
        hsq.pv<-par_res_file_i$p
        N.tot<-max(eqtl_gene_i$N)
        eqtl_gene_i$POS<-0
        snps<-eqtl_gene_i[,c('CHR','SNP','POS','BP','A1','A2')]
        names(snps)<-paste0('V',1:length(names(snps)))
        
        wgt.matrix<-as.matrix(data.frame(sbayesr=sbayesr_score$BETA,
                                         sbayesr_robust=sbayesr_robust_score$BETA,
                                         dbslmm=dbslmm_score$BETA,
                                         prscs=prscs_score$BETA,
                                         susie=susie_score$BETA,
                                         top1=eqtl_gene_i$slope,
                                         row.names=snps$V2))
        
        save(cv.performance, 
             hsq,
             hsq.pv,
             N.tot,
             snps,
             wgt.matrix,
             file = paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/RDat_files/',gene_i,'.RDat'))
      }
    }
  }
}

```

```{r, echo=T, eval=F}

library(data.table)

fusion_pos<-fread('/scratch/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/Whole_Blood/Whole_Blood.pos')
fusion_pos<-fusion_pos[fusion_pos$CHR == 22,]

dim(fusion_pos) # 234

eqtl_derived_files<-list.files(path='/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/RDat_files/', pattern='.RDat')
fusion_files<-list.files(path='/scratch/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/Whole_Blood/Whole_Blood/', pattern='.RDat')

length(eqtl_derived_files) # 159
length(fusion_files) # 6006

eqtl_derived_files<-gsub('.RDat','',eqtl_derived_files)
fusion_files<-gsub('.wgt.RDat','',gsub('Whole_Blood.','',fusion_files))

genes<-intersect(eqtl_derived_files, fusion_files)
length(genes) # 33

nsnp_res<-NULL
hsq_res<-NULL
cor_res<-list()

for(genes_i in genes){
  load(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/RDat_files/',genes_i,'.RDat'))
  eqtl_derived<-data.frame(wgt.matrix)
  if(sum(eqtl_derived$susie) == 0){
    eqtl_derived$susie<-NA
  }
  eqtl_derived$SNP<-row.names(eqtl_derived)
  sbayesr<-c(hsq, hsq.pv)
  
  load(paste0('/scratch/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/Whole_Blood/Whole_Blood/Whole_Blood.',genes_i,'.wgt.RDat'))
  fusion<-data.frame(wgt.matrix)
  fusion$SNP<-row.names(fusion)
  greml<-c(hsq, hsq.pv)
  
  nsnp_res<-rbind(nsnp_res, data.frame(gene_id=genes_i,
                                       nsnp_eqtl_derived=nrow(eqtl_derived),
                                       nsnp_fusion=nrow(fusion)))
  
  hsq_res<-rbind(hsq_res, data.frame(gene_id=genes_i,
                                     sbayesr_h2=sbayesr[1],
                                     sbayesr_se=sbayesr[2],
                                     sbayesr_p=sbayesr[3],
                                     greml_h2=greml[1],
                                     greml_se=greml[2],
                                     greml_p=greml[3]))

  both<-merge(eqtl_derived, fusion, by='SNP')
  
  if(nrow(both) > 0){
    cor_res[[genes_i]]<-cor(both[,-1], use='p')
  }
}

# Compare h2
cor(hsq_res$sbayesr_h2, hsq_res$greml_h2) # 0.7500682

library(ggplot2)
bitmap('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/h2_comp.png', units='px', width=1000, height=1000, res=300)
ggplot(hsq_res, aes(x=sbayesr_h2, y=greml_h2)) +
  geom_abline(intercept =0 , slope = 1, colour='red') +
  geom_point() +
  coord_fixed() +
  xlim(0,1) +
  ylim(0,1)
dev.off()

# Compare nsnp
cor(nsnp_res$nsnp_eqtl_derived, nsnp_res$nsnp_fusion) # 0.4849533

library(ggplot2)
bitmap('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/nsnp_comp.png', units='px', width=1000, height=1000, res=300)
ggplot(nsnp_res, aes(x=nsnp_eqtl_derived, y=nsnp_fusion)) +
  geom_abline(intercept =0 , slope = 1, colour='red') +
  geom_point() +
  coord_fixed()
dev.off()

# Check SNP-weight correlations
cor_res_melt<-melt(cor_res)
cor_res_melt$Var1<-gsub('top1.x','top1.GTEx',cor_res_melt$Var1)
cor_res_melt$Var2<-gsub('top1.x','top1.GTEx',cor_res_melt$Var2)
cor_res_melt$Var1<-gsub('top1.y','top1.FUSION',cor_res_melt$Var1)
cor_res_melt$Var2<-gsub('top1.y','top1.FUSION',cor_res_melt$Var2)
cor_res_melt$test<-paste0(cor_res_melt$Var1,'_',cor_res_melt$Var2)

cor_res_average<-NULL
for(i in unique(cor_res_melt$test)){
  cor_res_average<-rbind(cor_res_average, data.frame(test=i,
                                                     Mean=mean(cor_res_melt$value[cor_res_melt$test == i])))
}

bitmap('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/GTEx_v7/snp_weight_comp/cor_plot.png', units='px', width=3000, height=3000, res=300)
ggplot(cor_res_melt, aes(x=paste0(Var1,' vs. ',Var2), y=value)) +
  geom_boxplot() +
  coord_flip()
dev.off()

# SuSiE results are only available for 1 gene. LD mismatch seems too severe. Am I doing somthing wrong or is it required for an in sample LD reference to be used?

```

The correlation between eQTL derived TWAS weights and FUSION derived TWAS weights is similar to correlation between different FUSION models. The correlation between eQTL derived and FUSION weights is of interest, but not a good evaluation metric of methods, We should compare the correlation between predicted and observed expression, when using eQTL derived TWAS weights or largest FUSION TWAS weights. The low correlation between topSNP results from FUSION and GTEx indicates comparison of FUSION and eQTL susmtat derived weights are not very informative.

***

## Comparing predicted and observed gene expression in GTEx v8

Here we will predict gene expression levels in the GTEx v8 sample and then test the correlation with observered expression levels. Initially, we will use FUSION SNP-weights derived using the YFS whole blood sample, and eQTL data from eQTLGen whole blood meta-analysis (excl GTEx).

I do not have access to individual level data from GTeX v8, so this project is being carried out in collaboration with Zac Gerring and Eske Derks. Here, I will prepare the SNP-weights, the code to predict gene expression levels, and the code to compare predicted and observed expression levels. As an example target sample, I will use the EUR subset of the 1KG Phase 3 reference.

I have already written a script to predicted expression levels from TWAS weights. However, the script currently only uses the weights model with the best variance explained out of sample. To evaluate a range of methods, we will need predicted expression levels from all models.

***

### Generate SNP-weights from eQTLGen

```{bash, eval=F, echo=T}
# Download the cis-eQTL data exluding GTEx
mkdir /users/k1806347/brc_scratch/Data/eQTLGen/excl_GTEx
cd /users/k1806347/brc_scratch/Data/eQTLGen/excl_GTEx
wget https://molgenis26.gcc.rug.nl/downloads/eqtlgen/replication/2019-12-11-cisEQTL-ReplicationOfTopEffectsInGTEx_WoGTEx_TwoCohortFix.txt.gz

# I have realised that these replication results only inlcude the top SNP for each gene which we cannot use for deriving TWAS weights. I have emailed the eQTLGen PIs to ask for full leave-one-out results.

```

***

### Generate SNP-weights from FUSION released YFS eQTL data

FUSION only provide the N of the sample and the Z of each SNP for each gene. We can convert this to the required information using reference MAF, though using these approximations is not ideal. This may not be the best comparison since they are finish and there may therefore be MAF and LD mismatch with target sample. Though I assume FUSION developers restricted the analysis to those os EUR ancestry.

***

#### Format eQTL sumstats

```{r, echo=T, eval=F}
library(data.table)

# Read in the pos file
pos<-fread('/mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/YFS.BLOOD.RNAARR/YFS.BLOOD.RNAARR.pos')

# Read in EUR MAF
frq<-NULL
for(i in 1:22){
  frq<-rbind(frq, fread(paste0('/mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/LDREF/1000G.EUR.',i,'.frq')))
}

for(i in 1:nrow(pos)){
  print(i)
  
  load(paste0('/mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/YFS.BLOOD.RNAARR/',pos$WGT[i]))
  tmp<-data.table(cbind(snps, wgt.matrix[,which(colnames(wgt.matrix) == 'top1')]))
  names(tmp)<-c('CHR','SNP','POS','BP','A1','A2','Z')
  tmp$N<-pos$N[i]
  
  # Insert EUR MAF
  tmp_match<-merge(tmp, frq[,c('SNP','A1','A2','MAF'),with=F], by=c('SNP','A1','A2'))
  tmp_flip<-merge(tmp, frq[,c('SNP','A1','A2','MAF'),with=F], by.x=c('SNP','A1','A2'), by.y=c('SNP','A2','A1'))
  tmp_flip$MAF<-1-tmp_flip$MAF
  tmp<-rbind(tmp_match, tmp_flip)
  
  # Approximate BETA, SE, and P
  tmp$P<-2*pnorm(-abs(tmp$Z))
  tmp$BETA<-tmp$Z/sqrt((2*tmp$MAF)*(1-tmp$MAF)*(tmp$N+sqrt(abs(tmp$Z))))
  tmp$SE<-abs(tmp$BETA)/abs(tmp$Z)
  
  tmp$GENE<-pos$ID[i]
    
  tmp<-tmp[,c('GENE','CHR','SNP','BP','A1','A2','BETA','SE','Z','P','MAF','N'),with=F]
  
  if(i == 1){
    write.table(tmp, '/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS_eQTL_from_fusion.txt', col.names=T, row.names=F, quote=F)
  } else {
    write.table(tmp, '/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS_eQTL_from_fusion.txt', col.names=F, row.names=F, quote=F, append=T)
  }
}

tmp<-fread('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS_eQTL_from_fusion.txt')
names(tmp)[names(tmp) == 'MAF']<-'FREQ'
write.table(tmp, '/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS_eQTL_from_fusion.txt', col.names=T, row.names=F, quote=F)

# Make this the input format for eQTL_to_TWAS script.
# Incorporate DENTIST QC of sumstats?

```

***

#### Create TWAS weights

```{bash, eval=F, echo=T}

# Script modified to run only first 100 genes
sbatch -p brc,shared --mem 10G -n 1 /users/k1806347/brc_scratch/Software/Rscript.sh /users/k1806347/brc_scratch/Software/MyGit/eQTL_to_TWAS/compute_weights.R \
	--extract /users/k1806347/brc_scratch/Data/ldsc/w_hm3.snplist \
	--sumstats /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS_eQTL_from_fusion.txt \
	--gctb /users/k1806347/brc_scratch/Software/gctb_2.03beta_Linux/gctb_203.sh \
	--gctb_ref /users/k1806347/brc_scratch/Data/1KG/Phase3/LD_matrix/GCTB_shared/ukbEURu_hm3_shrunk_sparse/ukbEURu_hm3_v3_50k_chr \
	--plink_ref_chr /users/k1806347/brc_scratch/Data/1KG/Phase3/1KGPhase3.w_hm3.chr \
	--ld_blocks /users/k1806347/brc_scratch/Data/LDetect/EUR \
  --rscript /users/k1806347/brc_scratch/Software/Rscript.sh \
  --dbslmm /users/k1806347/brc_scratch/Software/DBSLMM/software \
  --plink /users/k1806347/brc_scratch/Software/plink1.9.sh \
  --PRScs_path /users/k1806347/brc_scratch/Software/PRScs.sh \
  --PRScs_ref_path /users/k1806347/brc_scratch/Software/PRScs/ldblk_1kg_eur \
	--output /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/twas_weights

```

```{bash, eval=F, echo=T}
# Rename RDat folder to match FUSION format weights
mv /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/RDat_files /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS.eQTL

```

```{R, echo=T, eval=F}
# Create .pos file
library(data.table)
pos<-fread('/mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/YFS.BLOOD.RNAARR/YFS.BLOOD.RNAARR.pos')

# Read in list of RDat files
rdat_list<-list.files(path='/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS.eQTL', pattern='.RDat')

# Subset pos file to those with RDat file
pos<-pos[pos$ID %in% gsub('.RDat','',rdat_list),]

pos$PANEL<-'YFS.eQTL'
pos$WGT<-paste0(pos$ID,'.RDat')

write.table(pos, '/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS.eQTL.pos', col.names=T, row.names=F, quote=F)

```


***

### Predict expression in 1KG

***

#### FUSION YFS models

```{bash, eval=F, echo=T}
sbatch -p brc,shared --mem 10G -n 1 /users/k1806347/brc_scratch/Software/Rscript.sh /users/k1806347/brc_scratch/Software/MyGit/Predicting-TWAS-features/FeaturePred.V2.0.R \
	--PLINK_prefix_chr /mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/LDREF/1000G.EUR. \
	--weights /mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/YFS.BLOOD.RNAARR/YFS.BLOOD.RNAARR.pos \
	--weights_dir /mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/YFS.BLOOD.RNAARR \
	--ref_ld_chr /mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/LDREF/1000G.EUR. \
	--plink /users/k1806347/brc_scratch/Software/plink1.9.sh \
	--n_cores 1 \
	--save_score T \
	--save_ref_expr T \
	--memory 10000 \
	--all_mod T \
	--pigz /users/k1806347/brc_scratch/Software/pigz \
	--ref_maf /mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/LDREF/1000G.EUR. \
	--output /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/1kg_pred_exp/YFS.BLOOD.RNAARR

```

***

#### YFS eQTL-based models

```{bash, eval=F, echo=T}
sbatch -p brc,shared --mem 10G -n 1 /users/k1806347/brc_scratch/Software/Rscript.sh /users/k1806347/brc_scratch/Software/MyGit/Predicting-TWAS-features/FeaturePred.V2.0.R \
	--PLINK_prefix_chr /mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/LDREF/1000G.EUR. \
	--weights /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS.eQTL.pos \
	--weights_dir /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS \
	--ref_ld_chr /mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/LDREF/1000G.EUR. \
	--plink /users/k1806347/brc_scratch/Software/plink1.9.sh \
	--n_cores 1 \
	--save_score T \
	--save_ref_expr T \
	--memory 10000 \
	--all_mod T \
	--pigz /users/k1806347/brc_scratch/Software/pigz \
	--ref_maf /mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/LDREF/1000G.EUR. \
	--output /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/1kg_pred_exp/YFS.eQTL

```

***

### Check correlation between predicted expression for YFS models

```{R, eval=F, echo=T}

library(data.table)
fusion<-fread('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/1kg_pred_exp/YFS.BLOOD.RNAARR/FeaturePredictions_YFS.BLOOD.RNAARR.txt.gz')
eqtl<-fread('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/1kg_pred_exp/YFS.eQTL/FeaturePredictions_YFS.eQTL.txt.gz')

genes<-unique(gsub('\\..*','',gsub('YFS.eQTL.','',names(eqtl)[-1:-2])))

fusion<-fusion[,grepl(paste0('FID|IID|',paste(genes,collapse='|')),names(fusion)), with=F]
both<-merge(fusion, eqtl, by=c('FID','IID'))

cor_res<-list()
for(i in genes){
  tmp<-both[,grepl(paste0('\\.',i,'\\.'), names(both)), with=F]
  names(tmp)<-gsub('YFS.eQTL.RDat.','eqtl.',gsub('YFS.BLOOD.RNAARR.YFS.','twas.',gsub(paste0(i,'.'),'', names(tmp))))
  cor_res[[i]]<-cor(tmp, use='p')
}

# Check SNP-weight correlations
cor_res_melt<-melt(cor_res)
cor_res_melt$test<-paste0(cor_res_melt$Var1,'_',cor_res_melt$Var2)

cor_res_average<-NULL
for(i in unique(cor_res_melt$test)){
  cor_res_average<-rbind(cor_res_average, data.frame(test=i,
                                                     Mean=mean(cor_res_melt$value[cor_res_melt$test == i])))
}

library(ggplot2)
bitmap('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/1kg_pred_exp/YFS_cor_plot.png', units='px', width=3000, height=3000, res=300)
ggplot(cor_res_melt, aes(x=paste0(Var1,' vs. ',Var2), y=value)) +
  geom_boxplot() +
  coord_flip()
dev.off()

# Looks as expected, except eQTL.top1 model doesn't correlate well with fusion.top1 model. In many instances the correlation is 1, but there are many low correlation. This seems to be due to changing the top1 Z score in fusion files to a BETA estimated based on N and MAF. In some instances this changes the top SNP to a SNP that is uncorrelated with the top SNP in the fusion files. Perhaps I should select the top SNP based on the p-value, rather than the BETA. I think selecting based on p-value may be more concordant with standard pT+clump PRS methodology, and will avoid SNPs with small N and large SE being selected. Thougb since we are filtering by N, this shouldn;t be a problem, and usin the SNP with the largest BETA may be more appropriate. However, the correlation between top1 and other models is lower when using the largest abs(BETA), suggesting using the most significant SNP may be more appropriate. This should be discussed with Sasha.

load('/mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/YFS.BLOOD.RNAARR/YFS.BLOOD.RNAARR/YFS.TRNAU1AP.wgt.RDat')
fusion_wgt<-data.frame(wgt.matrix)
load('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS.eQTL/TRNAU1AP.RDat')
eqtl_wgt<-data.frame(wgt.matrix)

head(fusion_wgt[rev(order(abs(fusion_wgt$top1))),])
head(eqtl_wgt[rev(order(abs(eqtl_wgt$top1))),])

```

***

### Predict expression in GTEx v8

***

This will be run by Zac. I will send him the score files and reference expression, and fusion LD reference data. I don't know the file paths so I have just put file names.

***

#### Compress files for Zac

```{bash, eval=F, echo=T}
mkdir -p /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/for_zac
cd /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/for_zac

# Copy FUSION reference with allele frequency files
cp -r /mnt/lustre/groups/biomarkers-brc-mh/TWAS_resource/FUSION/LDREF ./

# Copy FUSION .pos, .Rdat files, score files, and reference expression
cd /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/for_zac
mkdir FUSION.YFS
cd FUSION.YFS

cp /scratch/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/YFS.BLOOD.RNAARR/YFS.BLOOD.RNAARR.pos ./
cp -r /scratch/groups/biomarkers-brc-mh/TWAS_resource/FUSION/SNP-weights/YFS.BLOOD.RNAARR/YFS.BLOOD.RNAARR ./

# Copy eQTL-based .pos, .Rdat files, score files, and reference expression
cd /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/for_zac
mkdir eQTL.YFS
cd eQTL.YFS

cp /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS.eQTL.pos ./
cp -r /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/YFS/YFS.eQTL ./

# Copy pigz binary
cd /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/for_zac
cp /users/k1806347/brc_scratch/Software/pigz ./

# Copy plink binary
cd /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/for_zac
cp /users/k1806347/brc_scratch/Software/plink1.9/plink ./

# Copy FeaturePred.V2.0.R
cd /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/for_zac
cp /users/k1806347/brc_scratch/Software/MyGit/Predicting-TWAS-features/FeaturePred.V2.0.R ./

cd /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS
tar -zcvf for_zac.tar.gz for_zac

```

***

#### FUSION YFS models

```{bash, eval=F, echo=T}
cd /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/for_zac

Rscript FeaturePred.V2.0.R \
	--PLINK_prefix_chr LDREF/1000G.EUR. \
	--weights FUSION.YFS/YFS.BLOOD.RNAARR.pos \
	--weights_dir FUSION.YFS \
	--ref_ld_chr LDREF/1000G.EUR. \
	--plink ./plink \
	--n_cores 1 \
	--memory 10000 \
	--all_mod T \
	--pigz ./pigz \
	--ref_maf LDREF/1000G.EUR. \
	--output test/FUSION.YFS

```

***

#### YFS eQTL-based models

```{bash, eval=F, echo=T}
cd /scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/for_zac

Rscript FeaturePred.V2.0.R \
	--PLINK_prefix_chr LDREF/1000G.EUR. \
	--weights eQTL.YFS/YFS.eQTL.pos \
	--weights_dir eQTL.YFS \
	--ref_ld_chr LDREF/1000G.EUR. \
	--plink ./plink \
	--n_cores 1 \
	--memory 10000 \
	--all_mod T \
	--pigz ./pigz \
	--ref_maf LDREF/1000G.EUR. \
	--output test/eQTL.YFS
	
```

***

### Compare predicted and observed expression in GTEx v8

Again, this will be run by Zac. To develop the code, I will use the predicted expression levels in the 1KG reference, and pretend that the FUSION enet predicted expression levels are the observed gene expression.

```{r, eval=F, echo=T}
library(data.table)

# Read in predicted expression
fusion<-fread('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/1kg_pred_exp/YFS.BLOOD.RNAARR/FeaturePredictions_YFS.BLOOD.RNAARR.txt.gz')
eqtl<-fread('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/1kg_pred_exp/YFS.eQTL/FeaturePredictions_YFS.eQTL.txt.gz')

# Make column names more concise (format = <source>.<genename>.<method>)
names(fusion)<-gsub('YFS.BLOOD.RNAARR.YFS','fusion', names(fusion))
names(eqtl)<-gsub('.RDat','',gsub('YFS.','', names(eqtl)))

# Identify genes available in fusion and eqtl based models
fusion_eqtl_intersect<-intersect(unique(gsub('\\..*','',sub('.*?\\.','', names(fusion)[-1:-2]))),unique(gsub('\\..*','',sub('.*?\\.','', names(eqtl)[-1:-2]))))

# Combine the predicted expression from FUSION and eQTL data
pred_exp<-merge(fusion, eqtl, by=c('FID','IID'))
pred_exp<-pred_exp[,grepl(paste0('FID$|IID$|',paste0('\\.',fusion_eqtl_intersect,'\\.',collapse='|')), names(pred_exp)), with=F]

# Create pretend observed expression
obs_exp<-pred_exp[, grepl('FID$|IID$|enet$', names(pred_exp)), with=F]
names(obs_exp)<-gsub('.enet','',gsub('fusion.','',names(obs_exp)))

#######
# Compare predicted and observed expression
#######
# Identify genes available in predicted and observed datasets
pred_exp_genes<-unique(gsub('\\..*','',sub('.*?\\.','', names(pred_exp)[-1:-2])))
obs_exp_genes<-names(obs_exp)[-1:-2]

genes<-intersect(pred_exp_genes,obs_exp_genes)

# Calculate correlation between observed expression and predicted expression from each method
cor_res<-list()
for(i in genes){
  # Rename columns to make output consistent across genes and label the observed expression
  pred_exp_tmp<-pred_exp[,grepl(paste0('FID$|IID$|\\.',i,'\\.'), names(pred_exp)), with=F]
  names(pred_exp_tmp)<-gsub(paste0('.',i), '', names(pred_exp_tmp))
  
  obs_exp_tmp<-obs_exp[,c('FID','IID',i), with=F]
  names(obs_exp_tmp)[names(obs_exp_tmp) == i]<-'obs'
  
  # merge predicted and observed expression
  both_exp_tmp<-merge(pred_exp_tmp, obs_exp_tmp, by=c('FID','IID'))

  # Calculate correlation
  cor_res[[i]]<-cor(both_exp_tmp[,-1:-2,with=F], use='p')
}

# melt and combine all the results
cor_res_melt<-melt(cor_res)
cor_res_melt_obs<-cor_res_melt[cor_res_melt$Var1 == 'obs',]
cor_res_melt_obs$test<-paste0(cor_res_melt_obs$Var1,'_',cor_res_melt_obs$Var2)

cor_res_average<-NULL
for(i in unique(cor_res_melt_obs$test)){
  cor_res_average<-rbind(cor_res_average, data.frame(test=i,
                                                     Mean=mean(cor_res_melt_obs$value[cor_res_melt_obs$test == i])))
}

library(ggplot2)
png('/scratch/groups/biomarkers-brc-mh/TWAS_resource/eQTL_to_TWAS/data/1kg_pred_exp/pred_obs_cor_plot_example.png', units='px', width=1500, height=1000, res=300)
ggplot(cor_res_melt_obs, aes(x=paste0(Var1,' vs. ',Var2), y=value)) +
  geom_boxplot() +
  labs(x='Test', y='Correlation') +
  coord_flip()
dev.off()


```




